---
title: "Trajectory inference with Slingshot"
subtitle: "Local version, reads from the workshop data repo, no network access"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: true
  error: false
---

Local rework of `1_a_slingshot.html`.

Upstream: <https://nbisweden.github.io/workshop-archive/workshop-scRNAseq/2020-01-27/labs/compiled/slingshot/slingshot.html>
(NBIS workshop-scRNAseq, 2020-01-27 archive)

Method: Street et al. 2018, "Slingshot: cell lineage and pseudotime inference
for single-cell transcriptomics", BMC Genomics,
<https://doi.org/10.1186/s12864-018-4772-9>

::: {.callout-note}
## Status: transcribed from upstream, renders cleanly end to end

The code below is transcribed from the upstream lab, with the GEO download
replaced by a read from the staged data repo. Three changes were needed to run
under current package versions: `buildSNNGraph()` no longer takes `assay.type`
when a reduction is supplied, and `getLineages()` / `getCurves()` now return a
`PseudotimeOrdering`, so they are wrapped in `SlingshotDataSet()` before the
base-graphics `lines()` calls.
:::

## Setup

```{r}
#| label: setup
source(here::here("slides", "find_data_dir.R"))
data_dir <- find_data_dir("1_a_slingshot")
report_data_dir("1_a_slingshot")
```

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(Matrix)
  library(scran)
  library(scater)
  library(igraph)
  library(slingshot)
  library(tradeSeq)
  library(dplyr)
  library(cowplot)
  library(RColorBrewer)
})
```

## Load data

Mouse bone marrow progenitors, GEO GSE72857 (Paul et al. 2015).

Upstream downloads `GSE72857_umitab.txt.gz` from the NCBI FTP server, decompresses
it, reads it with `read.delim()`, and converts it to a sparse matrix. The staging
script does all of that once and saves the sparse matrix, so here we just read it.

::: {.callout-note}
## Staged file

`GSE72857_umitab.rds`, a sparse `dgCMatrix` of raw UMI counts, genes by cells.
Produced by `staging/stage_1_a_slingshot.R`.
:::

```{r}
#| label: load
umi_counts <- readRDS(file.path(data_dir, "GSE72857_umitab.rds"))

# Select every fifth cell, to keep the exercise fast.
umi_counts <- umi_counts[, c(TRUE, FALSE, FALSE, FALSE, FALSE)]
dim(umi_counts)
```

```{r}
#| label: palette
# Colour palette used throughout.
pal <- c(RColorBrewer::brewer.pal(9, "Set1"), RColorBrewer::brewer.pal(8, "Set2"))
```

## Basic processing with the Scran / Scater pipeline

Upstream offers a Scran/Scater route and a Seurat route, and they are
alternatives, not steps. We use Scran/Scater because it keeps the object a
`SingleCellExperiment`, which is what Slingshot wants anyway. The Seurat route is
kept below but not run.

The goal of this section is only to produce two things, a reduction to run the
trajectory on, and a clustering.

```{r}
#| label: scran-pipeline
data <- SingleCellExperiment(assays = list(counts = umi_counts))
data <- computeSumFactors(data, sizes = c(20, 40, 60, 80))
data <- logNormCounts(data)
var.fit <- modelGeneVar(data)
top.hvgs <- getTopHVGs(var.fit, n = 2000)
data <- runPCA(data, subset_row = top.hvgs)
g <- buildSNNGraph(data, k = 5, use.dimred = "PCA")
data$louvain_SNNk5 <- factor(igraph::cluster_louvain(g)$membership)
data <- runUMAP(data, dimred = "PCA", n_dimred = 50, ncomponents = 2,
                spread = 2, min_dist = 0.1, n_neighbors = 10, metric = "cosine")
```

```{r}
#| label: plot-clusters
plotReducedDim(data, dimred = "UMAP", colour_by = "louvain_SNNk5")
```

```{r}
#| label: extract-for-slingshot
# Slingshot takes plain matrices, not the SingleCellExperiment.
dimred     <- reducedDim(data, type = "UMAP")
clustering <- data$louvain_SNNk5
counts     <- as.matrix(counts(data)[top.hvgs, ])
```

## Alternative: the Seurat route

Not run. Shown because students who did the earlier days in Seurat will want to
see how they get to the same two objects.

```{r}
#| label: seurat-pipeline
#| eval: false
library(Seurat)
data_seurat <- CreateSeuratObject(counts = umi_counts)
data_seurat <- NormalizeData(data_seurat)
data_seurat <- FindVariableFeatures(data_seurat, nfeatures = 2000)
data_seurat <- ScaleData(data_seurat)
data_seurat <- RunPCA(data_seurat)
data_seurat <- FindNeighbors(data_seurat)
data_seurat <- FindClusters(data_seurat, resolution = 1)
data_seurat <- RunUMAP(data_seurat, n.neighbors = 10, dims = 1:50,
                       spread = 2, min.dist = 0.3)
DimPlot(data_seurat, group.by = "RNA_snn_res.1")

dimred     <- data_seurat@reductions$umap@cell.embeddings
clustering <- data_seurat$RNA_snn_res.1
counts     <- as.matrix(data_seurat@assays$RNA@counts[
  data_seurat@assays$RNA@var.features, ])
```

## Defining cell lineages with Slingshot

`slingshot()` is a wrapper around two steps. First define the lineages, which
are paths through the clusters. Then fit a curve through the data for each
lineage. They are shown separately here for clarity.

```{r}
#| label: lineages
set.seed(1)
lineages <- getLineages(data = dimred, clusterLabels = clustering)
lineages
```

```{r}
#| label: plot-lineages
plot(dimred, col = pal[clustering], asp = 1, pch = 16)
# getLineages() now returns a PseudotimeOrdering; the base-graphics lines()
# method is defined for SlingshotDataSet, so convert before plotting.
lines(SlingshotDataSet(lineages), lwd = 3, col = "black")
```

## Defining principal curves

Principal curves iteratively deform an initial curve to fit the data. Slingshot
runs one per lineage, and bundles curves that share clusters.

`approx_points = 300` is a speed compromise. With all cells this is slow.

```{r}
#| label: curves
curves <- getCurves(lineages, approx_points = 300, thresh = 0.01,
                    stretch = 0.8, allow.breaks = FALSE, shrink = 0.99)
curves
```

```{r}
#| label: plot-curves
plot(dimred, col = pal[clustering], asp = 1, pch = 16)
lines(SlingshotDataSet(curves), lwd = 3, col = "black")
```

## Finding differentially expressed genes

tradeSeq smooths expression along the trajectory with generalized additive
models, then tests whether coefficients differ between points on the trajectory.

```{r}
#| label: tradeseq-setup
BiocParallel::register(BiocParallel::SerialParam())
```

::: {.callout-warning}
## `fitGAM` is the slow step

Upstream filters hard before fitting, down to roughly 215 genes, purely to make
the tutorial finish. That filter is not a real analysis choice. In your own work
you would fit at least the variable genes.

If this is still too slow on the workshop machine, precompute the fitted object
in the staging script and read it here instead.
:::

```{r}
#| label: fitgam
# Stringent filter, for speed only.
filt_counts <- counts[rowSums(counts > 5) > ncol(counts) / 100, ]
dim(filt_counts)

sce <- fitGAM(counts = as.matrix(filt_counts), sds = curves)
```

```{r}
#| label: plot-gene-count
plotGeneCount(curves, filt_counts, clusters = clustering, models = sce)
```

### Genes that change with pseudotime

```{r}
#| label: association-test
pseudotime_association <- associationTest(sce)
pseudotime_association$fdr <- p.adjust(pseudotime_association$pvalue,
                                       method = "fdr")
pseudotime_association <- pseudotime_association[
  order(pseudotime_association$pvalue), ]
pseudotime_association$feature_id <- rownames(pseudotime_association)
head(pseudotime_association, 20)
```

```{r}
#| label: plot-helper
# Upstream defines a helper that immediately overwrites its own argument, so it
# always plots the same gene no matter what you pass it. Fixed here to actually
# use the argument.
plot_differential_expression <- function(feature_id) {
  cowplot::plot_grid(
    plotGeneCount(curves, filt_counts, gene = feature_id[1],
                  clusters = clustering, models = sce) +
      ggplot2::theme(legend.position = "none"),
    plotSmoothers(sce, as.matrix(counts), gene = feature_id[1]))
}
```

```{r}
#| label: plot-association
feature_id <- pseudotime_association |>
  dplyr::filter(pvalue < 0.05) |>
  dplyr::top_n(1, -waldStat) |>
  dplyr::pull(feature_id)

plot_differential_expression(feature_id)
```

### Genes that change between two pseudotime points

```{r}
#| label: start-end-test
pseudotime_start_end_association <- startVsEndTest(sce, pseudotimeValues = c(0, 1))
pseudotime_start_end_association$feature_id <-
  rownames(pseudotime_start_end_association)

feature_id <- pseudotime_start_end_association |>
  dplyr::filter(pvalue < 0.05) |>
  dplyr::top_n(1, waldStat) |>
  dplyr::pull(feature_id)

plot_differential_expression(feature_id)
```

### Genes that are different between lineages

```{r}
#| label: lineage-test
different_end_association <- diffEndTest(sce)
different_end_association$feature_id <- rownames(different_end_association)

feature_id <- different_end_association |>
  dplyr::filter(pvalue < 0.05) |>
  dplyr::arrange(dplyr::desc(waldStat)) |>
  dplyr::slice(1) |>
  dplyr::pull(feature_id)

plot_differential_expression(feature_id)
```

## Session info

```{r}
#| label: sessioninfo
sessionInfo()
```
