---
title: "Trajectory inference with Monocle 3"
subtitle: "Local version, reads from the workshop data repo, no network access"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: true
  error: false
---

Transcribed from `1_b_monocle3.html`.

Upstream: Julia Jakiela, "Inferring single cell trajectories with Monocle3 (R)",
Galaxy Training Network,
<https://training.galaxyproject.org/training-material/topics/single-cell/tutorials/scrna-case_monocle3-rstudio/tutorial.html>

Data: a mouse model of fetal growth restriction, Bacon et al. 2018.
Method: Cao et al. 2019, "The single-cell transcriptional landscape of
mammalian organogenesis", Nature, <https://doi.org/10.1038/s41586-019-0969-x>

::: {.callout-note}
## Status: transcribed, renders cleanly end to end

Every chunk is transcribed from the upstream tutorial. The whole document renders
without error against the staged `cell_data_set` and the current package
versions, with no code changes needed.
:::

::: {.callout-note}
## Local data file

Upstream downloads `AnnData_filtered.h5ad` from Zenodo and converts it to a
Monocle `cell_data_set`. Here we read one staged file,
`AnnData_filtered_cds.rds`, which is that conversion already done, an object of
class `cell_data_set`. The staging script builds it once, so the workshop does
not need network access, a Python/reticulate `anndata` install, or the
matrix-transpose and column-renaming steps.
:::

::: {.callout-warning}
## The usual student stumbling block

Monocle 3 infers a trajectory **within a single partition**. If `learn_graph()`
produces a disconnected mess, the answer is almost always that the cells span
several partitions, not that the parameters are wrong. That is what the
`partition_qval` tuning below is for.
:::

## Setup

```{r}
#| label: setup
source(here::here("slides", "find_data_dir.R"))
data_dir <- find_data_dir("1_b_monocle3")
report_data_dir("1_b_monocle3")
```

```{r}
#| label: libraries
library(monocle3)
library(magrittr)     # needed for %>%
library(viridisLite)
library(ggplot2)
```

## Upload, view and modify the files

Upstream at this point runs

```r
download.file('https://zenodo.org/record/7455590/files/AnnData_filtered.h5ad',
              'AnnData.h5ad')
ann <- anndata::read_h5ad('AnnData.h5ad')
expression_matrix <- ann$X
cell_metadata     <- ann$obs
gene_metadata     <- ann$var
expression_matrix <- as.matrix(expression_matrix)
expression_matrix <- t(expression_matrix)   # Monocle wants genes as rows
colnames(gene_metadata)[2] <- 'gene_short_name'
cds <- new_cell_data_set(expression_matrix, cell_metadata, gene_metadata)
```

All of that has been done once at staging time. We read the result.

```{r}
#| label: load-cds
cds <- readRDS(file.path(data_dir, "AnnData_filtered_cds.rds"))
cds
```

```{r}
#| label: inspect-cds
# Monocle needs genes as rows and cells as columns, and needs a column of
# gene_metadata named gene_short_name. Confirm both on the staged object.
head(exprs(cds), c(5, 5))
head(fData(cds))
colnames(fData(cds))
```

## Monocle workflow

## Pre-processing

Normalisation and PCA. `num_dim` is the number of principal components.
This step is slow, roughly 5 minutes at `num_dim = 210`.

```{r}
#| label: preprocess
# PCA pre-processing with 210 principal components
cds_preprocessing <- preprocess_cds(cds, method = "PCA", num_dim = 210)

# plot the variation in gene expression vs PCA components
plot_pc_variance_explained(cds_preprocessing)
```

## Batch correction and Dimensionality reduction

The dataset comprises 7 samples, so batch effects are a real risk.

```{r}
#| label: check-batch-column
colnames(colData(cds_preprocessing))    # confirm the batch column is called 'batch'
```

```{r}
#| label: umap-before-alignment
# reduce dimension first
cds_preprocessing_UMAP <- reduce_dimension(cds_preprocessing,
                                           reduction_method = "UMAP",
                                           preprocess_method = "PCA")

# then produce a plot to check for batch effects
plot_cells(cds_preprocessing_UMAP, color_cells_by = "batch",
           label_cell_groups = FALSE)
```

```{r}
#| label: align
# perform batch correction
cds_batch <- align_cds(cds_preprocessing_UMAP, preprocess_method = "PCA",
                       alignment_group = "batch")
```

```{r}
#| label: umap-after-alignment
# dimensionality reduction after alignment, now on the aligned space
cds_red_dim <- reduce_dimension(cds_batch, preprocess_method = "Aligned",
                                reduction_method = "UMAP")

# see the batch correction effect on a plot
plot_cells(cds_red_dim, color_cells_by = "batch", label_cell_groups = FALSE)
```

## Clustering: clusters

```{r}
#| label: cluster-default
# clustering with the default (automatic) resolution
cds_auto_cluster <- cluster_cells(cds_red_dim, reduction_method = "UMAP")

plot_cells(cds_auto_cluster, reduction_method = "UMAP",
           color_cells_by = "cluster", group_label_size = 5)
```

```{r}
#| label: cluster-resolution
# higher resolution gives more clusters, which helps assign cell types
cds_clustered <- cluster_cells(cds_red_dim, reduction_method = "UMAP",
                               resolution = 0.0002)

plot_cells(cds_clustered, reduction_method = "UMAP",
           color_cells_by = "cluster", group_label_size = 5)
```

## Clustering: partitions

Trajectory inference happens within one partition, so check the partitions
before learning the graph.

```{r}
#| label: partitions-default
plot_cells(cds_clustered, reduction_method = "UMAP",
           color_cells_by = "partition", label_cell_groups = FALSE)
```

Raising `partition_qval` above its default of 0.05 widens the partitions, that
is, it yields fewer of them. Check afterwards that the clusters did not change.

```{r}
#| label: partitions-qval
# changing the partition q-value
cds_clustered <- cluster_cells(cds_red_dim, reduction_method = "UMAP",
                               resolution = 0.0002, partition_qval = 1)

# see the partitions with the changed q-value
plot_cells(cds_clustered, reduction_method = "UMAP",
           color_cells_by = "partition", label_cell_groups = FALSE)

# check if clusters didn't change
plot_cells(cds_clustered, reduction_method = "UMAP",
           color_cells_by = "cluster", label_cell_groups = FALSE)
```

## Additional step: assigning cells to one partition

Not needed for this dataset. It is the last resort when `partition_qval` cannot
merge the partitions you need merged.

```{r}
#| label: manual-partition
# simulate a badly partitioned dataset
cds_partitions_extra <- cluster_cells(cds_red_dim, reduction_method = "UMAP",
                                      partition_qval = 0.0001)

plot_cells(cds_partitions_extra, reduction_method = "UMAP",
           color_cells_by = "partition", label_cell_groups = FALSE)

# store '1' the number of times equal to the number of cells
big_partition <- c(rep(1, length(cds_partitions_extra@colData@rownames)))

# take the barcode of each cell and assign '1' to each of them
names(big_partition) <- cds_partitions_extra@colData@rownames

# convert from numeric to factor
big_partition <- as.factor(big_partition)

# overwrite the partition slot of the CDS object
cds_partitions_extra@clusters$UMAP$partitions <- big_partition

plot_cells(cds_partitions_extra, reduction_method = "UMAP",
           color_cells_by = "partition", label_cell_groups = FALSE)
```

## Assigning cell types

### Supervised approach

Marker genes from Bacon et al. 2018:

| Marker | Cell type |
|---|---|
| Il2ra | Double negative (early T-cell) |
| Cd8b1, Cd8a, Cd4 | Double positive (middle T-cell) |
| Cd8b1, Cd8a, Cd4 high | Double positive (late middle T-cell) |
| Itm2a | Mature T-cell |
| Aif1 | Macrophages |
| Hba-a1 | RBC |

```{r}
#| label: marker-expression
plot_cells(cds_clustered,
           genes = c('Il2ra', 'Cd8b1', 'Cd8a', 'Cd4', 'Itm2a', 'Aif1', 'Hba-a1'),
           reduction_method = "UMAP")
```

```{r}
#| label: annotate
# keep the objects tidy so you can come back to any point of the analysis
cds_annotated <- cds_clustered

# create a new column 'cell_type' and initialise it with cluster values
colData(cds_annotated)$cell_type <- as.character(clusters(cds_annotated))

# Annotate clusters. Cluster numbering is not reproducible across package
# versions, so check the marker plot above before trusting this mapping.
# dplyr::recode leaves any unmatched level unchanged.
colData(cds_annotated)$cell_type <- dplyr::recode(
  colData(cds_annotated)$cell_type,
  '1' = 'DP-M1',    # double positive, middle T-cell (1st cluster)
  '2' = 'DP-M2',    # double positive, middle T-cell (2nd cluster)
  '3' = 'T-mat',    # mature T-cell
  '4' = 'DN',       # double negative, early T-cell
  '5' = 'DP-L',     # double positive, late middle T-cell
  '6' = 'DP-M3',    # double positive, middle T-cell (3rd cluster)
  '7' = 'Unknown',
  '8' = 'Unknown')

plot_cells(cds_annotated, color_cells_by = "cell_type",
           label_cell_groups = FALSE)
```

### Unsupervised approach

`top_markers()` is one of the slow steps. `reference_cells` speeds up the
significance test at some cost in sensitivity.

```{r}
#| label: top-markers
marker_test <- top_markers(cds_clustered, group_cells_by = "cluster",
                           reduction_method = "UMAP",
                           reference_cells = 1000, cores = 8)
```

```{r}
#| label: top-markers-filter
top_specific_markers <- marker_test %>%
  dplyr::filter(fraction_expressing >= 0.10) %>%   # fraction of expressing cells
  dplyr::group_by(cell_group) %>%
  dplyr::top_n(1, pseudo_R2)                       # one top gene per cluster

# Ties in pseudo_R2 give more genes per cluster than asked for. Break them.
top_specific_markers <- top_specific_markers %>%
  dplyr::distinct(pseudo_R2, .keep_all = TRUE)

# store the names of the marker genes
top_marker_names <- unique(top_specific_markers %>%
                             dplyr::pull(gene_short_name))
top_marker_names
```

```{r}
#| label: genes-by-group
plot_genes_by_group(cds_clustered,                    # our CDS object
                    top_marker_names,                 # genes to show
                    group_cells_by = "cluster",       # how to group cells
                    ordering_type = "maximal_on_diag")
```

```{r}
#| label: plot-found-markers
# plot the identified genes to check their expression in the bigger picture
plot_cells(cds_clustered,
           genes = c('Pcdhgc4', 'Pcdhga5', 'Gm5559', 'Gm10359', 'Ccr9',
                     'Cd8b1', 'Plac8', 'Il2ra', 'Cd52', 'Tmsb10', 'Mki67',
                     'Hmgb2', 'Pclaf', 'Npm1'),
           reduction_method = "UMAP")
```

### A marker file for Garnett

```{r}
#| label: garnett
# top_markers() again, now grouping cells by the assigned cell type
assigned_marker_test <- top_markers(cds_annotated,
                                    group_cells_by = "cell_type",
                                    reference_cells = 1000,
                                    cores = 8)

# filter these markers according to how stringent you want to be
garnett_markers <- assigned_marker_test %>%
  dplyr::filter(marker_test_q_value < 0.05 & specificity >= 0.25) %>%
  dplyr::group_by(cell_group) %>%
  dplyr::top_n(5, marker_score)

# exclude genes that are good markers for more than one cell type
garnett_markers_filtered <- garnett_markers %>%
  dplyr::group_by(gene_short_name) %>%
  dplyr::filter(dplyr::n() == 1)

# Write to a temp file rather than the repo, so rendering leaves no artefacts.
marker_file <- file.path(tempdir(), "marker_file.txt")
generate_garnett_marker_file(garnett_markers_filtered, file = marker_file)
cat(readLines(marker_file), sep = "\n")
```

## Trajectory inference

`use_partition = TRUE` learns a disjoint graph per partition, which is what we
want. The 'Unknown' cells sit in their own partition and should not be joined
to the T-cell trajectory.

```{r}
#| label: learn-graph
cds_trajectory <- learn_graph(cds_annotated, use_partition = TRUE)

plot_cells(cds_trajectory,
           color_cells_by = "cell_type",
           label_cell_groups = FALSE,
           label_groups_by_cluster = FALSE,
           label_leaves = FALSE,
           label_branch_points = FALSE)
```

### Option 1: Root nodes

```{r}
#| label: principal-points
# label_principal_points = TRUE shows the Y_<number> node names
plot_cells(cds_trajectory,
           color_cells_by = "cell_type",
           label_cell_groups = FALSE,
           label_groups_by_cluster = FALSE,
           label_leaves = FALSE,
           label_branch_points = FALSE,
           label_principal_points = TRUE,
           graph_label_size = 3)
```

Upstream reads a node name off that plot and hardcodes it, for example
`order_cells(cds_trajectory, root_pr_nodes = 'Y_14')`. Node numbering is not
stable across runs, so instead we use the tutorial's helper function, which
picks the root node from the annotated cell types. That keeps the document
non-interactive and independent of the node numbering.

```{r}
#| label: root-helper
# a helper function to identify the root principal point
get_correct_root_state <- function(cds, cell_phenotype, root_type) {
  cell_ids <- which(pData(cds)[, cell_phenotype] == root_type)

  closest_vertex <-
    cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  closest_vertex <- as.matrix(closest_vertex[colnames(cds), ])
  root_pr_nodes <-
    igraph::V(principal_graph(cds)[["UMAP"]])$name[
      as.numeric(names(which.max(table(closest_vertex[cell_ids, ]))))]
  root_pr_nodes
}

# find the node where the DN (double negative, earliest) cells reside
DN_node_id <- get_correct_root_state(cds_trajectory,
                                     cell_phenotype = 'cell_type', "DN")
DN_node_id

cds_order_1_helper <- order_cells(cds_trajectory, root_pr_nodes = DN_node_id)
```

### Option 2: Root cells

```{r}
#| label: root-cells
# all cells of the type we believe starts the trajectory
starting_cell_type <- 'DN'
index <- which(cds_trajectory@colData$cell_type == starting_cell_type)
DN_cells <- colnames(cds_trajectory)[index]

# on unannotated data you would instead name a cluster
starting_cluster <- colnames(cds_trajectory[, clusters(cds_trajectory) == 4])

cds_order_2 <- order_cells(cds_trajectory, root_cells = DN_cells)
```

Upstream also mentions calling `order_cells(cds_trajectory)` with no root
argument, which opens a cell-picking window. That cannot run in a rendered
document, so it is not included here.

## Plotting in pseudotime

```{r}
#| label: pseudotime
cds_order <- cds_order_1_helper

plot_cells(cds_order,
           color_cells_by = "pseudotime",
           label_cell_groups = FALSE,
           label_leaves = FALSE,
           label_branch_points = FALSE)

# access pseudotime for each cell and store it alongside the cell metadata
pseudotime <- pseudotime(cds_order)
cds_order@colData$pseudotime <- pseudotime
summary(pseudotime)
```

## Differential expression analysis

A small gene subset keeps the regression examples fast.

```{r}
#| label: de-subset
test_genes <- c('Il2ra', 'Cd8b1', 'Cd8a', 'Cd4', 'Itm2a', 'Aif1', 'Hba-a1')
cds_subset <- cds_order[rowData(cds_order)$gene_short_name %in% test_genes, ]
cds_subset
```

### Some more plotting

```{r}
#| label: violin
plot_genes_violin(cds_subset, group_cells_by = "cell_type", ncol = 2)
```

```{r}
#| label: genes-in-pseudotime
plot_genes_in_pseudotime(cds_subset, color_cells_by = "cell_type",
                         min_expr = 0.5)
```

### Regression analysis, advanced

```{r}
#| label: fit-models
# fit a generalized linear model for each gene
gene_fits <- fit_models(cds_subset, model_formula_str = "~cell_type + batch")

# extract a table of coefficients, Wald-tested against zero
fit_coefs <- coefficient_table(gene_fits)
head(fit_coefs)
```

```{r}
#| label: coef-filter
# the intercept term is not of interest
no_intercept_coefs <- fit_coefs %>% dplyr::filter(term != "(Intercept)")

# extract results for DP-M1 cells only
DP_M1_coefs <- fit_coefs %>% dplyr::filter(term == "cell_typeDP-M1")

# control the false discovery rate and keep a few columns
DP_M1_coefs_filtered <- DP_M1_coefs %>%
  dplyr::filter(q_value < 0.05) %>%
  dplyr::select(gene_short_name, term, q_value, estimate)

DP_M1_coefs_filtered
```

### Graph-autocorrelation analysis, advanced

`graph_test()` uses Moran's I. This runs on the full dataset and is the slowest
step in the tutorial.

```{r}
#| label: graph-test
graph_test_res <- graph_test(cds_order, neighbor_graph = "knn", cores = 8)
head(graph_test_res)
```

```{r}
#| label: gene-modules
# gene IDs for which q-value < 0.05
pr_deg_ids <- row.names(subset(graph_test_res, q_value < 0.05))

# group genes into modules, this runs UMAP on the genes rather than the cells
gene_module_df <- find_gene_modules(cds_order[pr_deg_ids, ], resolution = 1e-2)
```

```{r}
#| label: module-heatmap
cell_group_df <- tibble::tibble(
  cell = row.names(colData(cds_order)),
  cell_group = clusters(cds_order)[colnames(cds_order)])

agg_mat <- aggregate_gene_expression(cds_order, gene_module_df, cell_group_df)
row.names(agg_mat) <- stringr::str_c("Module ", row.names(agg_mat))
colnames(agg_mat)  <- stringr::str_c("Cluster ", colnames(agg_mat))

pheatmap::pheatmap(agg_mat, cluster_rows = TRUE, cluster_cols = TRUE,
                   scale = "column", clustering_method = "ward.D2",
                   fontsize = 6)
```

```{r}
#| label: plot-modules
# Module numbering varies between runs, so take whichever modules exist rather
# than the upstream hardcoded c(40, 39, 36, 17).
modules_to_plot <- head(sort(unique(gene_module_df$module)), 4)

plot_cells(cds_order,
           genes = gene_module_df %>%
             dplyr::filter(module %in% modules_to_plot),
           group_cells_by = "cluster",
           color_cells_by = "cluster",
           show_trajectory_graph = FALSE)
```

## 3D plotting

```{r}
#| label: plot-3d
cds_3d <- reduce_dimension(cds_order, preprocess_method = 'Aligned',
                           max_components = 3)

plot_cells_3d(cds_3d, color_cells_by = "cell_type")
```

## Export plots

```{r}
#| label: export
plot_pseudotime <- plot_cells(cds_order,
                              color_cells_by = "pseudotime",
                              label_cell_groups = FALSE,
                              label_leaves = FALSE,
                              label_branch_points = FALSE)

# Written to a temp directory so rendering leaves nothing behind in the repo.
svg(file.path(tempdir(), "plot_pseudotime.svg"))
plot_pseudotime
dev.off()
```

## Session info

```{r}
#| label: sessioninfo
sessionInfo()
```
