---
title: "Pseudobulk differential expression with DESeq2"
subtitle: "Local version, reads from the workshop data repo, no network access"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: true
  error: false
---

::: {.callout-important}
## Read this before you trust anything below

This document is a transcription of the upstream HBC lesson
*Single-cell RNA-seq: Pseudobulk differential expression analysis*
(<https://hbctraining.github.io/scRNA-seq/lessons/pseudobulk_DESeq2_scrnaseq.html>),
adapted only so that it reads its input from the workshop data repo instead of
a downloaded `data/` folder.

The document renders end to end. Two chunks fail on purpose and are recovered by
a fallback in the very next chunk, so the rendered page shows exactly two error
blocks and then keeps going.

The upstream page carries an **ARCHIVED** banner and the lesson script is dated
**May 2020**. Expect API drift against current DESeq2, Seurat, and scater. Two
calls are known to be affected. `scater::calculateQCMetrics()` is now defunct,
and `Matrix.utils` was archived on CRAN so `aggregate.Matrix()` is unavailable.
Both are transcribed as the tutorial has them, each followed by a modern
equivalent. The document sets `error: false`, so any unexpected failure stops
the render and is reported. These two chunks, `qc-tutorial` and
`aggregate-tutorial`, carry a chunk-level `error: true` because their failure
is the point being taught.
:::

## Learning objectives

- Understand how to prepare single-cell RNA-seq raw count data for pseudobulk
  differential expression analysis.
- Use DESeq2 to perform pseudobulk differential expression analysis on a
  specific cell type cluster.
- Create functions to iterate the pseudobulk differential expression analysis
  across different cell types.

## Differential expression analysis with DESeq2

After identifying the cell type identities of the scRNA-seq clusters, we often
want to perform differential expression analysis between conditions within
particular cell types. Functions exist within Seurat to do this, but the
p-values are often inflated because each cell is treated as a sample. Single
cells within a sample are not independent of each other. If we treat cells as
samples we are investigating variation within an individual, not variation
across a population.

The current best practice is a pseudobulk approach:

1. Subset to the cells for the cell type(s) of interest.
2. Extract the raw counts after QC filtering.
3. Aggregate the counts and metadata to the sample level.
4. Perform the DE analysis. At least two biological replicates per condition
   are required, and more are recommended.

## Setting up the R environment

```{r}
#| label: setup
data_dir <- "~/MEGA/workshop_data/2_a_pseudobulk_de"
in_data_dir <- function(basename) {
  return(file.path(data_dir, basename))
}
```

### Where the data came from

Upstream tells students to right-click and download the RDS object into a
`data/` folder. The verbatim download link on the lesson page is:

```
https://www.dropbox.com/s/dgdooocb1a03dvk/scRNA-seq_input_data_for_DE.rds?dl=1
```

For the workshop that file is already staged, so the chunks below read it from
`data_dir` and never touch the network.

### Load libraries

```{r}
#| label: libraries
# May 2020
# Single-cell RNA-seq analysis - Pseudobulk DE analysis with DESeq2

# Load libraries
library(scater)
library(Seurat)
library(tidyverse)
library(cowplot)
# library(Matrix.utils)  # archived on CRAN; not loaded. See the aggregation
#                        # section, which shows aggregate.Matrix() failing and
#                        # uses a Matrix-only replacement instead.
library(edgeR)
library(dplyr)
library(magrittr)
library(Matrix)
library(purrr)
library(reshape2)
library(S4Vectors)
library(tibble)
library(SingleCellExperiment)
library(pheatmap)
library(apeglm)
library(png)
library(DESeq2)
library(RColorBrewer)
```

The table below records which packages the tutorial uses and their versions,
including the archived `Matrix.utils`, which is intentionally not loaded above.
It is a quick way to confirm the environment before running the analysis.

```{r}
#| label: package-availability
needed <- c("scater", "Seurat", "tidyverse", "cowplot", "Matrix.utils",
            "edgeR", "dplyr", "magrittr", "Matrix", "purrr", "reshape2",
            "S4Vectors", "tibble", "SingleCellExperiment", "pheatmap",
            "apeglm", "png", "DESeq2", "RColorBrewer", "DEGreport", "scuttle")
data.frame(
  package = needed,
  installed = vapply(needed, requireNamespace, logical(1), quietly = TRUE),
  version = vapply(needed,
                   function(p) {
                     if (requireNamespace(p, quietly = TRUE)) {
                       as.character(utils::packageVersion(p))
                     } else {
                       NA_character_
                     }
                   },
                   character(1)),
  row.names = NULL)
```

### Load the RDS object

```{r}
#| label: load-data
# Read in the dataset
sce <- readRDS(file.path(data_dir, "scRNA-seq_input_data_for_DE.rds"))
sce
```

### Deriving the aggregation metadata

::: {.callout-important}
## The staged object is the raw Kang object, so we build the three key columns here

The upstream lesson assumes `sce` already carries `cluster_id`, `sample_id`, and
`group_id`. The object distributed for this workshop is the raw Kang et al. 2017
`SingleCellExperiment`, whose `colData` instead has `cell` (cell type), `ind`
(individual), and `stim` (condition). We derive the three columns the tutorial
needs, exactly as `muscat::prepSCE()` would: `cluster_id` from the cell type,
`group_id` from the condition, and `sample_id` from condition plus individual
(so `ctrl` and individual `101` give sample `ctrl101`). Without this step every
chunk below fails, because `sce$cluster_id` and friends are `NULL`.
:::

```{r}
#| label: derive-metadata
# Drop the handful of cells with no cell-type assignment.
sce <- sce[, !is.na(sce$cell)]

sce$cluster_id <- factor(sce$cell)
sce$group_id   <- factor(sce$stim)
sce$sample_id  <- factor(paste0(sce$stim, sce$ind))

head(colData(sce)[, c("cluster_id", "sample_id", "group_id")])
```

## Exploring the dataset

The data is the Kang et al. 2017 lupus PBMC dataset, the same one used for the
rest of the single-cell workflow. Here it is the non-pooled count data with
eight control samples and eight interferon stimulated samples, so there are real
biological replicates to work with.

```{r}
#| label: explore-counts
# Explore the raw counts for the dataset

## Check the assays present
assays(sce)

## Explore the raw counts for the dataset
dim(counts(sce))

counts(sce)[1:6, 1:6]
```

We do not run `head()` on this object, since it would print thousands of
columns. We look at the first six rows and columns instead.

```{r}
#| label: explore-metadata
## Explore the cellular metadata for the dataset
dim(colData(sce))

head(colData(sce))
```

For every cell we have the condition (`ctrl` or `stim`), the sample ID, and the
cell type.

## Acquiring necessary metrics for aggregation across cells in a sample

```{r}
#| label: clusters-and-samples
# Named vector of cluster names
kids <- purrr::set_names(levels(sce$cluster_id))
kids

# Total number of clusters
nk <- length(kids)
nk

# Named vector of sample names
sids <- purrr::set_names(levels(sce$sample_id))

# Total number of samples
ns <- length(sids)
ns
```

To perform sample-level DE we need sample-level metadata. We reorder the
single-cell metadata to match the order of the sample ID factor levels, then
take the sample-level information from the first cell of each sample.

```{r}
#| label: sample-metadata
# Generate sample level metadata

## Determine the number of cells per sample
table(sce$sample_id)

## Turn named vector into a numeric vector of number of cells per sample
n_cells <- as.numeric(table(sce$sample_id))

## Determine how to reorder the samples (rows) of the metadata to match the
## order of sample names in the sids vector
m <- match(sids, sce$sample_id)

## Create the sample level metadata by combining the reordered metadata with
## the number of cells corresponding to each sample.
ei <- data.frame(colData(sce)[m, ],
                 n_cells, row.names = NULL) %>%
  dplyr::select(-"cluster_id")
ei
```

## Quality control

Upstream removes count outliers and low count genes with `scater`, following the
2019 Bioconductor tutorial. A real analysis would do a more careful and more
hands-on job of QC.

```{r}
#| label: qc-tutorial
#| error: true
# Perform QC if not already performed
dim(sce)

# Calculate quality control (QC) metrics. This is the upstream call, and it fails
# on purpose: calculateQCMetrics() is defunct in current scater. The next chunk
# computes the same is_outlier flag with scuttle, the modern equivalent.
sce <- calculateQCMetrics(sce)
```

::: {.callout-warning}
## `calculateQCMetrics()` was removed from scater

`scater::calculateQCMetrics()` and the `total_features_by_counts` column it
created were deprecated in Bioconductor 3.10 and removed after that. The modern
equivalent lives in `scuttle`, where the per-cell count of detected genes is
called `detected`. The chunk below computes `sce$is_outlier` the modern way, but
only if the tutorial chunk above failed to create it, so the tutorial code stays
authoritative when it happens to work.
:::

```{r}
#| label: qc-modern-fallback
if (is.null(sce$is_outlier)) {
  message("calculateQCMetrics() did not run. Using scuttle::perCellQCMetrics().")
  qc <- scuttle::perCellQCMetrics(sce)
  sce$total_features_by_counts <- qc$detected
  sce$is_outlier <- scuttle::isOutlier(
    metric = qc$detected,
    nmads = 2, type = "both", log = TRUE)
}
table(sce$is_outlier)
```

```{r}
#| label: qc-filter
# Remove outlier cells
sce <- sce[, !sce$is_outlier]
dim(sce)

## Remove lowly expressed genes which have less than 10 cells with any counts
sce <- sce[rowSums(counts(sce) > 1) >= 10, ]

dim(sce)
```

## Count aggregation to sample level

We take the sum of counts for each sample within each cell type.

```{r}
#| label: aggregate-tutorial
#| error: true
# Aggregate the counts per sample_id and cluster_id

# Subset metadata to only include the cluster and sample IDs to aggregate across
groups <- colData(sce)[, c("cluster_id", "sample_id")]

# Aggregate across cluster-sample groups. This is the upstream call, and it fails
# on purpose: aggregate.Matrix() lived in the archived Matrix.utils. The
# Matrix-only replacement two chunks down produces the same sums, and the chunk
# after it inspects the result.
pb <- aggregate.Matrix(t(counts(sce)),
                       groupings = groups, fun = "sum")
```

::: {.callout-important}
## `Matrix.utils` is archived on CRAN

`Matrix.utils` was archived on CRAN in 2020, so `aggregate.Matrix()` may simply
not be installable on a current R. The commented-out chunk below does the same
sum-by-group using only `Matrix`, which is a base recommended package. A sparse
model matrix of group indicators, crossproduct with the cell-by-gene matrix,
gives exactly the group-by-gene sums that `aggregate.Matrix(fun = "sum")`
returns, with the same `cluster_sample` row names.

The chunk after it runs that fallback for real, but only when `pb` does not
already exist, so the document can carry on to DESeq2 either way.
:::

```{r}
#| label: aggregate-fallback-shown
# # Matrix-only replacement for Matrix.utils::aggregate.Matrix(fun = "sum")
#
# group_key <- factor(paste(colData(sce)$cluster_id,
#                           colData(sce)$sample_id,
#                           sep = "_"))
#
# # One indicator column per cluster-sample group, cells in rows.
# mm <- Matrix::sparse.model.matrix(~ 0 + group_key)
# colnames(mm) <- levels(group_key)
#
# # crossprod(mm, cells_by_genes) sums each gene over the cells of each group.
# pb <- Matrix::crossprod(mm, Matrix::t(counts(sce)))
```

```{r}
#| label: aggregate-fallback-run
if (!exists("pb")) {
  message("aggregate.Matrix() unavailable. Using the Matrix-only fallback.")
  group_key <- factor(paste(colData(sce)$cluster_id,
                            colData(sce)$sample_id,
                            sep = "_"))
  mm <- Matrix::sparse.model.matrix(~ 0 + group_key)
  colnames(mm) <- levels(group_key)
  pb <- Matrix::crossprod(mm, Matrix::t(counts(sce)))
}
class(pb)
dim(pb)
pb[1:6, 1:6]
```

The output is a sparse matrix with cell type-sample rows and gene columns. For
example, within B cells, sample `ctrl101` has a count for gene `NOC2L`.

To run DE per cell type we need to split the data by cell type, and transpose
each piece so that genes are rows and samples are columns. Not every sample
contains cells of every cell type, so we first work out which rows belong to
which cluster.

```{r}
#| label: split-by-cluster
# Not every cluster is present in all samples; create a vector that represents
# how to split samples
splitf <- sapply(stringr::str_split(rownames(pb),
                                    pattern = "_",
                                    n = 2),
                 `[`, 1)

# Turn into a list and split the list into components for each cluster and
# transform, so rows are genes and columns are samples and make rownames as the
# sample IDs
pb <- split.data.frame(pb,
                       factor(splitf)) %>%
  lapply(function(u)
    set_colnames(t(u),
                 stringr::str_extract(rownames(u), "(?<=_)[:alnum:]+")))

class(pb)

# Explore the different components of list
str(pb)
```

```{r}
#| label: cells-per-cluster-sample
# Print out the table of cells in each cluster-sample group
options(width = 100)
table(sce$cluster_id, sce$sample_id)
```

## Differential gene expression with DESeq2

DESeq2 normalizes the count data to account for differences in library size and
RNA composition between samples. We then use the normalized counts for QC plots
at the gene and sample level, and finally run the differential expression test.

### Sample-level metadata

We need metadata for all samples, including cluster ID, sample ID, and the
condition of interest (`group_id`), plus any other sample-level metadata such as
batch, sex, or age. `ei` holds the sample ID and condition, so we combine it
with the cluster IDs.

```{r}
#| label: de-samples
# Get sample names for each of the cell type clusters

# prep. data.frame for plotting
get_sample_ids <- function(x) {
  pb[[x]] %>%
    colnames()
}

de_samples <- map(1:length(kids), get_sample_ids) %>%
  unlist()
```

```{r}
#| label: de-cluster-ids
# Get cluster IDs for each of the samples

samples_list <- map(1:length(kids), get_sample_ids)

get_cluster_ids <- function(x) {
  rep(names(pb)[x],
      each = length(samples_list[[x]]))
}

de_cluster_ids <- map(1:length(kids), get_cluster_ids) %>%
  unlist()
```

```{r}
#| label: metadata-table
# Create a data frame with the sample IDs, cluster IDs and condition

gg_df <- data.frame(cluster_id = de_cluster_ids,
                    sample_id = de_samples)

gg_df <- left_join(gg_df, ei[, c("sample_id", "group_id")])

metadata <- gg_df %>%
  dplyr::select(cluster_id, sample_id, group_id)

metadata
```

## Subsetting dataset to cluster(s) of interest

```{r}
#| label: clusters-vector
# Generate vector of cluster IDs
clusters <- levels(factor(metadata$cluster_id))
clusters
```

::: {.callout-note}
Upstream writes `clusters <- levels(metadata$cluster_id)`. Since R 4.0,
`data.frame()` defaults to `stringsAsFactors = FALSE`, so `metadata$cluster_id`
is a character vector and `levels()` on it returns `NULL`. That would silently
make `clusters` empty and break every chunk below. Wrapping it in `factor()`
restores the intent of the original.
:::

We run the DE analysis on B cells, the first element of the vector.

```{r}
#| label: subset-b-cells
clusters[1]

# Subset the metadata to only the B cells
cluster_metadata <- metadata[which(metadata$cluster_id == clusters[1]), ]
head(cluster_metadata)

# Assign the rownames of the metadata to be the sample IDs
rownames(cluster_metadata) <- cluster_metadata$sample_id
head(cluster_metadata)

# Subset the counts to only the B cells
counts <- pb[[clusters[1]]]

cluster_counts <- data.frame(counts[, which(colnames(counts) %in% rownames(cluster_metadata))])

# Check that all of the row names of the metadata are the same and in the same
# order as the column names of the counts in order to use as input to DESeq2
all(rownames(cluster_metadata) == colnames(cluster_counts))
```

`group_id` has to be a factor for the `levels()` calls in the contrast code
below to work, so we make sure of that here.

```{r}
#| label: group-id-factor
cluster_metadata$group_id <- factor(cluster_metadata$group_id)
levels(cluster_metadata$group_id)
```

## Create DESeq2 object

The design formula should also include any other metadata columns whose
variation we want to regress out, for example batch, sex, or age. Here we only
have the comparison of interest, `group_id`.

```{r}
#| label: dds
dds <- DESeqDataSetFromMatrix(cluster_counts,
                              colData = cluster_metadata,
                              design = ~ group_id)
dds
```

## Quality control, sample level

Normalization and log2 transformation improve the distances used by unsupervised
clustering. DESeq2 uses the median of ratios method for normalization and a
regularized log transform (rlog) for sample-level QC, since it moderates the
variance across the mean.

::: {.callout-note}
The DESeq2 vignette suggests that large datasets, hundreds of samples, use the
variance stabilizing transformation `vst()` instead of `rlog()`, since `rlog()`
can be slow. With 16 samples here, `rlog()` is fine.
:::

### Principal component analysis

```{r}
#| label: pca
# Transform counts for data visualization
rld <- rlog(dds, blind = TRUE)

# Plot PCA
DESeq2::plotPCA(rld, intgroup = "group_id")
```

We expect a clean separation on PC1 by condition, which would say that the
condition of interest is the largest source of variation in the dataset.

### Hierarchical clustering

The heatmap shows the correlation of gene expression for all pairwise
combinations of samples. Most genes are not differentially expressed, so samples
generally correlate above 0.80. Values below 0.80 may flag an outlier or sample
contamination.

```{r}
#| label: sample-correlation-heatmap
# Extract the rlog matrix from the object and compute pairwise correlation values
rld_mat <- assay(rld)
rld_cor <- cor(rld_mat)

# Plot heatmap
pheatmap(rld_cor, annotation = cluster_metadata[, c("group_id"), drop = FALSE])
```

## Running DESeq2

DESeq2 models the raw counts using size factors to account for differences in
library depth, estimates gene-wise dispersions and shrinks them, then fits the
negative binomial model and tests with the Wald test or the likelihood ratio
test. All of that happens inside the single `DESeq()` call.

```{r}
#| label: run-deseq
# Run DESeq2 differential expression analysis
dds <- DESeq(dds)
```

```{r}
#| label: dispersion-plot
# Plot dispersion estimates
plotDispEsts(dds)
```

Dispersions should decrease with increasing mean and follow the line of best
fit.

## Results

We compare the stimulated group relative to the control.

```{r}
#| label: results
# Output results of Wald test for contrast for stim vs ctrl
levels(cluster_metadata$group_id)[2]
levels(cluster_metadata$group_id)[1]

contrast <- c("group_id", levels(cluster_metadata$group_id)[2], levels(cluster_metadata$group_id)[1])

# resultsNames(dds)
res <- results(dds,
               contrast = contrast,
               alpha = 0.05)

res <- lfcShrink(dds,
                 contrast = contrast,
                 res = res,
                 type = "normal")
res
```

::: {.callout-note}
Current DESeq2 defaults `lfcShrink()` to `type = "apeglm"`, which shrinks a
single `coef` and rejects a `contrast =` argument. Upstream relied on the older
default of `type = "normal"`, so we now pass `type = "normal"` explicitly. The
alternative, `type = "apeglm"` with `coef = resultsNames(dds)[2]`, is the modern
recommendation but changes the reported fold changes.
:::

We will write out the significant genes and make a few plots.

- Table of results for all genes.
- Table of results for significant genes, adjusted p below 0.05.
- Scatterplot of normalized expression of the top 20 most significant genes.
- Heatmap of all significant genes.
- Volcano plot of results.

The tutorial writes CSV files into a `results/` folder that it assumes you
created by hand when you set up the project, so we create it here.

```{r}
#| label: results-dir
results_dir <- here::here("slides", "wed_morning", "wed_morn_tutorials", "results")
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
results_dir
```

### Table of results for all genes

```{r}
#| label: res-table
# Turn the results object into a tibble for use with tidyverse functions
res_tbl <- res %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble()

# Check results output
res_tbl
```

```{r}
#| label: write-all-genes
# Write all results to file
write.csv(res_tbl,
          file.path(results_dir,
                    paste0(clusters[1], "_",
                           levels(cluster_metadata$group_id)[2], "_vs_",
                           levels(cluster_metadata$group_id)[1],
                           "_all_genes.csv")),
          quote = FALSE,
          row.names = FALSE)
```

::: {.callout-note}
Upstream builds this filename from `levels(cluster_metadata$sample)[2]`. There
is no `sample` column, only `sample_id`, so that expression is `NULL` and
`paste0()` drops it, producing a file named `B cells__vs__all_genes.csv` with
the condition names missing. We use `group_id` instead, which is what was meant.
Upstream's significant-genes call has a second typo on the same line,
`paste0("results", ...)` with no trailing slash, which writes into the working
directory rather than into `results/`. Both are fixed here.
:::

### Table of results for significant genes

```{r}
#| label: sig-res
# Set thresholds
padj_cutoff <- 0.05

# Subset the significant results
sig_res <- dplyr::filter(res_tbl, padj < padj_cutoff) %>%
  dplyr::arrange(padj)

# Check significant genes output
sig_res
```

```{r}
#| label: write-sig-genes
# Write significant results to file
write.csv(sig_res,
          file.path(results_dir,
                    paste0(clusters[1], "_",
                           levels(cluster_metadata$group_id)[2], "_vs_",
                           levels(cluster_metadata$group_id)[1],
                           "_sig_genes.csv")),
          quote = FALSE,
          row.names = FALSE)
```

### Scatterplot of normalized expression of top 20 most significant genes

This plot is a good check that we are reading the fold change direction
correctly.

```{r}
#| label: top20-scatter
## ggplot of top genes
normalized_counts <- counts(dds,
                            normalized = TRUE)

## Order results by padj values
top20_sig_genes <- sig_res %>%
  dplyr::arrange(padj) %>%
  dplyr::pull(gene) %>%
  head(n = 20)

top20_sig_norm <- data.frame(normalized_counts) %>%
  rownames_to_column(var = "gene") %>%
  dplyr::filter(gene %in% top20_sig_genes)

gathered_top20_sig <- top20_sig_norm %>%
  gather(colnames(top20_sig_norm)[2:length(colnames(top20_sig_norm))],
         key = "samplename", value = "normalized_counts")

gathered_top20_sig <- inner_join(ei[, c("sample_id", "group_id")],
                                 gathered_top20_sig,
                                 by = c("sample_id" = "samplename"))

## plot using ggplot2
ggplot(gathered_top20_sig) +
  geom_point(aes(x = gene,
                 y = normalized_counts,
                 color = group_id),
             position = position_jitter(w = 0.1, h = 0)) +
  scale_y_log10() +
  xlab("Genes") +
  ylab("log10 Normalized Counts") +
  ggtitle("Top 20 Significant DE Genes") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5))
```

### Heatmap of all significant genes

```{r}
#| label: sig-gene-heatmap
# Extract normalized counts for only the significant genes
sig_norm <- data.frame(normalized_counts) %>%
  rownames_to_column(var = "gene") %>%
  dplyr::filter(gene %in% sig_res$gene)

# Set a color palette
heat_colors <- brewer.pal(6, "YlOrRd")

# Run pheatmap using the metadata data frame for the annotation
pheatmap(sig_norm[, 2:length(colnames(sig_norm))],
         color = heat_colors,
         cluster_rows = TRUE,
         show_rownames = FALSE,
         annotation = cluster_metadata[, c("group_id", "cluster_id")],
         border_color = NA,
         fontsize = 10,
         scale = "row",
         fontsize_row = 10,
         height = 20)
```

### Volcano plot of results

```{r}
#| label: volcano
## Obtain logical vector where TRUE values denote padj values < 0.05 and fold
## change > 1.5 in either direction
res_table_thres <- res_tbl %>%
  mutate(threshold = padj < 0.05 & abs(log2FoldChange) >= 0.58)

## Volcano plot
ggplot(res_table_thres) +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold)) +
  ggtitle("Volcano plot of stimulated B cells relative to control") +
  xlab("log2 fold change") +
  ylab("-log10 adjusted p-value") +
  scale_y_continuous(limits = c(0, 50)) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))
```

## Useful scripts for running analyses on many different cell type clusters

### Script to run DESeq2 on all cell type clusters, Wald test

This runs DESeq2 on every cell type cluster, contrasting one level of the
condition of interest against another with the Wald test. Upstream writes its
output into `DESeq2/pairwise/`, and we keep that layout underneath
`results_dir`.

```{r}
#| label: wald-all-clusters
dir.create(file.path(results_dir, "DESeq2"), showWarnings = FALSE)
dir.create(file.path(results_dir, "DESeq2", "pairwise"), showWarnings = FALSE)

# Function to run DESeq2 and get results for all clusters
## x is index of cluster in clusters vector on which to run function
## A is the sample group to compare
## B is the sample group to compare against (base level)

get_dds_resultsAvsB <- function(x, A, B) {
  cluster_metadata <- metadata[which(metadata$cluster_id == clusters[x]), ]
  rownames(cluster_metadata) <- cluster_metadata$sample_id
  cluster_metadata$group_id <- factor(cluster_metadata$group_id)
  counts <- pb[[clusters[x]]]
  cluster_counts <- data.frame(counts[, which(colnames(counts) %in% rownames(cluster_metadata))])

  # all(rownames(cluster_metadata) == colnames(cluster_counts))

  dds <- DESeqDataSetFromMatrix(cluster_counts,
                                colData = cluster_metadata,
                                design = ~ group_id)

  # Transform counts for data visualization
  rld <- rlog(dds, blind = TRUE)

  # Plot PCA
  DESeq2::plotPCA(rld, intgroup = "group_id")
  ggsave(file.path(results_dir, paste0(clusters[x], "_specific_PCAplot.png")))

  # Extract the rlog matrix from the object and compute pairwise correlation values
  rld_mat <- assay(rld)
  rld_cor <- cor(rld_mat)

  # Plot heatmap
  png(file.path(results_dir, paste0(clusters[x], "_specific_heatmap.png")))
  pheatmap(rld_cor, annotation = cluster_metadata[, c("group_id"), drop = FALSE])
  dev.off()

  # Run DESeq2 differential expression analysis
  dds <- DESeq(dds)

  # Plot dispersion estimates
  png(file.path(results_dir, paste0(clusters[x], "_dispersion_plot.png")))
  plotDispEsts(dds)
  dev.off()

  # Output results of Wald test for contrast for A vs B
  contrast <- c("group_id", levels(cluster_metadata$group_id)[A], levels(cluster_metadata$group_id)[B])

  # resultsNames(dds)
  res <- results(dds,
                 contrast = contrast,
                 alpha = 0.05)

  res <- lfcShrink(dds,
                   contrast = contrast,
                   res = res,
                   type = "normal")
  # Set thresholds
  padj_cutoff <- 0.05

  # Turn the results object into a tibble for use with tidyverse functions
  res_tbl <- res %>%
    data.frame() %>%
    rownames_to_column(var = "gene") %>%
    as_tibble()

  write.csv(res_tbl,
            file.path(results_dir, "DESeq2", "pairwise",
                      paste0(clusters[x], "_",
                             levels(cluster_metadata$group_id)[A], "_vs_",
                             levels(cluster_metadata$group_id)[B],
                             "_all_genes.csv")),
            quote = FALSE,
            row.names = FALSE)

  # Subset the significant results
  sig_res <- dplyr::filter(res_tbl, padj < padj_cutoff) %>%
    dplyr::arrange(padj)

  write.csv(sig_res,
            file.path(results_dir, "DESeq2", "pairwise",
                      paste0(clusters[x], "_",
                             levels(cluster_metadata$group_id)[A], "_vs_",
                             levels(cluster_metadata$group_id)[B],
                             "_sig_genes.csv")),
            quote = FALSE,
            row.names = FALSE)

  ## ggplot of top genes
  normalized_counts <- counts(dds,
                              normalized = TRUE)

  ## Order results by padj values
  top20_sig_genes <- sig_res %>%
    dplyr::arrange(padj) %>%
    dplyr::pull(gene) %>%
    head(n = 20)

  top20_sig_norm <- data.frame(normalized_counts) %>%
    rownames_to_column(var = "gene") %>%
    dplyr::filter(gene %in% top20_sig_genes)

  gathered_top20_sig <- top20_sig_norm %>%
    gather(colnames(top20_sig_norm)[2:length(colnames(top20_sig_norm))],
           key = "samplename", value = "normalized_counts")

  gathered_top20_sig <- inner_join(ei[, c("sample_id", "group_id")],
                                   gathered_top20_sig,
                                   by = c("sample_id" = "samplename"))

  ## plot using ggplot2
  ggplot(gathered_top20_sig) +
    geom_point(aes(x = gene,
                   y = normalized_counts,
                   color = group_id),
               position = position_jitter(w = 0.1, h = 0)) +
    scale_y_log10() +
    xlab("Genes") +
    ylab("log10 Normalized Counts") +
    ggtitle("Top 20 Significant DE Genes") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(plot.title = element_text(hjust = 0.5))
  ggsave(file.path(results_dir, "DESeq2", "pairwise",
                   paste0(clusters[x], "_",
                          levels(cluster_metadata$group_id)[A], "_vs_",
                          levels(cluster_metadata$group_id)[B],
                          "_top20_DE_genes.png")))
}
```

Upstream calls this with a bare `map()`, which aborts the whole sweep at the
first cluster that fails. Small clusters such as Megakaryocytes can easily fail,
so we wrap each call so that one failure does not hide the other seven.

```{r}
#| label: wald-all-clusters-run
# Run the script on all clusters comparing stim condition relative to control
# condition
wald_outcomes <- map(1:length(clusters),
                     function(i) {
                       tryCatch(get_dds_resultsAvsB(i, A = 2, B = 1),
                                error = function(e) {
                                  message("Cluster ", clusters[i], " failed: ",
                                          conditionMessage(e))
                                  conditionMessage(e)
                                })
                     })
names(wald_outcomes) <- clusters
```

### Script to run DESeq2 on all cell type clusters, likelihood ratio test

```{r}
#| label: lrt-setup
# Likelihood ratio test
dir.create(file.path(results_dir, "DESeq2", "lrt"), showWarnings = FALSE)

# Create DESeq2Dataset object
clusters <- levels(factor(metadata$cluster_id))

metadata <- gg_df %>%
  dplyr::select(cluster_id, sample_id, group_id)

metadata$group <- paste0(metadata$cluster_id, "_", metadata$group_id) %>%
  factor()
```

```{r}
#| label: lrt-function
# DESeq2
library(DEGreport)
get_dds_LRTresults <- function(x) {

  cluster_metadata <- metadata[which(metadata$cluster_id == clusters[x]), ]
  rownames(cluster_metadata) <- cluster_metadata$sample_id
  cluster_metadata$group_id <- factor(cluster_metadata$group_id)
  counts <- pb[[clusters[x]]]
  cluster_counts <- data.frame(counts[, which(colnames(counts) %in% rownames(cluster_metadata))])

  # all(rownames(cluster_metadata) == colnames(cluster_counts))

  dds <- DESeqDataSetFromMatrix(cluster_counts,
                                colData = cluster_metadata,
                                design = ~ group_id)

  dds_lrt <- DESeq(dds, test = "LRT", reduced = ~ 1)

  # Extract results
  res_LRT <- results(dds_lrt)

  # Create a tibble for LRT results
  res_LRT_tb <- res_LRT %>%
    data.frame() %>%
    rownames_to_column(var = "gene") %>%
    as_tibble()

  # Save all results
  write.csv(res_LRT_tb,
            file.path(results_dir, "DESeq2", "lrt",
                      paste0(clusters[x], "_LRT_all_genes.csv")),
            quote = FALSE,
            row.names = FALSE)

  # Subset to return genes with padj < 0.05
  sigLRT_genes <- res_LRT_tb %>%
    dplyr::filter(padj < 0.05)

  # Save sig results
  write.csv(sigLRT_genes,
            file.path(results_dir, "DESeq2", "lrt",
                      paste0(clusters[x], "_LRT_sig_genes.csv")),
            quote = FALSE,
            row.names = FALSE)

  # Transform counts for data visualization
  rld <- rlog(dds_lrt, blind = TRUE)

  # Extract the rlog matrix from the object and compute pairwise correlation values
  rld_mat <- assay(rld)
  rld_cor <- cor(rld_mat)

  # Obtain rlog values for those significant genes
  cluster_rlog <- rld_mat[sigLRT_genes$gene, ]

  cluster_meta_sig <- cluster_metadata[which(rownames(cluster_metadata) %in% colnames(cluster_rlog)), ]

  # # Remove samples without replicates
  # cluster_rlog <- cluster_rlog[, -1]
  # cluster_metadata <- cluster_metadata[which(rownames(cluster_metadata) %in% colnames(cluster_rlog)), ]

  # Use the `degPatterns` function from the 'DEGreport' package to show gene
  # clusters across sample groups
  cluster_groups <- degPatterns(cluster_rlog, metadata = cluster_meta_sig,
                                time = "group_id", col = NULL)
  ggsave(file.path(results_dir, "DESeq2", "lrt",
                   paste0(clusters[x], "_LRT_DEgene_groups.png")))

  # Let's see what is stored in the `df` component
  write.csv(cluster_groups$df,
            file.path(results_dir, "DESeq2", "lrt",
                      paste0(clusters[x], "_LRT_DEgene_groups.csv")),
            quote = FALSE,
            row.names = FALSE)

  saveRDS(cluster_groups,
          file.path(results_dir, "DESeq2", "lrt",
                    paste0(clusters[x], "_LRT_DEgene_groups.rds")))
  save(dds_lrt, cluster_groups, res_LRT, sigLRT_genes,
       file = file.path(results_dir, "DESeq2", "lrt",
                        paste0(clusters[x], "_all_LRTresults.Rdata")))
}
```

```{r}
#| label: lrt-run
lrt_outcomes <- map(1:length(clusters),
                    function(i) {
                      tryCatch(get_dds_LRTresults(i),
                               error = function(e) {
                                 message("Cluster ", clusters[i], " failed: ",
                                         conditionMessage(e))
                                 conditionMessage(e)
                               })
                    })
names(lrt_outcomes) <- clusters
```

::: {.callout-note}
The LRT here compares `~ group_id` against `~ 1`, which with a two-level factor
is close to the Wald test above rather than a genuinely multi-group design. The
upstream lesson presents it as a template to adapt when there are three or more
groups.
:::

## Session information

```{r}
#| label: session-info
sessionInfo()
```
