---
title: "Differential expression testing in Seurat"
subtitle: "Local version, reads from the workshop data repo, no network access"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: true
  error: false
---

Local rework of `2_b_de_vignette.html`.

Upstream: <https://satijalab.org/seurat/articles/de_vignette>

::: {.callout-important}
## Status: transcribed from upstream, not yet executed

Every chunk below is transcribed from the upstream Seurat vignette and has **not
been run yet**. This first render is the test. `error: false` is set in the
YAML, so any chunk failure stops the render and is reported rather than being
printed into the document and passed over.
:::

::: {.callout-tip}
## Pairs with `2_a_pseudobulk_de_local.qmd`

Seurat's default `FindMarkers()` treats each **cell** as an independent
replicate. Cells from the same donor are not independent, so this inflates
significance and produces large numbers of false positives. The pseudobulk
section below, and the companion document `2_a_pseudobulk_de_local.qmd`, show
the fix. Aggregate to one profile per donor per cell type, then test at the
sample level.

See Squair et al. 2021, "Confronting false discoveries in single-cell
differential expression", Nature Communications,
<https://doi.org/10.1038/s41467-021-25960-2>
:::

::: {.callout-note}
## Data files this document expects

Upstream calls `SeuratData::LoadData("ifnb")`, which downloads and installs the
`ifnb.SeuratData` package. That is a network call, so it is replaced here by a
plain `readRDS()`.

The staging script must place these files in the data directory for
`2_b_de_vignette`:

- `ifnb.rds` , the `ifnb` `Seurat` object (interferon-beta stimulated human
  PBMCs, 2 conditions, about 14k cells), i.e. the result of
  `saveRDS(SeuratData::LoadData("ifnb"), "ifnb.rds")` run on a machine with
  network access.
- `ye1.ctrl.8.10.sm.best` and `ye2.stim.8.10.sm.best` , the demuxlet donor
  assignments that the upstream vignette pulls from GitHub. Optional. If they
  are absent the chunk falls back to the original URLs, which needs network.
:::

::: {.callout-warning}
## Optional packages needed for the alternative tests

`FindMarkers()` dispatches on `test.use`. Most options are covered by Seurat's
own dependencies, but several need extra packages installed separately:

- `test.use = "DESeq2"` needs Bioconductor **DESeq2**. It is used in the
  pseudobulk section below, so a missing DESeq2 breaks that whole section.
- `test.use = "MAST"` needs Bioconductor **MAST**. It is used in the last
  section.
- `test.use = "wilcox"` (the default) uses **presto** when it is installed, and
  `test.use = "wilcox_limma"` needs **limma**.

A missing package will surface as an error in this test render rather than
stopping it.
:::

## Setup

```{r}
#| label: setup
source(here::here("slides", "find_data_dir.R"))
data_dir <- find_data_dir("2_b_de_vignette")
report_data_dir("2_b_de_vignette")
```

## Load in the data

This vignette highlights some example workflows for performing differential
expression in Seurat. For demonstration purposes we use the interferon-beta
stimulated human PBMCs dataset (`ifnb`), which upstream is available via the
`SeuratData` package and here is read from a staged `.rds` file.

```{r}
#| label: libraries
library(Seurat)
library(ggplot2)
```

```{r}
#| label: load-data
ifnb <- readRDS(file.path(data_dir, "ifnb.rds"))
ifnb
```

## Perform default differential expression tests

The bulk of Seurat's differential expression features can be accessed through
the `FindMarkers()` function. By default, Seurat performs differential
expression (DE) testing based on the non-parametric Wilcoxon rank sum test. To
test for DE genes between two specific groups of cells, specify the `ident.1`
and `ident.2` parameters.

```{r}
#| label: default-de
# Normalize the data
ifnb <- NormalizeData(ifnb)

# Find DE features between CD16 Mono and CD14 Mono
Idents(ifnb) <- "seurat_annotations"
monocyte.de.markers <- FindMarkers(ifnb, ident.1 = "CD16 Mono", ident.2 = "CD14 Mono")
# view results
head(monocyte.de.markers)
```

The results data frame has the following columns:

- `p_val` : p-value (unadjusted)
- `avg_log2FC` : log fold-change of the average expression between the two
  groups. Positive values indicate that the feature is more highly expressed in
  the first group.
- `pct.1` : the percentage of cells where the feature is detected in the first
  group
- `pct.2` : the percentage of cells where the feature is detected in the second
  group
- `p_val_adj` : adjusted p-value, based on Bonferroni correction using all
  features in the dataset.

If the `ident.2` parameter is omitted or set to `NULL`, `FindMarkers()` will
test for differentially expressed features between the group specified by
`ident.1` and all other cells. Additionally, the parameter `only.pos` can be set
to `TRUE` to only search for positive markers, i.e. features that are more
highly expressed in the `ident.1` group.

```{r}
#| label: default-de-one-vs-rest
# Find differentially expressed features between CD16 Monocytes and all other cells, only
# search for positive markers
monocyte.de.markers <- FindMarkers(ifnb, ident.1 = "CD16 Mono", ident.2 = NULL, only.pos = TRUE)
# view results
head(monocyte.de.markers)
```

## Perform DE analysis within the same cell type across conditions

Since this dataset contains treatment information (control versus stimulated
with interferon-beta), we can also ask what genes change in different conditions
for cells of the same type. First we create a column in the `meta.data` slot to
hold both the cell type and treatment information and switch the current
`Idents` to that column. Then we use `FindMarkers()` to find the genes that are
different between control and stimulated CD14 monocytes.

```{r}
#| label: celltype-stim-de
ifnb$celltype.stim <- paste(ifnb$seurat_annotations, ifnb$stim, sep = "_")
Idents(ifnb) <- "celltype.stim"
mono.de <- FindMarkers(ifnb, ident.1 = "CD14 Mono_STIM", ident.2 = "CD14 Mono_CTRL", verbose = FALSE)
head(mono.de, n = 10)
```

However, the p-values obtained from this analysis should be interpreted with
caution, because these tests treat each cell as an independent replicate and
ignore inherent correlations between cells originating from the same sample.
Such analyses have been shown to find a large number of false positive
associations, as demonstrated by Squair et al. 2021
(<https://doi.org/10.1038/s41467-021-25960-2>), Zimmerman et al. 2021
(<https://doi.org/10.1038/s41467-021-21038-1>), Junttila et al. 2022
(<https://doi.org/10.1093/bib/bbac286>), and others. Below we show how
pseudobulking can be used to account for such within-sample correlation.

## Perform DE analysis after pseudobulking

To pseudobulk we use `AggregateExpression()` to sum together gene counts of all
the cells from the same sample for each cell type. This results in one gene
expression profile per sample and cell type. We can then perform DE analysis
using DESeq2 on the sample level. This treats the samples, rather than the
individual cells, as independent observations.

First we need to retrieve the sample information for each cell. This is not
loaded in the metadata. Upstream loads it from the GitHub repo of the source
data for the original paper. Here we prefer staged local copies and fall back to
the upstream URLs only if the local files are missing.

```{r}
#| label: add-donor-ids
# load the inferred sample IDs of each cell
ctrl_local <- file.path(data_dir, "ye1.ctrl.8.10.sm.best")
stim_local <- file.path(data_dir, "ye2.stim.8.10.sm.best")
ctrl_src <- if (file.exists(ctrl_local)) {
  ctrl_local
} else {
  url("https://raw.githubusercontent.com/yelabucsf/demuxlet_paper_code/master/fig3/ye1.ctrl.8.10.sm.best")
}
stim_src <- if (file.exists(stim_local)) {
  stim_local
} else {
  url("https://raw.githubusercontent.com/yelabucsf/demuxlet_paper_code/master/fig3/ye2.stim.8.10.sm.best")
}
ctrl <- read.table(ctrl_src, head = TRUE, stringsAsFactors = FALSE)
stim <- read.table(stim_src, head = TRUE, stringsAsFactors = FALSE)
info <- rbind(ctrl, stim)

# rename the cell IDs by substituting the '-' into '.'
info$BARCODE <- gsub(pattern = "\\-", replacement = "\\.", info$BARCODE)

# only keep the cells with high-confidence sample ID
info <- info[grep(pattern = "SNG", x = info$BEST), ]

# remove cells with duplicated IDs in both ctrl and stim groups
info <- info[!duplicated(info$BARCODE) & !duplicated(info$BARCODE, fromLast = TRUE), ]

# now add the sample IDs to ifnb
rownames(info) <- info$BARCODE
info <- info[, c("BEST"), drop = FALSE]
names(info) <- c("donor_id")
ifnb <- AddMetaData(ifnb, metadata = info)

# remove cells without donor IDs
ifnb$donor_id[is.na(ifnb$donor_id)] <- "unknown"
ifnb <- subset(ifnb, subset = donor_id != "unknown")
```

We can now perform pseudobulking (`AggregateExpression()`) based on the donor
IDs.

```{r}
#| label: pseudobulk
# pseudobulk the counts based on donor-condition-celltype
pseudo_ifnb <- AggregateExpression(ifnb, assays = "RNA", return.seurat = TRUE,
                                   group.by = c("stim", "donor_id", "seurat_annotations"))

# each 'cell' is a donor-condition-celltype pseudobulk profile
tail(Cells(pseudo_ifnb))

pseudo_ifnb$celltype.stim <- paste(pseudo_ifnb$seurat_annotations, pseudo_ifnb$stim, sep = "_")
```

Next we perform DE testing on the pseudobulk level for CD14 monocytes, and
compare it against the previous single-cell-level DE results.

```{r}
#| label: pseudobulk-de
Idents(pseudo_ifnb) <- "celltype.stim"

bulk.mono.de <- FindMarkers(object = pseudo_ifnb,
                            ident.1 = "CD14 Mono_STIM",
                            ident.2 = "CD14 Mono_CTRL",
                            test.use = "DESeq2")
head(bulk.mono.de, n = 15)
```

```{r}
#| label: compare-sc-bulk
# compare the DE P-values between the single-cell level and the pseudobulk level results
names(bulk.mono.de) <- paste0(names(bulk.mono.de), ".bulk")
bulk.mono.de$gene <- rownames(bulk.mono.de)

names(mono.de) <- paste0(names(mono.de), ".sc")
mono.de$gene <- rownames(mono.de)

merge_dat <- merge(mono.de, bulk.mono.de, by = "gene")
merge_dat <- merge_dat[order(merge_dat$p_val.bulk), ]

# Number of genes that are marginally significant in both, marginally significant
# only in bulk, and marginally significant only in single-cell
common <- merge_dat$gene[which(merge_dat$p_val.bulk < 0.05 &
                                 merge_dat$p_val.sc < 0.05)]
only_sc <- merge_dat$gene[which(merge_dat$p_val.bulk > 0.05 &
                                  merge_dat$p_val.sc < 0.05)]
only_bulk <- merge_dat$gene[which(merge_dat$p_val.bulk < 0.05 &
                                    merge_dat$p_val.sc > 0.05)]
print(paste0("# Common: ", length(common)))
print(paste0("# Only in single-cell: ", length(only_sc)))
print(paste0("# Only in bulk: ", length(only_bulk)))
```

Upstream reports 3,519 genes with evidence of differential expression (prior to
multiple hypothesis testing) in both analyses, 1,649 genes that only appear
differentially expressed in the single-cell analysis, and just 204 genes that
only appear differentially expressed in the bulk analysis. The single-cell
p-values are systematically smaller. We can investigate these discrepancies with
`VlnPlot()`.

First, the top genes that are differentially expressed in both analyses.

```{r}
#| label: violin-common
# create a new column to annotate sample-condition-celltype in the single-cell dataset
ifnb$donor_id.stim <- paste0(ifnb$stim, "-", ifnb$donor_id)

# generate violin plot
Idents(ifnb) <- "celltype.stim"
print(merge_dat[merge_dat$gene %in% common[1:2], c("gene", "p_val.sc", "p_val.bulk")])

VlnPlot(ifnb, features = common[1:2], idents = c("CD14 Mono_CTRL", "CD14 Mono_STIM"),
        group.by = "stim")
```

```{r}
#| label: violin-common-by-donor
VlnPlot(ifnb, features = common[1:2], idents = c("CD14 Mono_CTRL", "CD14 Mono_STIM"),
        group.by = "donor_id.stim", ncol = 1)
```

In both the pseudobulk and single-cell analyses, the p-values for these two
genes are astronomically small. For both of these genes, when just comparing all
stimulated CD14 monocytes to all control CD14 monocytes across samples, we see
much higher expression in the stimulated cells. When breaking down these cells
by sample, we continue to see consistently higher expression levels in the
stimulated samples compared to the control samples. In other words, this finding
is not driven by just one or two samples. Because of this consistency, we find
this signal in both analyses.

By contrast, we can examine examples of genes that are only DE under the
single-cell analysis.

```{r}
#| label: violin-only-sc
print(merge_dat[merge_dat$gene %in% c("SRGN", "HLA-DRA"), c("gene", "p_val.sc", "p_val.bulk")])

VlnPlot(ifnb, features = c("SRGN", "HLA-DRA"), idents = c("CD14 Mono_CTRL", "CD14 Mono_STIM"),
        group.by = "stim")
```

```{r}
#| label: violin-only-sc-by-donor
VlnPlot(ifnb, features = c("SRGN", "HLA-DRA"), idents = c("CD14 Mono_CTRL", "CD14 Mono_STIM"),
        group.by = "donor_id.stim", ncol = 1)
```

Here, SRGN and HLA-DRA both have very small p-values in the single-cell analysis
(on the order of $10^{-21}$ and $10^{-9}$), but much larger p-values around 0.18
in the pseudobulk analysis. While there appears to be a difference between
control and stimulated cells when ignoring sample information, the signal is
much weaker on the sample level, and we can see notable variability from sample
to sample.

## Perform DE analysis using alternative tests

Seurat supports many other DE tests. The currently supported values of
`test.use` are:

- `"wilcox"` : Wilcoxon rank sum test (default, using the `presto` package)
- `"wilcox_limma"` : Wilcoxon rank sum test (using the `limma` package)
- `"bimod"` : likelihood-ratio test for single cell feature expression
  (McDavid et al., Bioinformatics, 2013,
  <https://doi.org/10.1093/bioinformatics/bts714>)
- `"roc"` : standard AUC classifier
- `"t"` : Student's t-test
- `"poisson"` : likelihood ratio test assuming an underlying Poisson
  distribution. Use only for UMI-based datasets
- `"negbinom"` : likelihood ratio test assuming an underlying negative binomial
  distribution. Use only for UMI-based datasets
- `"LR"` : uses a logistic regression framework to determine differentially
  expressed genes. Constructs a logistic regression model predicting group
  membership based on each feature individually and compares this to a null
  model with a likelihood ratio test.
- `"MAST"` : GLM-framework that treats cellular detection rate as a covariate
  (Finak et al., Genome Biology, 2015,
  <https://doi.org/10.1186/s13059-015-0844-5>)
- `"DESeq2"` : DE based on a model using the negative binomial distribution
  (Love et al., Genome Biology, 2014,
  <https://doi.org/10.1186/s13059-014-0550-8>)

For MAST and DESeq2, please ensure that these packages are installed separately
in order to use them as part of Seurat. Once installed, the `test.use` parameter
can be used to specify which DE test to use.

```{r}
#| label: mast
# Test for DE features using the MAST package
Idents(ifnb) <- "seurat_annotations"
head(FindMarkers(ifnb, ident.1 = "CD14 Mono", ident.2 = "CD16 Mono", test.use = "MAST"))
```

## Session Info

```{r}
#| label: sessioninfo
sessionInfo()
```
