---
title: "Cell type composition: COVID PBMCs mapped to a healthy reference"
subtitle: "Local version, reads from the workshop data repo, no network access"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: true
  error: true
---

Local rework of `3_a_composition_covid.html`.

Upstream: <https://satijalab.org/seurat/articles/covid_sctmapping>

The reference is the annotated CITE-seq PBMC reference of Hao, Hao, et al.,
"Integrated analysis of multimodal single-cell data", Cell 2021,
<https://doi.org/10.1016/j.cell.2021.04.048>, distributed via Zenodo,
<https://zenodo.org/record/7779017>.

The query is a merged object built from three public COVID-19 PBMC studies in
the CZI cellxgene collections, <https://cellxgene.cziscience.com/collections>,
totalling 1,498,064 cells from 277 donors.

::: {.callout-important}
## Status: transcribed from upstream, not yet executed

Every chunk below is transcribed from the upstream Seurat vignette and has
**not** been run. This first render is the test. `error: true` is set in the
YAML on purpose, so that the render completes and lists *all* the failures
rather than stopping at the first one. Read the rendered HTML as a punch list.
:::

::: {.callout-warning}
## Teaching point: this vignette applies no statistical test

The upstream "Differential composition analysis" section compares cell type
proportions between healthy donors and COVID-19 patients by building a
contingency table of donor by predicted cell type, converting the counts to
per-donor relative abundances, and drawing two boxplots. It reports a reduction
in MAIT cells and an increase in plasmablasts among COVID-19 patients. It never
tests anything. There is no p-value, no confidence interval, and no correction
for the many cell types that were looked at.

That is a reasonable first look, and it is exactly the setup for
`3_b_composition_milo_local.qmd`, which does test. Make the gap explicit on the
slide rather than letting students infer that eyeballing boxplots is the method.
:::

::: {.callout-note}
## Local data files this document expects

Staged under `find_data_dir("3_a_composition_covid")`:

| File | Contents |
|---|---|
| `pbmc_multimodal_2023.rds` | The Seurat CITE-seq PBMC reference, with the `spca` reduction and the `wnn.umap` reduction model, plus the `celltype.l1` and `celltype.l2` metadata columns. Upstream path `/brahms/hartmana/vignette_data/pbmc_multimodal_2023.rds`, public source <https://zenodo.org/record/7779017>. |
| `merged_covid_object.rds` | The merged, un-normalized COVID PBMC query object from the three studies, built in the BPCells interaction vignette. Upstream path `/brahms/mollag/seurat_v5/vignette_data/merged_covid_object.rds`. |
| `merged_covid_object_mapped.rds` | Optional. The query object *after* `MapQuery`, so the slow mapping step can be skipped. See the mapping section for what it must contain. |

The staging script must use these exact filenames.
:::

## Setup

```{r}
#| label: setup
source(here::here("slides", "find_data_dir.R"))
data_dir <- find_data_dir("3_a_composition_covid")
report_data_dir("3_a_composition_covid")
```

```{r}
#| label: libraries
library(Seurat)
library(BPCells)
library(dplyr)
library(patchwork)
library(ggplot2)
options(future.globals.maxSize = 10 * 1024^3)
```

## Introduction: Reference mapping analysis in Seurat v5

Reference mapping identifies consistent cell type labels across studies, which
is what makes cross-dataset analysis possible. The individual datasets here are
manageable, but their aggregate does not fit in memory, so the query matrices
live on disk through BPCells and are never all loaded at once. The three studies
also used different annotation ontologies. Mapping to one common reference
replaces those with a single harmonized label set.

## Load the PBMC Reference Dataset and Query Datasets

Upstream reads both objects from the Satija lab filesystem. Locally we read the
staged copies. The query object holds the three studies as three separate
on-disk layers in its `RNA` assay.

```{r}
#| label: load
reference <- readRDS(file.path(data_dir, "pbmc_multimodal_2023.rds"))
object <- readRDS(file.path(data_dir, "merged_covid_object.rds"))
object <- NormalizeData(object, verbose = FALSE)
```

## Mapping

Anchors are found between reference and query in the precomputed supervised PCA
(`spca`), which is the recommended reduction for CITE-seq reference datasets.
In Seurat v5 a single `FindTransferAnchors` plus `MapQuery` call maps all three
datasets, because they all live in one query object.

::: {.callout-tip}
## Candidate for precomputing at staging time

This is the slow step, 1.5 million query cells against a 162,000-cell reference.
It is not what this session is about. If the workshop machine cannot afford the
runtime, stage a post-mapping object as `merged_covid_object_mapped.rds` and
have this section read that instead of recomputing.

The staged post-mapping object needs to carry:

- the `ref.umap` reduction, for the two `DimPlot` calls,
- the metadata columns `predicted.celltype.l1` and `predicted.celltype.l2`,
- the original metadata columns `cell_type`, `publication`, `donor_id`, and
  `disease`, which the split UMAPs and the composition analysis depend on,
- the `RNA` assay counts, if the pseudobulk differential expression section at
  the end is also to be run.

With that object staged, the reference itself is only needed for the mapping
step and could be dropped from the workshop data repo entirely.
:::

```{r}
#| label: mapping
anchor <- FindTransferAnchors(
  reference = reference,
  query = object,
  reference.reduction = 'spca',
  normalization.method = 'SCT',
  dims = 1:50)
object <- MapQuery(
  anchorset = anchor,
  query = object,
  reference = reference,
  refdata = list(
    celltype.l1 = "celltype.l1",
    celltype.l2 = "celltype.l2"
    ),
  reduction.model = 'wnn.umap'
)
```

## Explore the mapping results

All cells from the three studies are projected into the UMAP space defined by
the reference. Each cell gets an annotation at two levels of granularity,
`predicted.celltype.l1` and `predicted.celltype.l2`. The first plot shows the
original, study-specific annotations in `cell_type`. The second shows the
harmonized annotations produced by mapping. The contrast between the two is the
argument for mapping in the first place.

```{r}
#| label: dimplot-original-annotations
DimPlot(object, reduction = 'ref.umap', group.by = 'cell_type', alpha = 0.1, label = TRUE, split.by = 'publication', ncol = 3, label.size = 3) + NoLegend()
```

```{r}
#| label: dimplot-predicted-annotations
DimPlot(object, reduction = 'ref.umap', group.by = 'predicted.celltype.l2', alpha = 0.1, label = TRUE, split.by = 'publication', ncol = 3, label.size = 3) + NoLegend()
```

## Differential composition analysis

This is the section the workshop is here for.

The harmonized annotations are used to look for differences in the proportion of
each cell type between healthy individuals and COVID-19 patients. Upstream
reports a reduction in MAIT cells and an increase in plasmablasts among COVID-19
patients.

The recipe is: cross-tabulate donor against predicted cell type, drop donors
with 50 or fewer cells, divide each donor's row by its row total to get relative
abundances, then attach each donor's disease status from a second
cross-tabulation.

```{r}
#| label: composition-table
df_comp <- as.data.frame.matrix(table(object$donor_id, object$predicted.celltype.l2))
select.donors <- rownames(df_comp)[rowSums(df_comp)> 50]
df_comp <- df_comp[select.donors, ]
df_comp_relative <- sweep(x = df_comp, MARGIN = 1, STATS = rowSums(df_comp), FUN = '/')

df_disease <-  as.data.frame.matrix(table(object$donor_id, object$disease))[select.donors, ]

df_comp_relative$disease <- 'other'
df_comp_relative$disease[df_disease$normal!=0] <- 'normal'
df_comp_relative$disease[df_disease$`COVID-19`!=0] <- 'COVID-19'
df_comp_relative$disease <- factor(df_comp_relative$disease, levels = c('normal','COVID-19','other'))
df_comp_relative <- df_comp_relative[df_comp_relative$disease %in% c('normal','COVID-19'),]
```

The last line above is the vignette's own base-R subsetting. The `dplyr`
equivalent, for students who prefer it, is
`df_comp_relative <- dplyr::filter(df_comp_relative, disease %in% c('normal', 'COVID-19'))`.

Note what the disease assignment does. A donor is labelled `normal` if any of
its cells are tabulated under `normal`, then relabelled `COVID-19` if any of its
cells are tabulated under `COVID-19`, so `COVID-19` wins for any donor that has
both. Everything else stays `other` and is dropped on the last line.

```{r}
#| label: composition-boxplots
p1 <-  ggplot(data = df_comp_relative, mapping = aes(x = disease, y = MAIT, fill = disease)) +
  geom_boxplot(outlier.shape  = NA) +
  scale_fill_manual(values = c("#377eb8", "#e41a1c")) +
  xlab("") + ylab('relative abundance') +
  ggtitle('MAIT') +
  geom_jitter(color="black", size=0.4, alpha=0.9 ) +
  theme_bw() +
  theme( axis.title = element_text(size = 12),
         axis.text = element_text(size = 12),
         plot.title = element_text(size = 15, hjust = 0.5, face = "bold")
  )

p2 <-  ggplot(data = df_comp_relative, mapping = aes(x = disease, y = Plasmablast, fill = disease)) +
  geom_boxplot(outlier.shape  = NA) +
  scale_fill_manual(values = c("#377eb8", "#e41a1c")) +
  xlab("") + ylab('relative abundance') +
    ggtitle('Plasmablast') +
  geom_jitter(color="black", size=0.4, alpha=0.9 ) +
  theme_bw() +
  theme( axis.title = element_text(size = 12),
         axis.text = element_text(size = 12),
         plot.title = element_text(size = 15, hjust = 0.5, face = "bold")
  )

p1 + p2 + plot_layout(ncol = 2)
```

That is where the upstream section ends. Two boxplots, no test.

## Differential expression analysis

Beyond composition, an aggregation-based (pseudobulk) workflow explores
differential genes between healthy individuals and COVID-19 donors. All cells
sharing a cell type and a donor are aggregated with `AggregateExpression`. The
result is a Seurat object in which each "cell" is the pseudobulk profile of one
cell type in one individual.

```{r}
#| label: pseudobulk-aggregate
bulk <- AggregateExpression(object,
  return.seurat = TRUE,
  assays  = 'RNA',
  group.by = c("predicted.celltype.l2", "donor_id", "disease")
)

bulk <- subset(bulk, subset = disease %in% c('normal', 'COVID-19') )
bulk <- subset(bulk, subset = predicted.celltype.l2 !=  'Doublet')
bulk$disease <- factor(bulk$disease, levels = c('normal', 'COVID-19'))
```

With the pseudobulk object in hand, cell type-specific differential expression
between healthy and COVID-19 donors becomes possible. Upstream only visualizes a
few interferon-stimulated genes, which are typically upregulated during viral
infection.

```{r}
#| label: pseudobulk-isg-violins
#| fig-height: 12
p1 <- VlnPlot(
  object = bulk, features = 'IFI6', group.by = 'predicted.celltype.l2',
  split.by = 'disease', cols = c("#377eb8", "#e41a1c"))
p2 <- VlnPlot(
  object = bulk, features = c('ISG15'), group.by = 'predicted.celltype.l2',
  split.by = 'disease', cols = c("#377eb8", "#e41a1c"))
p3 <- VlnPlot(
  object = bulk, features = c('IFIT5'), group.by = 'predicted.celltype.l2',
  split.by = 'disease', cols = c("#377eb8", "#e41a1c"))
p1 + p2 + p3 + plot_layout(ncol = 1)
```

Note that this section, like the composition section, stops at pictures. The
violins are split by disease and no test is run on them either.

## Session Info

```{r}
#| label: sessioninfo
sessionInfo()
```
