---
title: "Cell type composition: COVID PBMCs, healthy versus infected"
subtitle: "Local version, reads from the workshop data repo, no network access"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: true
  error: false
---

Local rework of `3_a_composition_covid.html`.

Upstream: <https://satijalab.org/seurat/articles/covid_sctmapping>

::: {.callout-important}
## This version substitutes a smaller, public dataset

The upstream vignette maps a 1,498,064-cell COVID PBMC query onto a
162,000-cell reference. **That query object has no public download**, it lives
inside the Satija lab filesystem, so the vignette cannot be reproduced as
written.

We substitute a smaller public dataset that already carries the annotations the
composition analysis needs, so no reference and no mapping step are required:

Wilk et al. 2020, "A single-cell atlas of the peripheral immune response in
patients with severe COVID-19", Nature Medicine,
<https://doi.org/10.1038/s41591-020-0944-y>, distributed by CELLxGENE,
<https://cellxgene.cziscience.com/collections/b9fc3d70-5a72-4479-a046-c2cc1ab19efc>.
41,305 PBMCs, 13 donors (7 COVID-19, 6 healthy), 26 cell types.

The teaching point is unchanged, and the headline result is the same: MAIT cells
are reduced and plasmablasts increased in COVID-19.
:::

::: {.callout-warning}
## Teaching point: this analysis applies no statistical test

Like the upstream vignette, the composition analysis below compares cell type
proportions between healthy and COVID-19 donors with a table and boxplots. It
reports reduced MAIT cells and increased plasmablasts. It never tests anything.
No p-value, no confidence interval, no correction for the many cell types looked
at.

That is a reasonable first look, and it is exactly the setup for
`3_b_composition_milo_local.qmd`, which does test. Make the gap explicit on the
slide rather than letting students infer that eyeballing boxplots is the method.
An optional propeller block at the end shows how little it takes to add a real
test.
:::

::: {.callout-note}
## Local data file this document expects

Staged under `find_data_dir("3_a_composition_covid")` by
`staging/stage_3_a_composition_covid.R`:

- `covid_pbmc_wilk.rds`, a `Seurat` object of raw counts with the metadata
  columns `donor_id`, `disease` (levels `normal` and `COVID-19`), and
  `cell_type`.
:::

## Setup

```{r}
#| label: setup
source(here::here("slides", "find_data_dir.R"))
data_dir <- find_data_dir("3_a_composition_covid")
report_data_dir("3_a_composition_covid")
```

```{r}
#| label: libraries
library(Seurat)
library(dplyr)
library(patchwork)
library(ggplot2)
```

## What upstream does, and why we skip it

The upstream vignette spends its first half on **reference mapping**: it projects
a query of COVID PBMCs onto an annotated healthy PBMC reference, transferring a
harmonized label (`predicted.celltype.l2`) onto every query cell. That is worth
understanding, and `integration_mapping.html` in this folder covers the mechanics.

We skip it here for one reason: the CELLxGENE distribution of the Wilk data
already carries curated `cell_type` labels, so the annotations we would have
obtained by mapping are simply present in the object. That lets the tutorial get
straight to the composition question, which is what this session is about.

## Load the data

```{r}
#| label: load
object <- readRDS(file.path(data_dir, "covid_pbmc_wilk.rds"))
object

table(object$disease)
sort(table(object$cell_type), decreasing = TRUE)
```

## A quick standard pipeline for a UMAP

The staged object is raw counts only, with no reduction. We run a compact
standard Seurat pipeline so there is a UMAP to look at. None of this is the point
of the session, it is just to see the annotations in a low-dimensional space.

```{r}
#| label: pipeline
object <- NormalizeData(object)
object <- FindVariableFeatures(object, nfeatures = 2000)
object <- ScaleData(object)
object <- RunPCA(object, npcs = 30)
object <- RunUMAP(object, dims = 1:30)
```

```{r}
#| label: umap
#| fig-width: 11
#| fig-height: 6
DimPlot(object, reduction = "umap", group.by = "cell_type",
        label = TRUE, repel = TRUE, label.size = 3) + NoLegend()
```

```{r}
#| label: umap-by-disease
DimPlot(object, reduction = "umap", group.by = "disease",
        cols = c("#377eb8", "#e41a1c"))
```

## Differential composition analysis

This is the section the workshop is here for.

We look for differences in the proportion of each cell type between healthy and
COVID-19 donors. The recipe: cross-tabulate donor against cell type, drop donors
with 50 or fewer cells, divide each donor's row by its row total to get relative
abundances, then attach each donor's disease status.

```{r}
#| label: composition-table
df_comp <- as.data.frame.matrix(table(object$donor_id, object$cell_type))
select.donors <- rownames(df_comp)[rowSums(df_comp) > 50]
df_comp <- df_comp[select.donors, ]
df_comp_relative <- sweep(x = df_comp, MARGIN = 1,
                          STATS = rowSums(df_comp), FUN = "/")

# This dataset splits plasmablasts into IgA and IgG subtypes. Collapse them into
# one Plasmablast column so the comparison matches the upstream story.
pb_cols <- grep("plasmablast", colnames(df_comp_relative),
                value = TRUE, ignore.case = TRUE)
df_comp_relative$Plasmablast <- rowSums(df_comp_relative[, pb_cols, drop = FALSE])

# The MAIT population is spelled out in this dataset's ontology labels.
mait_col <- "mucosal-associated invariant T cell"

df_disease <- as.data.frame.matrix(table(object$donor_id, object$disease))[select.donors, ]

df_comp_relative$disease <- "other"
df_comp_relative$disease[df_disease$normal != 0] <- "normal"
df_comp_relative$disease[df_disease$`COVID-19` != 0] <- "COVID-19"
df_comp_relative$disease <- factor(df_comp_relative$disease,
                                   levels = c("normal", "COVID-19", "other"))
df_comp_relative <- df_comp_relative |>
  dplyr::filter(disease %in% c("normal", "COVID-19"))
```

Note what the disease assignment does. A donor is labelled `normal` if any of its
cells fall under `normal`, then relabelled `COVID-19` if any fall under
`COVID-19`, so `COVID-19` wins for a donor with both. In this dataset each donor
is one arm, so that is moot here, but it is the vignette's own logic and worth
keeping visible.

```{r}
#| label: composition-boxplots
plot_abundance <- function(df, column, title) {
  ggplot(df, aes(x = disease, y = .data[[column]], fill = disease)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(color = "black", size = 0.4, alpha = 0.9) +
    scale_fill_manual(values = c("#377eb8", "#e41a1c")) +
    xlab("") + ylab("relative abundance") + ggtitle(title) +
    theme_bw() +
    theme(axis.title = element_text(size = 12),
          axis.text = element_text(size = 12),
          plot.title = element_text(size = 15, hjust = 0.5, face = "bold"))
}

p1 <- plot_abundance(df_comp_relative, mait_col, "MAIT")
p2 <- plot_abundance(df_comp_relative, "Plasmablast", "Plasmablast")
p1 + p2 + plot_layout(ncol = 2)
```

MAIT cells are reduced and plasmablasts increased in COVID-19, the same result
the upstream vignette reports. **And, like upstream, that is where it stops. Two
boxplots, no test.**

## Optional: add the test the vignette omits

`propeller` turns these per-donor proportions into a real result with an FDR
attached, in about three lines. It is a good place to make the point that the
boxplots above are a picture, not an analysis.

Phipson et al. 2022, "propeller: testing for differences in cell type
proportions in single cell data", Bioinformatics,
<https://doi.org/10.1093/bioinformatics/btac582>

```{r}
#| label: propeller
#| eval: false
# Not run by default, needs the speckle package. Remove eval:false to run it.
library(speckle)
propeller(clusters = object$cell_type,
          sample   = object$donor_id,
          group    = object$disease)
```

## Differential expression analysis

Beyond composition, an aggregation-based (pseudobulk) workflow looks at
differential genes between healthy and COVID-19 donors. All cells sharing a cell
type and a donor are aggregated with `AggregateExpression`, giving one profile
per cell type per donor.

```{r}
#| label: pseudobulk-aggregate
bulk <- AggregateExpression(object,
  return.seurat = TRUE,
  assays  = "RNA",
  group.by = c("cell_type", "donor_id", "disease"))

bulk <- subset(bulk, subset = disease %in% c("normal", "COVID-19"))
bulk$disease <- factor(bulk$disease, levels = c("normal", "COVID-19"))
```

With the pseudobulk object in hand, we can look at interferon-stimulated genes,
which are typically upregulated during viral infection. As upstream does, this
only visualizes them, it does not test.

```{r}
#| label: pseudobulk-isg-violins
#| fig-height: 12
p1 <- VlnPlot(bulk, features = "IFI6", group.by = "cell_type",
              split.by = "disease", cols = c("#377eb8", "#e41a1c"))
p2 <- VlnPlot(bulk, features = "ISG15", group.by = "cell_type",
              split.by = "disease", cols = c("#377eb8", "#e41a1c"))
p3 <- VlnPlot(bulk, features = "IFIT3", group.by = "cell_type",
              split.by = "disease", cols = c("#377eb8", "#e41a1c"))
p1 + p2 + p3 + plot_layout(ncol = 1)
```

Like the composition section, this stops at pictures. The violins are split by
disease and no test is run on them either.

## Session info

```{r}
#| label: sessioninfo
sessionInfo()
```
