---
title: "Differential abundance testing with Milo"
subtitle: "Local version, reads from the workshop data repo, no network access"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: true
  error: true
---

Local rework of `3_b_composition_milo.html`.

Upstream: <https://bioconductor.org/packages/release/bioc/vignettes/miloR/inst/doc/milo_gastrulation.html>

Method: Dann et al. 2022, "Differential abundance testing on single-cell data
using k-nearest neighbor graphs", Nature Biotechnology,
<https://doi.org/10.1038/s41587-021-01033-z>

Dataset: Pijuan-Sala et al. 2019, "A single-cell molecular map of mouse
gastrulation and early organogenesis", Nature,
<https://doi.org/10.1038/s41586-019-0933-9>

::: {.callout-important}
## Status: transcribed from upstream, first render is a test

The code below is transcribed from the upstream vignette, with the
`EmbryoAtlasData()` download replaced by a read from the staged data repo. It has
not been executed yet. `error: true` is set in the YAML on purpose, so a test
render reports every failure rather than stopping at the first one.
:::

::: {.callout-caution}
## Do not put the download back

Upstream calls `EmbryoAtlasData()`, which pulls the mouse gastrulation atlas
through `ExperimentHub`. `ExperimentHub` caches per user, so twenty students on
twenty accounts means twenty multi-GB downloads at the same moment. That call is
deliberately absent here. It lives in `staging/stage_3_b_composition_milo.R` and
is run once, ahead of time.
:::

## Setup

```{r}
#| label: setup
source(here::here("slides", "find_data_dir.R"))
data_dir <- find_data_dir("3_b_composition_milo")
report_data_dir("3_b_composition_milo")
```

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(miloR)
  library(SingleCellExperiment)
  library(scater)
  library(scran)
  library(dplyr)
  library(patchwork)
  library(ggplot2)
})
```

## Load data

Eight samples, four at stage E7.0 and four at E7.5, so the contrast is
developmental time. The object already carries a `pca.corrected` reduction, built
after fastMNN batch correction, and cells with an all-NA reduction have already
been dropped by the staging script.

::: {.callout-note}
## Staged file

`embryo_atlas_subset.rds`, a `SingleCellExperiment`. Produced by
`staging/stage_3_b_composition_milo.R`, which is where the `EmbryoAtlasData()`
call and the NA filter live.
:::

```{r}
#| label: load
embryo_data <- readRDS(file.path(data_dir, "embryo_atlas_subset.rds"))
embryo_data
```

## Visualize the data

```{r}
#| label: umap
embryo_data <- runUMAP(embryo_data, dimred = "pca.corrected", name = "umap")
plotReducedDim(embryo_data, colour_by = "stage", dimred = "umap")
```

## Create a Milo object

`Milo` extends `SingleCellExperiment` with slots for the neighbourhoods.

```{r}
#| label: milo-object
embryo_milo <- Milo(embryo_data)
embryo_milo
```

## Construct the KNN graph

`k` affects the power of the test, because each neighbourhood needs enough cells
per sample to estimate between-replicate variance. Too large and it over-smooths.
Start with the same `k` you used for clustering and UMAP.

```{r}
#| label: build-graph
embryo_milo <- buildGraph(embryo_milo, k = 30, d = 30,
                          reduced.dim = "pca.corrected")
```

## Define representative neighbourhoods

`prop` is the fraction of cells sampled as neighbourhood indices. `refined = TRUE`
is the recommended sampling scheme.

```{r}
#| label: make-nhoods
embryo_milo <- makeNhoods(embryo_milo, prop = 0.1, k = 30, d = 30,
                          refined = TRUE, reduced_dims = "pca.corrected")
```

Rule of thumb: mean neighbourhood size should be above 5 times the number of
samples. If it is lower, raise `k` or `prop` and rebuild.

```{r}
#| label: nhood-size-hist
plotNhoodSizeHist(embryo_milo)
```

## Count cells in neighbourhoods

Milo uses the variation in cell numbers between replicates of the same condition
to test for differential abundance, so the counting is by sample.

```{r}
#| label: count-cells
embryo_milo <- countCells(embryo_milo,
                          meta.data = as.data.frame(colData(embryo_milo)),
                          sample = "sample")
head(nhoodCounts(embryo_milo))
```

## Defining the experimental design

Testing is a negative binomial GLM from edgeR. The design matrix maps each sample
to its condition, here `stage`, plus `sequencing.batch` as a known technical
covariate.

::: {.callout-tip}
## Worth lingering on

This is the same design-matrix idea as DESeq2 in `2_a_pseudobulk_de_local.qmd`,
applied to neighbourhood counts instead of gene counts.
:::

```{r}
#| label: design
embryo_design <- data.frame(colData(embryo_milo))[,
  c("sample", "stage", "sequencing.batch")]

# Batch is stored as an integer, and must be a factor.
embryo_design$sequencing.batch <- as.factor(embryo_design$sequencing.batch)
embryo_design <- dplyr::distinct(embryo_design)
rownames(embryo_design) <- embryo_design$sample

embryo_design
```

## Computing neighbourhood connectivity

Milo adapts the spatial FDR correction from cydar, weighting each p-value by the
reciprocal of the kth nearest neighbour distance. Those distances have to be
stored first.

```{r}
#| label: nhood-distance
# This is the slowest step in the workflow.
embryo_milo <- calcNhoodDistance(embryo_milo, d = 30,
                                 reduced.dim = "pca.corrected")
```

## Testing

```{r}
#| label: test
da_results <- testNhoods(embryo_milo,
                         design = ~ sequencing.batch + stage,
                         design.df = embryo_design,
                         reduced.dim = "pca.corrected")
head(da_results)
```

```{r}
#| label: top-hits
da_results |>
  dplyr::arrange(SpatialFDR) |>
  head(20)
```

## Inspecting the results

```{r}
#| label: pvalue-hist
ggplot(da_results, aes(PValue)) + geom_histogram(bins = 50)
```

```{r}
#| label: volcano
ggplot(da_results, aes(logFC, -log10(SpatialFDR))) +
  geom_point() +
  geom_hline(yintercept = 1)   # 10% FDR
```

Each node is a neighbourhood, edges show how many cells two neighbourhoods share,
and node position comes from the index cell's position in the single-cell UMAP.
Significant neighbourhoods are coloured by log fold change.

```{r}
#| label: nhood-graph
embryo_milo <- buildNhoodGraph(embryo_milo)

umap_pl <- plotReducedDim(embryo_milo, dimred = "umap", colour_by = "stage",
                          text_by = "celltype", text_size = 3,
                          point_size = 0.5) +
  guides(fill = "none")

nh_graph_pl <- plotNhoodGraphDA(embryo_milo, da_results, layout = "umap",
                                alpha = 0.1)

umap_pl + nh_graph_pl + plot_layout(guides = "collect")
```

## Differential abundance by cell type

Label each neighbourhood by its most abundant cell type, and record what fraction
of the neighbourhood carries that label.

```{r}
#| label: annotate
da_results <- annotateNhoods(embryo_milo, da_results, coldata_col = "celltype")
head(da_results)
```

```{r}
#| label: celltype-fraction
ggplot(da_results, aes(celltype_fraction)) + geom_histogram(bins = 50)
```

```{r}
#| label: relabel-mixed
# Neighbourhoods that are a blend of cell types get called Mixed rather than
# being forced into one label.
da_results$celltype <- ifelse(da_results$celltype_fraction < 0.7,
                              "Mixed", da_results$celltype)
```

```{r}
#| label: beeswarm
plotDAbeeswarm(da_results, group.by = "celltype")
```

::: {.callout-tip}
## The teaching payoff

Early types such as epiblast and primitive streak are enriched at the earlier
stage, ectoderm later. Note the "Mixed" neighbourhoods that also come up
significant. Those are transitional states, and they are exactly what a
cluster-level test would have averaged away. That is the argument for Milo over
`propeller` or `scCODA` in continuum data.
:::

## Finding markers of DA populations

`findNhoodGroupMarkers` runs a one-versus-all differential expression test for a
group of neighbourhoods. The grouping has to be defined first, in a `NhoodGroup`
column.

```{r}
#| label: lognorm
embryo_milo <- logNormCounts(embryo_milo)
```

```{r}
#| label: custom-grouping
# Group the significant neighbourhoods that shifted in one direction against
# everything else. subset.row restricts to 10 genes for illustration, exactly as
# upstream does. Drop that argument for a real analysis.
da_results$NhoodGroup <- as.numeric(da_results$SpatialFDR < 0.1 &
                                    da_results$logFC < 0)
da_nhood_markers <- findNhoodGroupMarkers(
  embryo_milo, da_results,
  subset.row = rownames(embryo_milo)[1:10])
head(da_nhood_markers)
```

## Session info

```{r}
#| label: sessioninfo
sessionInfo()
```
