---
title: "scATAC-seq vignette (Signac PBMC)"
format: html
---

Adapted from <https://stuartlab.org/signac/articles/pbmc_vignette.html>.

```{r}
library(Signac)
library(Seurat)
library(GenomicRanges)
library(ggplot2)
library(patchwork)
library(GenomeInfoDb) # for standardChromosomes()
library(biovizBase)   # required indirectly by Signac plotting
library(AnnotationHub)

# Where the pre-staged workshop data lives. See slides/find_data_dir.R.
source(here::here("slides", "find_data_dir.R"))
data_dir <- find_data_dir()
```

### Checkpoint files

::: {.callout-important}
## Nothing here is recomputed during class

Timed on 2026-07-26, on a 30 GB laptop. This tutorial cannot be run from
scratch in a class slot, and it does not fit in 16 GB of RAM:

| Step | Cost from scratch |
|------|-------------------|
| `CreateChromatinAssay` + `NucleosomeSignal` + `TSSEnrichment` | tens of minutes, and it streams the 2.6 GB fragments file |
| `GeneActivity` | tens of minutes |
| `FindTransferAnchors` / `TransferData` | tens of minutes |
| `FindMarkers(test.use = "wilcox")` over ~100k peaks | **hours without `presto`** |
| `FindMarkers(test.use = "LR")` over ~100k peaks | **hours, one logistic regression per peak** |

Peak memory during the fragment and `GeneActivity` steps was about 15 GB,
which will not survive on a 16 GB student machine alongside RStudio and a
browser.

So every expensive step is checkpointed to `.rds` in the workshop data
directory, and in class each one is loaded, not computed. Loading is seconds
and the loaded object is a few GB. The `eval: false` staging blocks are what
produced those checkpoints, run once ahead of the workshop.

Two things must be true before the session, or students will silently fall
into the recompute path:

1. The four `pbmc_atac_ck*.rds` files are present in the data directory.
2. `presto` is installed. Without it `FindMarkers` falls back to a much
   slower implementation. `remotes::install_github("immunogenomics/presto")`.
:::

Checkpoints are read from the workshop data directory. If one is missing, the
tutorial recomputes it and writes the result to `gitignore/` locally rather
than to the shared data directory, which students will not have write access
to.

```{r}
# Read checkpoints from the staged data directory, write any that have to be
# recomputed into the local gitignore/ directory.
ck_read_dir <- data_dir
ck_write_dir <- here::here("gitignore")
dir.create(ck_write_dir, showWarnings = FALSE)

ck_qc      <- in_data_dir("pbmc_atac_ck1_qc.rds")       # post-QC, pre-subset
ck_dimred  <- in_data_dir("pbmc_atac_ck2_dimred.rds")   # post-UMAP + GeneActivity
ck_labeled <- in_data_dir("pbmc_atac_ck3_labeled.rds")  # post-label-transfer + relabel
ck_markers <- in_data_dir("pbmc_atac_ck4_markers.rds")  # da_peaks + lr_da_peaks

# Say plainly which of these will be loaded and which will be recomputed, so a
# slow run is diagnosed in the first seconds rather than an hour in.
for (ck in c(ck_qc, ck_dimred, ck_labeled, ck_markers)) {
  message(
    if (file.exists(ck)) "will LOAD      " else "will RECOMPUTE ",
    basename(ck)
  )
}
if (!requireNamespace("presto", quietly = TRUE)) {
  message(
    "presto is NOT installed. FindMarkers will be slow if it has to run."
  )
}
```

The large 10x Genomics ATAC files (h5 peak matrix, fragments.tsv.gz +
index, ~3 GB total) and the reference scRNA-seq object (`pbmc_10k_v3.rds`)
belong in the workshop data directory, not in this repository. As of
2026-07-26 only the h5 peak matrix and `singlecell.csv.gz` are staged there,
so the block below is what stages the rest. Run it ahead of the workshop, never
during a session.

```{r}
#| eval: false
# Run once. ~3 GB total. Writes into the workshop data directory.
base_url <- "https://cf.10xgenomics.com/samples/cell-atac/2.1.0/10k_pbmc_ATACv2_nextgem_Chromium_Controller/"
for (f in c(
  "10k_pbmc_ATACv2_nextgem_Chromium_Controller_filtered_peak_bc_matrix.h5",
  "10k_pbmc_ATACv2_nextgem_Chromium_Controller_singlecell.csv",
  "10k_pbmc_ATACv2_nextgem_Chromium_Controller_fragments.tsv.gz",
  "10k_pbmc_ATACv2_nextgem_Chromium_Controller_fragments.tsv.gz.tbi"
)) {
  curl::curl_download(paste0(base_url, f), destfile = file.path(data_dir, f))
}
# Pre-processed scRNA-seq Seurat object used for label transfer.
curl::curl_download(
  "https://signac-objects.s3.amazonaws.com/pbmc_10k_v3.rds",
  destfile = file.path(data_dir, "pbmc_10k_v3.rds")
)
```

### Build the Signac object and compute QC metrics

This block builds the Seurat/Signac object from the 10x outputs, keeps
standard chromosomes only, attaches Ensembl 98 gene annotations, and
computes the nucleosome signal, TSS enrichment, % reads in peaks, and
blacklist ratio per cell. The resulting object is cached to `ck_qc`.

```{r}
if (file.exists(ck_qc)) {
  pbmc <- readRDS(ck_qc)
  message("Loaded pbmc from ", ck_qc)
} else {
  counts <- Read10X_h5(
    filename = file.path(
      data_dir,
      "10k_pbmc_ATACv2_nextgem_Chromium_Controller_filtered_peak_bc_matrix.h5"
    )
  )

  metadata <- read.csv(
    file = file.path(
      data_dir,
      "10k_pbmc_ATACv2_nextgem_Chromium_Controller_singlecell.csv.gz"
    ),
    header = TRUE,
    row.names = 1
  )

  chrom_assay <- CreateChromatinAssay(
    counts = counts,
    sep = c(":", "-"),
    fragments = file.path(
      data_dir,
      "10k_pbmc_ATACv2_nextgem_Chromium_Controller_fragments.tsv.gz"
    ),
    min.cells = 10,
    min.features = 200
  )

  pbmc <- CreateSeuratObject(
    counts = chrom_assay,
    assay = "peaks",
    meta.data = metadata
  )

  # Keep only standard chromosomes.
  peaks.keep <- seqnames(granges(pbmc)) %in% standardChromosomes(granges(pbmc))
  pbmc <- pbmc[as.vector(peaks.keep), ]

  # Pull Ensembl 98 EnsDb (Homo sapiens) and the ENCODE hg38 blacklist
  # from AnnotationHub. `AH75011` is the accession for the Ensembl 98
  # EnsDb (a SQLite-backed Ensembl annotation database mapped to GRCh38).
  ah <- AnnotationHub()
  ensdb_v98 <- ah[["AH75011"]]
  blacklist_regions <- ah[["AH107305"]] # ENCODE blacklist for hg38

  annotations <- GetGRangesFromEnsDb(ensdb = ensdb_v98)
  seqlevels(annotations) <- paste0("chr", seqlevels(annotations))
  genome(annotations) <- "hg38"
  Annotation(pbmc) <- annotations

  pbmc <- NucleosomeSignal(object = pbmc)
  pbmc <- TSSEnrichment(object = pbmc)
  pbmc$pct_reads_in_peaks <- pbmc$peak_region_fragments / pbmc$passed_filters * 100
  pbmc$blacklist_ratio <- FractionCountsInRegion(
    object = pbmc, assay = "peaks", regions = blacklist_regions
  )

  saveRDS(pbmc, ck_qc)
  message("Saved checkpoint ", ck_qc)
}

pbmc
```

```{r}
pbmc[["peaks"]]
```

```{r}
granges(pbmc)
```

### QC plots

```{r}
DensityScatter(pbmc, x = "nCount_peaks", y = "TSS.enrichment", log_x = TRUE, quantiles = TRUE)
```

```{r}
pbmc$nucleosome_group <- ifelse(pbmc$nucleosome_signal > 4, "NS > 4", "NS < 4")
FragmentHistogram(object = pbmc, group.by = "nucleosome_group")
```

```{r}
VlnPlot(
  object = pbmc,
  features = c("nCount_peaks", "TSS.enrichment", "blacklist_ratio", "nucleosome_signal", "pct_reads_in_peaks"),
  pt.size = 0.1,
  ncol = 5
)
```

### Filter low-quality cells, normalize, dim-reduce, build gene-activity assay

Cached to `ck_dimred`.

```{r}
if (file.exists(ck_dimred)) {
  pbmc <- readRDS(ck_dimred)
  message("Loaded pbmc from ", ck_dimred)
} else {
  pbmc <- subset(
    x = pbmc,
    subset = nCount_peaks > 9000 &
      nCount_peaks < 100000 &
      pct_reads_in_peaks > 40 &
      blacklist_ratio < 0.01 &
      nucleosome_signal < 4 &
      TSS.enrichment > 4
  )

  pbmc <- RunTFIDF(pbmc)
  pbmc <- FindTopFeatures(pbmc, min.cutoff = "q0")
  pbmc <- RunSVD(pbmc)

  pbmc <- RunUMAP(object = pbmc, reduction = "lsi", dims = 2:30)
  pbmc <- FindNeighbors(object = pbmc, reduction = "lsi", dims = 2:30)
  pbmc <- FindClusters(object = pbmc, verbose = FALSE, algorithm = 3)

  gene.activities <- GeneActivity(pbmc)
  pbmc[["RNA"]] <- CreateAssayObject(counts = gene.activities)
  pbmc <- NormalizeData(
    object = pbmc, assay = "RNA",
    normalization.method = "LogNormalize",
    scale.factor = median(pbmc$nCount_RNA)
  )

  saveRDS(pbmc, ck_dimred)
  message("Saved checkpoint ", ck_dimred)
}

pbmc
```

```{r}
DepthCor(pbmc)
```

```{r}
DimPlot(object = pbmc, label = TRUE) + NoLegend()
```

```{r}
DefaultAssay(pbmc) <- "RNA"

FeaturePlot(
  object = pbmc,
  features = c("MS4A1", "CD3D", "LEF1", "NKG7", "TREM1", "LYZ"),
  pt.size = 0.1,
  max.cutoff = "q95",
  ncol = 3
)
```

### Integrate with scRNA-seq via label transfer

Cached to `ck_labeled` (the relabeled `pbmc`). The reference `pbmc_rna` is
loaded fresh because it is a separate object.

```{r}
pbmc_rna <- readRDS(file.path(data_dir, "pbmc_10k_v3.rds"))
pbmc_rna <- UpdateSeuratObject(pbmc_rna)
```

```{r}
if (file.exists(ck_labeled)) {
  pbmc <- readRDS(ck_labeled)
  message("Loaded pbmc from ", ck_labeled)
} else {
  transfer.anchors <- FindTransferAnchors(
    reference = pbmc_rna,
    query = pbmc,
    reduction = "cca"
  )

  predicted.labels <- TransferData(
    anchorset = transfer.anchors,
    refdata = pbmc_rna$celltype,
    weight.reduction = pbmc[["lsi"]],
    dims = 2:30
  )

  pbmc <- AddMetaData(object = pbmc, metadata = predicted.labels)

  saveRDS(pbmc, ck_labeled)
  message("Saved checkpoint ", ck_labeled)
}
```

```{r}
plot1 <- DimPlot(
  object = pbmc_rna, group.by = "celltype",
  label = TRUE, repel = TRUE
) + NoLegend() + ggtitle("scRNA-seq")

plot2 <- DimPlot(
  object = pbmc, group.by = "predicted.id",
  label = TRUE, repel = TRUE
) + NoLegend() + ggtitle("scATAC-seq")

plot1 + plot2
```

#### How confident are the predicted labels?

```{r}
VlnPlot(pbmc, "prediction.score.max", group.by = "predicted.id")
```

Cells called as platelets are suspect, since platelets are anucleate and
should not have an ATAC signal. The block below pulls every
`prediction.score.*` column for these cells and orders them by mean score.

```{r}
metadata_attributes <- colnames(pbmc[[]])
prediction_score_attributes <- grep("^prediction.score.", metadata_attributes, value = TRUE)
prediction_score_attributes <- setdiff(prediction_score_attributes, "prediction.score.max")

predicted_platelets <- which(pbmc$predicted.id == "Platelet")
platelet_scores <- pbmc[[]][predicted_platelets, prediction_score_attributes]

ordered_columns <- names(sort(colMeans(platelet_scores, na.rm = TRUE), decreasing = TRUE))
ordered_platelet_scores_df <- platelet_scores[, ordered_columns]

print(ordered_platelet_scores_df)
```

```{r}
predicted_id_counts <- table(pbmc$predicted.id)

# Keep only predicted.id values with more than 20 cells.
major_predicted_ids <- names(predicted_id_counts[predicted_id_counts > 20])
pbmc <- pbmc[, pbmc$predicted.id %in% major_predicted_ids]

# Set cell identities to the per-cell predicted labels.
Idents(pbmc) <- pbmc$predicted.id

# Replace each cluster label with its most likely predicted label.
for (i in levels(pbmc)) {
  cells_to_reid <- WhichCells(pbmc, idents = i)
  newid <- names(which.max(table(pbmc$predicted.id[cells_to_reid])))
  Idents(pbmc, cells = cells_to_reid) <- newid
}
```

### Find differentially accessible peaks between cell types

`presto` greatly accelerates `FindMarkers`. It comes from GitHub
(<https://github.com/immunogenomics/presto>) and must be installed before the
workshop, never during a session. The setup chunk at the top of this document
reports whether it is present.

Both differential-accessibility tests below are checkpointed together into
`ck_markers`. They are the single most expensive part of this tutorial: the
Wilcoxon test scans roughly 100k peaks, and the logistic-regression test fits
one model per peak. In class these load from the `.rds`.

```{r}
DefaultAssay(pbmc) <- "peaks"

if (file.exists(ck_markers)) {
  marker_results <- readRDS(ck_markers)
  da_peaks <- marker_results$da_peaks
  lr_da_peaks <- marker_results$lr_da_peaks
  message("Loaded differential accessibility results from ", ck_markers)
} else {
  # wilcox is the default option for test.use. This is the step that needs
  # presto. Without it, expect hours rather than minutes.
  da_peaks <- FindMarkers(
    object = pbmc,
    ident.1 = "CD4 Naive",
    ident.2 = "CD14+ Monocytes",
    test.use = "wilcox",
    min.pct = 0.1
  )

  # Logistic regression with total fragment count as a latent variable. LR
  # fits a model per peak and gets no benefit from presto, so it is restricted
  # to the peaks the Wilcoxon test already flagged. Running it genome-wide
  # takes hours and changes none of the conclusions drawn below.
  lr_features <- head(rownames(da_peaks[order(da_peaks$p_val), ]), 500)
  lr_da_peaks <- FindMarkers(
    object = pbmc,
    ident.1 = "CD4 Naive",
    ident.2 = "CD14+ Monocytes",
    test.use = "LR",
    latent.vars = "nCount_peaks",
    features = lr_features,
    min.pct = 0.1
  )

  saveRDS(
    list(da_peaks = da_peaks, lr_da_peaks = lr_da_peaks),
    file.path(ck_write_dir, "pbmc_atac_ck4_markers.rds")
  )
  message("Saved checkpoint pbmc_atac_ck4_markers.rds")
}

head(da_peaks)
```

```{r}
head(lr_da_peaks)
```

```{r}
plot1 <- VlnPlot(
  object = pbmc,
  features = rownames(da_peaks)[1],
  pt.size = 0.1,
  idents = c("CD4 Naive", "CD14+ Monocytes")
)
plot2 <- FeaturePlot(
  object = pbmc,
  features = rownames(da_peaks)[1],
  pt.size = 0.1
)

plot1 | plot2
```

```{r}
open_cd4naive <- rownames(da_peaks[da_peaks$avg_log2FC > 3, ])
open_cd14mono <- rownames(da_peaks[da_peaks$avg_log2FC < -3, ])

closest_genes_cd4naive <- ClosestFeature(pbmc, regions = open_cd4naive)
closest_genes_cd14mono <- ClosestFeature(pbmc, regions = open_cd14mono)
```

```{r}
head(closest_genes_cd4naive)
```

```{r}
head(closest_genes_cd14mono)
```

#### GO enrichment of nearest genes

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
```

```{r}
cd4naive_ego <- enrichGO(
  gene = closest_genes_cd4naive$gene_id,
  keyType = "ENSEMBL",
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05,
  readable = TRUE
)

barplot(cd4naive_ego, showCategory = 20)
```

```{r}
cd14mono_ego <- enrichGO(
  gene = closest_genes_cd14mono$gene_id,
  keyType = "ENSEMBL",
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05,
  readable = TRUE
)

barplot(cd14mono_ego, showCategory = 20)
```

#### Plotting genomic regions

```{r}
pbmc <- SortIdents(pbmc)
```

```{r}
regions_highlight <- subsetByOverlaps(
  StringToGRanges(open_cd4naive),
  LookupGeneCoords(pbmc, "CD4")
)

CoveragePlot(
  object = pbmc,
  region = "CD4",
  region.highlight = regions_highlight,
  extend.upstream = 1000,
  extend.downstream = 1000
)
```

```{r}
regions_highlight <- subsetByOverlaps(
  StringToGRanges(open_cd14mono),
  LookupGeneCoords(pbmc, "LYZ")
)

CoveragePlot(
  object = pbmc,
  region = "LYZ",
  region.highlight = regions_highlight,
  extend.upstream = 1000,
  extend.downstream = 5000
)
```

```{r}
sessionInfo()
```
