---
title: "CITE-seq / multimodal vignette"
format: html
---

Adapted from <https://satijalab.org/seurat/articles/multimodal_vignette>.

```{r}
library(Seurat)
library(ggplot2)
library(patchwork)
```

The CITE-seq UMI matrices (RNA and ADT, GSE100866) are shipped with this
repository. If they are missing, the cell below will download them.

```{r}
#| eval: false
# Run once, change `eval: true` if you need to re-download.
curl::curl_download(
  "https://github.com/steverozen/cite_and_atac_wkshp/raw/refs/heads/main/GSE100866_CBMC_8K_13AB_10X-ADT_umi.csv.gz",
  destfile = here::here("GSE100866_CBMC_8K_13AB_10X-ADT_umi.csv.gz")
)
curl::curl_download(
  "https://github.com/steverozen/cite_and_atac_wkshp/raw/refs/heads/main/GSE100866_CBMC_8K_13AB_10X-RNA_umi.csv.gz",
  destfile = here::here("GSE100866_CBMC_8K_13AB_10X-RNA_umi.csv.gz")
)
```

This next cell may take a minute or two.

```{r}
# Load in the RNA UMI matrix.
#
# Note that this dataset also contains ~5% of mouse cells, which we can use as
# negative controls for the protein measurements. For this reason, the gene
# expression matrix has HUMAN_ or MOUSE_ appended to the beginning of each
# gene.
cbmc.rna <- as.sparse(read.csv(
  file = here::here("GSE100866_CBMC_8K_13AB_10X-RNA_umi.csv.gz"),
  sep = ",", header = TRUE, row.names = 1
))

# To make life a bit easier going forward, we discard all but the top 100 most
# highly expressed mouse genes, and remove the HUMAN_ prefix from the human
# genes.
cbmc.rna <- CollapseSpeciesExpressionMatrix(cbmc.rna) # Seurat function

# Load in the ADT UMI matrix (ADT = antibody derived tags).
cbmc.adt <- as.sparse(read.csv(
  file = here::here("GSE100866_CBMC_8K_13AB_10X-ADT_umi.csv.gz"),
  sep = ",", header = TRUE, row.names = 1
))

# Since measurements were made in the same cells, the two matrices have
# identical column names.
all.equal(colnames(cbmc.rna), colnames(cbmc.adt))
```

### Setup a Seurat object, add the RNA and protein data

```{r}
# Create a Seurat object based on the scRNA-seq data.
cbmc <- CreateSeuratObject(counts = cbmc.rna)

# By default, the cbmc object contains an assay storing RNA measurements.
Assays(cbmc)
```

```{r}
# Create a new assay to store ADT information.
adt_assay <- CreateAssay5Object(counts = cbmc.adt)

# Add this assay to the previously created Seurat object.
cbmc[["ADT"]] <- adt_assay

# Validate that the object now contains multiple assays.
Assays(cbmc)
```

```{r}
# Extract a list of features measured in the ADT assay.
rownames(cbmc[["ADT"]])
```

```{r}
# List the current default assay.
DefaultAssay(cbmc)
```

```{r}
# Switch the default to ADT.
DefaultAssay(cbmc) <- "ADT"
DefaultAssay(cbmc)
```

```{r}
# All operations below are performed on the RNA assay. Set and verify that the
# default assay is RNA.
DefaultAssay(cbmc) <- "RNA"
DefaultAssay(cbmc)
```

## Important: for production, use sctransform for the first 3 steps below

```{r}
# Perform visualization and clustering steps.
cbmc <- NormalizeData(cbmc)
cbmc <- FindVariableFeatures(cbmc)
cbmc <- ScaleData(cbmc)
cbmc <- RunPCA(cbmc, verbose = FALSE)
cbmc <- FindNeighbors(cbmc, dims = 1:30) ### Go to slides explaining cluster
cbmc <- FindClusters(cbmc, resolution = 0.8, verbose = FALSE)
cbmc <- RunUMAP(cbmc, dims = 1:30)
DimPlot(cbmc, label = TRUE)
```

```{r}
# Normalize ADT data.
DefaultAssay(cbmc) <- "ADT"
cbmc <- NormalizeData(cbmc, normalization.method = "CLR", margin = 2)
DefaultAssay(cbmc) <- "RNA"

# The following command is an alternative but returns the same result.
cbmc <- NormalizeData(cbmc, normalization.method = "CLR", margin = 2, assay = "ADT")

# Visualize CD19 levels for RNA and protein. By setting the default assay we
# can visualize one or the other.
DefaultAssay(cbmc) <- "ADT"
p1 <- FeaturePlot(cbmc, "CD19", cols = c("lightgrey", "darkgreen")) + ggtitle("CD19 protein")
DefaultAssay(cbmc) <- "RNA"
p2 <- FeaturePlot(cbmc, "CD19") + ggtitle("CD19 RNA")

# Place plots side-by-side.
p1 | p2
```

```{r}
# Alternately, we can use specific assay keys to specify a specific modality.
# Here we ask what the RNA key is.
Key(cbmc[["RNA"]])
```

```{r}
# Here we ask what the ADT key is.
Key(cbmc[["ADT"]])
```

```{r}
# Including the key in the feature name overrides the default assay, e.g.
# prepending "adt_" to "CD19".
p1 <- FeaturePlot(cbmc, "adt_CD19", cols = c("lightgrey", "darkgreen")) + ggtitle("CD19 protein")
p2 <- FeaturePlot(cbmc, "rna_CD19") + ggtitle("CD19 RNA")
p1 | p2
```

```{r}
# As CD19 is a B cell marker, we can identify cluster 6 as expressing CD19 on
# the surface.
VlnPlot(cbmc, "adt_CD19")
```

```{r}
# We can also identify alternative protein and RNA markers for this cluster
# through differential expression.
adt_markers <- FindMarkers(cbmc, ident.1 = 6, assay = "ADT")
rna_markers <- FindMarkers(cbmc, ident.1 = 6, assay = "RNA")

head(adt_markers)
```

```{r}
head(rna_markers)
```

### Additional visualizations of multimodal data

```{r}
# Draw ADT scatter plots (like biaxial plots for FACS). Note that you can even
# 'gate' cells if desired by using HoverLocator and FeatureLocator.
FeatureScatter(cbmc, feature1 = "adt_CD19", feature2 = "adt_CD3")
```

```{r}
# View relationship between protein and RNA.
FeatureScatter(cbmc, feature1 = "adt_CD3", feature2 = "rna_CD3E")
```

```{r}
FeatureScatter(cbmc, feature1 = "adt_CD4", feature2 = "adt_CD8")
```

```{r}
# Raw (non-normalized) ADT counts. The values are quite high, particularly
# compared to RNA. This is due to the significantly higher protein copy
# number in cells, which reduces 'drop-out' in ADT data.
FeatureScatter(cbmc, feature1 = "adt_CD4", feature2 = "adt_CD8", slot = "counts")
```

### Loading data from 10X multi-modal experiments

```{r}
#| eval: false
# Download once, ~49 MB. Change to eval: true if you need to download.
download.file(
  "https://cf.10xgenomics.com/samples/cell-exp/3.0.0/pbmc_10k_protein_v3/pbmc_10k_protein_v3_filtered_feature_bc_matrix.tar.gz",
  destfile = here::here("pbmc_10k_protein_v3_filtered_feature_bc_matrix.tar.gz")
)
```

```{r}
if (!dir.exists(here::here("filtered_feature_bc_matrix"))) {
  untar(
    here::here("pbmc_10k_protein_v3_filtered_feature_bc_matrix.tar.gz"),
    exdir = here::here()
  )
}
```

```{r}
# Read10X returns a named list with "Gene Expression" and "Antibody Capture".
pbmc10k.data <- Read10X(data.dir = here::here("filtered_feature_bc_matrix"))

rownames(x = pbmc10k.data[["Antibody Capture"]]) <- gsub(
  pattern = "_[control_]*TotalSeqB", replacement = "",
  x = rownames(x = pbmc10k.data[["Antibody Capture"]])
)

pbmc10k <- CreateSeuratObject(
  counts = pbmc10k.data[["Gene Expression"]], min.cells = 3, min.features = 200
)
pbmc10k <- NormalizeData(pbmc10k)
pbmc10k[["ADT"]] <- CreateAssayObject(
  pbmc10k.data[["Antibody Capture"]][, colnames(x = pbmc10k)]
)
pbmc10k <- NormalizeData(pbmc10k, assay = "ADT", normalization.method = "CLR")

plot1 <- FeatureScatter(pbmc10k, feature1 = "adt_CD19", feature2 = "adt_CD3", pt.size = 1)
plot2 <- FeatureScatter(pbmc10k, feature1 = "adt_CD4", feature2 = "adt_CD8a", pt.size = 1)
plot3 <- FeatureScatter(pbmc10k, feature1 = "adt_CD3", feature2 = "CD3E", pt.size = 1)
(plot1 + plot2 + plot3) & NoLegend()
```

```{r}
sessionInfo()
```
