---
title: "Seurat PBMC3K Guided Clustering Tutorial"
subtitle: "Source: `pbmc3k_tutorial.qmd`"
author: "Steve Rozen (sr110@duke.edu, steverozen@pm.me)"
date: last-modified
date-format: "YYYY-MM-DD HH:mm"
format:
  html:
    page-layout: full
    include-in-header:
      text: |
        <style>
          .quarto-container, .page-columns, .main-container {
            max-width: 100% !important;
          }
        </style>
---

## Introduction

This document follows the [Seurat PBMC3K guided clustering tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial),
which walks through a standard single-cell RNA-seq analysis workflow on a dataset
of Peripheral Blood Mononuclear Cells (PBMCs) from 10X Genomics.
The dataset contains `r 2700` cells sequenced on the Illumina NextSeq 500.

## Setup

```{r}
#| label: setup
#| message: false
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
```

## Load Data

Read the 10X Genomics Cell Ranger output and create a Seurat object.
Genes detected in fewer than 3 cells and cells with fewer than 200 detected
genes are excluded.

```{r}
#| label: load-data
#| warning: false
# Suppress warning about underscores in feature names (10X format uses them)
pbmc.data <- Read10X(data.dir = "filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(
  counts  = pbmc.data,
  project = "pbmc3k",
  min.cells    = 3,
  min.features = 200
)
pbmc
```

### Examine the count matrix

```{r}
#| label: examine-counts
#| message: false
pbmc.data[c("CD3D", "TCL1A", "MS4A1"), 1:30]
```

The `.` values in the matrix represent zeros (no molecules detected). Since
most values in scRNA-seq are 0, Seurat uses a sparse-matrix representation
to save memory:

```{r}
#| label: sparse-vs-dense
dense.size  <- object.size(as.matrix(pbmc.data))
sparse.size <- object.size(pbmc.data)
dense.size / sparse.size
```

## Quality Control

Common QC metrics include:

- Number of unique genes per cell (`nFeature_RNA`)
- Total molecules per cell (`nCount_RNA`)
- Percentage of reads mapping to mitochondrial genes (`percent.mt`) --
  high values indicate dying cells

```{r}
#| label: qc-metrics
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
head(pbmc@meta.data, 5)
```

### QC Visualization

```{r}
#| label: qc-violin
#| fig-width: 10
#| fig-height: 5
#| warning: false
# Warning about "data" layer not found is expected before normalization
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
#| label: qc-scatter
#| fig-width: 10
#| fig-height: 5
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

### Filter cells

Retain cells with 200--2,500 unique genes and < 5% mitochondrial content.

```{r}
#| label: filter-cells
pbmc <- subset(
  pbmc,
  subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5
)
pbmc
```

## Normalization

Log-normalize: divide each cell's gene counts by total expression, multiply by
10,000, then log-transform.

```{r}
#| label: normalize
#| message: false
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
```

## Identify Highly Variable Features

Select the top 2,000 genes with the highest cell-to-cell variation using
variance-stabilizing transformation (VST).

```{r}
#| label: variable-features
#| message: false
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)
top10 <- head(VariableFeatures(pbmc), 10)
top10
```

```{r}
#| label: variable-features-plot
#| fig-width: 10
#| fig-height: 5
#| message: false
#| warning: false
# Suppress repel nudge message and log-10 infinite values warning (expected for genes with zero variance)
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

## Scale Data

Shift each gene's expression so that mean = 0 across cells and scale so that
variance = 1, preventing highly-expressed genes from dominating downstream
analyses.

```{r}
#| label: scale-data
#| message: false
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
```

## Linear Dimensional Reduction (PCA)

```{r}
#| label: run-pca
#| message: false
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
```

```{r}
#| label: pca-loadings
print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
```

```{r}
#| label: pca-viz-loadings
#| fig-width: 10
#| fig-height: 5
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
```

```{r}
#| label: pca-dimplot
#| fig-width: 7
#| fig-height: 5
DimPlot(pbmc, reduction = "pca") + NoLegend()
```

### PCA Heatmaps

```{r}
#| label: pca-heatmap-1
#| fig-width: 8
#| fig-height: 6
DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)
```

```{r}
#| label: pca-heatmap-15
#| fig-width: 10
#| fig-height: 18
DimHeatmap(pbmc, dims = 1:15, cells = 500, balanced = TRUE)
```

## Determine Dimensionality

```{r}
#| label: elbow-plot
#| fig-width: 7
#| fig-height: 5
ElbowPlot(pbmc)
```

The elbow appears around PC 9--10. We use 10 PCs for downstream analysis.

## Cluster Cells

Build a K-nearest neighbor graph in PCA space, then apply the Louvain
algorithm with resolution 0.5.

```{r}
#| label: find-clusters
#| message: false
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)
```

```{r}
#| label: cluster-ids
head(Idents(pbmc), 5)
```

## Non-linear Dimensional Reduction (UMAP)

```{r}
#| label: run-umap
#| message: false
#| warning: false
# Suppress UMAP method change message (using R-native UWOT, which is correct)
pbmc <- RunUMAP(pbmc, dims = 1:10)
```

```{r}
#| label: umap-plot
#| fig-width: 8
#| fig-height: 6
DimPlot(pbmc, reduction = "umap")
```

## Save Intermediate Object

```{r}
#| label: save-intermediate

filepath <- "output/pbmc_tutorial.rds"
dir.create(dirname(filepath), recursive = TRUE, showWarnings = FALSE)
saveRDS(pbmc, file = filepath)
```

## Find Differentially Expressed Markers

### Cluster 2 markers

```{r}
#| label: cluster2-markers
cluster2.markers <- FindMarkers(pbmc, ident.1 = 2)
head(cluster2.markers, n = 5)
```

### Cluster 5 vs. clusters 0 and 3

```{r}
#| label: cluster5-markers
cluster5.markers <- FindMarkers(pbmc, ident.1 = 5, ident.2 = c(0, 3))
head(cluster5.markers, n = 5)
```

### All cluster markers

```{r}
#| label: all-markers
#| message: false
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE)
pbmc.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1)
```

### ROC test for cluster 0

```{r}
#| label: roc-markers
cluster0.markers <- FindMarkers(
  pbmc,
  ident.1        = 0,
  logfc.threshold = 0.25,
  test.use       = "roc",
  only.pos       = TRUE
)
head(cluster0.markers)
```

## Visualize Markers

### Violin plots

```{r}
#| label: marker-violin
#| fig-width: 10
#| fig-height: 5
VlnPlot(pbmc, features = c("MS4A1", "CD79A"))
```

```{r}
#| label: marker-violin-raw
#| fig-width: 10
#| fig-height: 5
# Suppress "default search for data layer" warning (expected when requesting counts layer)
suppressWarnings(
  print(VlnPlot(pbmc, features = c("NKG7", "PF4"), layer = "counts", log = TRUE))
)
```

### Feature plots on UMAP

```{r}
#| label: feature-plot
#| fig-width: 10
#| fig-height: 10
FeaturePlot(
  pbmc,
  features = c("MS4A1", "GNLY", "CD3E", "CD14",
               "FCER1A", "FCGR3A", "LYZ", "PPBP", "CD8A")
)
```

### Marker heatmap

```{r}
#| label: marker-heatmap
#| fig-width: 10
#| fig-height: 10
pbmc.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 10) %>%
  ungroup() -> top10_markers

DoHeatmap(pbmc, features = top10_markers$gene) + NoLegend()
```

## Assign Cell Type Identity

Map clusters to cell types based on canonical markers.

```{r}
#| label: assign-cell-types
new.cluster.ids <- c(
  "Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B",
  "CD8 T", "FCGR3A+ Mono", "NK", "DC", "Platelet"
)
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
```

```{r}
#| label: umap-labeled
#| fig-width: 10
#| fig-height: 7
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```

## Publication-Quality Figure

```{r}
#| label: pub-umap
#| fig-width: 12
#| fig-height: 7
base_size <- 14

plot <- DimPlot(pbmc, reduction = "umap", label = TRUE,
                label.size = 0.7 * base_size / ggplot2::.pt) +
  xlab("UMAP 1") + ylab("UMAP 2") +
  theme(
    axis.title  = element_text(size = rel(1.3)),
    legend.text = element_text(size = rel(1.3))
  ) +
  guides(colour = guide_legend(override.aes = list(size = 10)))
plot

ggsave(
  filename = "output/images/pbmc3k_umap.jpg",
  height = 7, width = 12,
  plot = plot, quality = 50
)
```

## Save Final Object

```{r}
#| label: save-final
saveRDS(pbmc, file = "output/pbmc3k_final.rds")
```

## References

- Hao Y et al. (2024) "Dictionary learning for integrative, multimodal and scalable single-cell analysis." *Nat Biotechnol* 42:293--304. <https://doi.org/10.1038/s41587-023-01767-y>
- McInnes L, Healy J, Melville J (2018) "UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction." *arXiv:1802.03426*. <https://doi.org/10.48550/arXiv.1802.03426>

::: {.callout-note collapse="true"}
## Session Info

```{r}
#| label: session-info
#| echo: false
#| comment: ""
sessioninfo::session_info(info = "all")
```

:::
