---
title: "Seurat SCTransform Normalization Tutorial"
subtitle: "Source: `sctransform_vignette.qmd`"
author: "Steve Rozen (sr110@duke.edu, steverozen@pm.me)"
date: last-modified
date-format: "YYYY-MM-DD HH:mm"
format:
  html:
    page-layout: full
    include-in-header:
      text: |
        <style>
          .quarto-container, .page-columns, .main-container {
            max-width: 100% !important;
          }
        </style>
---

## Introduction

This document follows the [Seurat SCTransform vignette](https://satijalab.org/seurat/articles/sctransform_vignette),
which demonstrates an improved normalization workflow using
[sctransform](https://github.com/satijalab/sctransform).
SCTransform replaces the three separate steps of `NormalizeData()`,
`ScaleData()`, and `FindVariableFeatures()` with a single regularized
negative binomial regression that removes technical variation while preserving
biological heterogeneity. It uses the same PBMC 3K dataset from 10X Genomics.

## Setup

```{r}
#| label: setup
#| message: false
library(Seurat)
library(ggplot2)
library(sctransform)
```

## Load Data

```{r}
#| label: load-data
#| warning: false
# Suppress warning about underscores in feature names (10X format uses them)
pbmc_data <- Read10X(data.dir = "filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(counts = pbmc_data)
pbmc
```

## Quality Control

Calculate the percentage of mitochondrial reads per cell.

```{r}
#| label: qc-mito
pbmc <- PercentageFeatureSet(pbmc, pattern = "^MT-", col.name = "percent.mt")
```

## SCTransform Normalization

`SCTransform()` replaces `NormalizeData()`, `ScaleData()`, and
`FindVariableFeatures()` in a single command. Here we also regress out
mitochondrial percentage to reduce its influence on clustering.

```{r}
#| label: sctransform
#| message: false
#| warning: false
pbmc <- SCTransform(pbmc, vars.to.regress = "percent.mt", verbose = FALSE)
```

## Dimensionality Reduction

SCTransform's improved variance stabilization allows using more PCs than the
standard log-normalize workflow. Here we use 30 PCs.

### PCA

```{r}
#| label: run-pca
#| message: false
pbmc <- RunPCA(pbmc, verbose = FALSE)
```

### UMAP

```{r}
#| label: run-umap
#| message: false
#| warning: false
# Suppress UMAP method change message (using R-native UWOT, which is correct)
pbmc <- RunUMAP(pbmc, dims = 1:30, verbose = FALSE)
```

## Clustering

```{r}
#| label: clustering
#| message: false
pbmc <- FindNeighbors(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindClusters(pbmc, verbose = FALSE)
```

## UMAP Visualization

```{r}
#| label: umap-clusters
#| fig-width: 8
#| fig-height: 6
DimPlot(pbmc, label = TRUE)
```

## Marker Gene Visualization

### Violin plots

```{r}
#| label: marker-violin
#| fig-width: 10
#| fig-height: 8
#| warning: false
# Suppress "data layer" warning (expected with SCT assay)
VlnPlot(
  pbmc,
  features = c("CD8A", "GZMK", "CCL5", "S100A4",
               "ANXA1", "CCR7", "ISG15", "CD3D"),
  pt.size = 0.2, ncol = 2
)
```

### Feature plots -- T cell and CD4 markers

```{r}
#| label: feature-plot-1
#| fig-width: 10
#| fig-height: 8
FeaturePlot(
  pbmc,
  features = c("CD8A", "GZMK", "CCL5", "S100A4", "ANXA1", "CCR7"),
  pt.size = 0.2, ncol = 2
)
```

### Feature plots -- T cell, B cell, and NK markers

```{r}
#| label: feature-plot-2
#| fig-width: 10
#| fig-height: 8
FeaturePlot(
  pbmc,
  features = c("CD3D", "ISG15", "TCL1A", "FCER2", "XCL1", "FCGR3A"),
  pt.size = 0.2, ncol = 2
)
```

## Save Object

```{r}
#| label: save-object
saveRDS(pbmc, file = "output/pbmc_sctransform.rds")
```

## References

- Hao Y et al. (2024) "Dictionary learning for integrative, multimodal and scalable single-cell analysis." *Nat Biotechnol* 42:293--304. <https://doi.org/10.1038/s41587-023-01767-y>
- Hafemeister C, Satija R (2019) "Normalization and variance stabilization of single-cell RNA-seq data using regularized negative binomial regression." *Genome Biol* 20:296. <https://doi.org/10.1186/s13059-019-1874-1>
- Choudhary S, Satija R (2022) "Comparison and evaluation of statistical error models for scRNA-seq." *Genome Biol* 23:27. <https://doi.org/10.1186/s13059-021-02584-9>
- McInnes L, Healy J, Melville J (2018) "UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction." *arXiv:1802.03426*. <https://doi.org/10.48550/arXiv.1802.03426>

::: {.callout-note collapse="true"}
## Session Info

```{r}
#| label: session-info
#| echo: false
#| comment: ""
sessioninfo::session_info(info = "all")
```

:::
