---
title: "tidyplots vs. ggplot2: the Get-started tutorial, side by side"
subtitle: "Every example from the tidyplots Get-started guide, with a matching ggplot2 version"
author: "Bioinformatics & Research Computing"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 4
    fig-height: 3
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
---

## About this handout

This document walks through Jan Broder Engler's
[tidyplots *Get started* tutorial](https://jbengler.github.io/tidyplots/articles/tidyplots.html)
example by example. For each step we show the original **tidyplots** code and its
plot, then a **ggplot2** version that produces the same figure. The point is to
see how the two grammars express the same idea.

The mental difference in one line:

- **tidyplots** pipes a data frame through `add_*()`, `remove_*()`, and
  `adjust_*()` verbs. The statistic and its visual form are named together
  (`add_mean_bar()`, `add_sem_errorbar()`).
- **ggplot2** maps data columns to aesthetics inside `aes()`, then stacks
  `geom_*()` and `stat_summary()` layers with `+`. Statistics and geoms are
  separate dials.

Both sit on the grammar of graphics, so the translation is usually mechanical.
The interesting cases are where tidyplots bundles several ggplot2 moves into one
verb (the SEM error bar, the fixed plot-area size, the built-in theme).

## Setup

```{r}
#| label: setup
library(tidyplots)   # the package the tutorial teaches
library(ggplot2)     # the package we translate each example into
library(ggbeeswarm)  # geom_beeswarm(), needed to match add_data_points_beeswarm()
```

A note on `stat_summary()`, which does most of the work on the ggplot2 side:

- `fun = mean` computes a single y value per x group. Pair it with
  `geom = "col"` for a bar or `geom = "crossbar"` (with `fun.min`/`fun.max` also
  set to `mean`) for a flat dash at the mean.
- `fun.data = mean_se` returns the mean together with `mean ± 1 standard error`,
  which is exactly what tidyplots' `add_sem_errorbar()` draws.

## Data preparation

tidyplots ships demo datasets. The tutorial uses `study`: 20 participants, four
treatment groups (A, B, C, D), and a measured `score`.

```{r}
#| label: data
study
```

## Plotting

### The empty plot

We start the same way in both: declare the data and which columns map to x and y.
In tidyplots the mapping goes inside `tidyplot()`. In ggplot2 it goes inside
`aes()`. Neither has drawn a geom yet, so both panels are empty.

#### tidyplots

```{r}
#| label: empty-tp
study |>
  tidyplot(x = treatment, y = score)
```

#### ggplot2

```{r}
#| label: empty-gg
ggplot(study, aes(x = treatment, y = score))
```

### Add data points

tidyplots adds a layer with `add_data_points()`. ggplot2 adds a layer with
`geom_point()`. (tidyplots nudges overlapping points apart a little, plain
`geom_point()` stacks them in a straight column. We match the beeswarm spread
later.)

#### tidyplots

```{r}
#| label: points-tp
study |>
  tidyplot(x = treatment, y = score) |>
  add_data_points()
```

#### ggplot2

```{r}
#| label: points-gg
ggplot(study, aes(x = treatment, y = score)) +
  geom_point()
```

### Add a mean bar

`add_mean_bar(alpha = 0.4)` computes the group mean and draws it as a
semi-transparent bar. In ggplot2 this is `stat_summary(fun = mean, geom = "col")`:
`fun = mean` is the statistic, `geom = "col"` is how to draw it.

#### tidyplots

```{r}
#| label: meanbar-tp
study |>
  tidyplot(x = treatment, y = score) |>
  add_data_points() |>
  add_mean_bar(alpha = 0.4)
```

#### ggplot2

```{r}
#| label: meanbar-gg
ggplot(study, aes(x = treatment, y = score)) +
  geom_point() +
  stat_summary(fun = mean, geom = "col", alpha = 0.4)
```

### Swap the bar for a dash, add SEM error bars

`add_mean_dash()` draws the mean as a short horizontal dash instead of a bar, and
`add_sem_errorbar()` adds standard-error-of-the-mean error bars. In ggplot2 the
dash is a `crossbar` with its top and bottom both pinned to the mean, and the
error bar uses `fun.data = mean_se`.

#### tidyplots

```{r}
#| label: dash-tp
study |>
  tidyplot(x = treatment, y = score) |>
  add_data_points() |>
  add_mean_dash() |>
  add_sem_errorbar()
```

#### ggplot2

```{r}
#| label: dash-gg
ggplot(study, aes(x = treatment, y = score)) +
  geom_point() +
  # fun.min and fun.max both = mean -> the crossbar collapses to a flat dash
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", width = 0.5) +
  # mean_se returns mean and mean +/- 1 standard error
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4)
```

### Color as a third variable

Color encodes a *variable*, so in tidyplots it goes in the initial `tidyplot()`
call. Here `x = group`, `color = dose`. In ggplot2, `color = dose` goes in
`aes()`. tidyplots automatically separates (dodges) the colored groups side by
side, so on the ggplot2 side we add `position_dodge()` to every layer to match.

#### tidyplots

```{r}
#| label: color-tp
study |>
  tidyplot(x = group, y = score, color = dose) |>
  add_data_points() |>
  add_mean_dash() |>
  add_sem_errorbar()
```

#### ggplot2

```{r}
#| label: color-gg
dodge <- position_dodge(width = 0.8)   # shared dodge so layers line up
ggplot(study, aes(x = group, y = score, color = dose)) +
  geom_point(position = dodge) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", width = 0.5, position = dodge) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4,
               position = dodge)
```

## Remove

`remove_*()` strips elements. Removing the legend title and the whole y-axis in
tidyplots is two verbs. In ggplot2 you set the legend title to `NULL` in `labs()`
and blank the y-axis pieces inside `theme()`.

#### tidyplots

```{r}
#| label: remove-tp
study |>
  tidyplot(x = group, y = score, color = dose) |>
  add_data_points() |>
  add_mean_dash() |>
  add_sem_errorbar() |>
  remove_legend_title() |>
  remove_y_axis()
```

#### ggplot2

```{r}
#| label: remove-gg
dodge <- position_dodge(width = 0.8)
ggplot(study, aes(x = group, y = score, color = dose)) +
  geom_point(position = dodge) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", width = 0.5, position = dodge) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4,
               position = dodge) +
  labs(color = NULL) +                       # drop the legend title
  theme(axis.title.y = element_blank(),      # drop the y-axis title
        axis.text.y  = element_blank(),      # drop the y tick labels
        axis.ticks.y = element_blank(),      # drop the y tick marks
        axis.line.y  = element_blank())      # drop the y axis line
```

## Adjust

The `adjust_*()` family tweaks display details. This is the base plot we keep
adjusting, points plus a mean bar plus SEM error bars, colored by treatment.

#### tidyplots

```{r}
#| label: base-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_mean_bar(alpha = 0.4) |>
  add_sem_errorbar()
```

#### ggplot2

When the bar is colored, tidyplots fills it with the group color at low alpha.
In ggplot2, point outline color comes from `color` and bar interior from `fill`,
so we map both. `color = NA` removes the bar outline.

```{r}
#| label: base-gg
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_point() +
  stat_summary(aes(fill = treatment), fun = mean, geom = "col",
               alpha = 0.4, color = NA) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4)
```

### Fixed plot-area size and beeswarm points

`add_data_points_beeswarm(shape = 1)` spreads overlapping points into a swarm
with hollow circles, and `adjust_size(width = 20, height = 20)` fixes the
*plot-area* size to 20 mm by 20 mm. ggbeeswarm gives us the swarm. ggplot2 has no
built-in "set the plot area to N millimetres" control, you size the whole device
through the chunk's `fig-width`/`fig-height` (or `ggsave()`), so the match here is
approximate.

#### tidyplots

```{r}
#| label: beeswarm-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points_beeswarm(shape = 1) |>
  add_mean_bar(alpha = 0.4) |>
  add_sem_errorbar() |>
  adjust_size(width = 20, height = 20)
```

#### ggplot2

```{r}
#| label: beeswarm-gg
#| fig-width: 3
#| fig-height: 3
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_beeswarm(shape = 1) +                 # hollow circles, spread into a swarm
  stat_summary(aes(fill = treatment), fun = mean, geom = "col",
               alpha = 0.4, color = NA) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4)
```

### Titles and captions

`adjust_title()` and friends set the various labels. In ggplot2 every label lives
in a single `labs()` call. Setting the legend title to `""` mirrors
`adjust_legend_title("")`.

#### tidyplots

```{r}
#| label: titles-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_mean_bar(alpha = 0.4) |>
  add_sem_errorbar() |>
  adjust_title("This is my fantastic plot title") |>
  adjust_x_axis_title("Treatment group") |>
  adjust_y_axis_title("Disease score") |>
  adjust_legend_title("") |>
  adjust_caption("Here goes the caption")
```

#### ggplot2

```{r}
#| label: titles-gg
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_point() +
  stat_summary(aes(fill = treatment), fun = mean, geom = "col",
               alpha = 0.4, color = NA) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  labs(title   = "This is my fantastic plot title",
       x       = "Treatment group",
       y       = "Disease score",
       color   = "",                    # empty legend title
       fill    = "",
       caption = "Here goes the caption")
```

### Custom colors

`adjust_colors()` takes a vector of colors. The ggplot2 equivalents are
`scale_color_manual()` for the points and `scale_fill_manual()` for the bars,
both fed the same vector.

#### tidyplots

```{r}
#| label: colors-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_mean_bar(alpha = 0.4) |>
  add_sem_errorbar() |>
  adjust_colors(new_colors = c("#644296", "#F08533", "#3B78B0", "#D1352C"))
```

#### ggplot2

```{r}
#| label: colors-gg
my_colors <- c("#644296", "#F08533", "#3B78B0", "#D1352C")
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_point() +
  stat_summary(aes(fill = treatment), fun = mean, geom = "col",
               alpha = 0.4, color = NA) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  scale_color_manual(values = my_colors) +
  scale_fill_manual(values = my_colors)
```

### A built-in color scheme

tidyplots ships color schemes such as `colors_discrete_seaside`. It is just a
character vector of hex codes, so on the ggplot2 side we can hand the very same
object to `scale_color_manual()` / `scale_fill_manual()`.

#### tidyplots

```{r}
#| label: scheme-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_mean_bar(alpha = 0.4) |>
  add_sem_errorbar() |>
  adjust_colors(new_colors = colors_discrete_seaside)
```

#### ggplot2

```{r}
#| label: scheme-gg
seaside <- as.character(colors_discrete_seaside)   # strip the tidyplots class
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_point() +
  stat_summary(aes(fill = treatment), fun = mean, geom = "col",
               alpha = 0.4, color = NA) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  scale_color_manual(values = seaside) +
  scale_fill_manual(values = seaside)
```

### Rename axis labels

`rename_x_axis_labels()` takes a named vector mapping old label to new label.
The ggplot2 equivalent is the `labels` argument of `scale_x_discrete()`.

#### tidyplots

```{r}
#| label: rename-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_mean_bar(alpha = 0.4) |>
  add_sem_errorbar() |>
  rename_x_axis_labels(new_names = c("A" = "This",
                                     "B" = "is",
                                     "C" = "totally",
                                     "D" = "new"))
```

#### ggplot2

```{r}
#| label: rename-gg
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_point() +
  stat_summary(aes(fill = treatment), fun = mean, geom = "col",
               alpha = 0.4, color = NA) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  scale_x_discrete(labels = c("A" = "This", "B" = "is",
                              "C" = "totally", "D" = "new"))
```

### Reorder axis labels

`reorder_x_axis_labels("D", "C")` brings D and C to the front. In ggplot2 you set
the category order with the `limits` argument of `scale_x_discrete()`.

#### tidyplots

```{r}
#| label: reorder-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_mean_bar(alpha = 0.4) |>
  add_sem_errorbar() |>
  reorder_x_axis_labels("D", "C")
```

#### ggplot2

```{r}
#| label: reorder-gg
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_point() +
  stat_summary(aes(fill = treatment), fun = mean, geom = "col",
               alpha = 0.4, color = NA) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  scale_x_discrete(limits = c("D", "C", "A", "B"))
```

### Sort axis labels by value

`sort_x_axis_labels()` orders the categories by their y value (the mean score
here). In ggplot2, `reorder(treatment, score, FUN = mean)` reorders the factor by
its group mean.

#### tidyplots

```{r}
#| label: sort-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_mean_bar(alpha = 0.4) |>
  add_sem_errorbar() |>
  sort_x_axis_labels()
```

#### ggplot2

```{r}
#| label: sort-gg
ggplot(study, aes(x = reorder(treatment, score, FUN = mean),
                  y = score, color = treatment)) +
  geom_point() +
  stat_summary(aes(fill = treatment), fun = mean, geom = "col",
               alpha = 0.4, color = NA) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  labs(x = "treatment")               # reorder() renames the axis, so restore it
```

### Reverse axis labels

`reverse_x_axis_labels()` flips the order. In ggplot2, `scale_x_discrete(limits = rev)`
reverses the existing category order.

#### tidyplots

```{r}
#| label: reverse-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_mean_bar(alpha = 0.4) |>
  add_sem_errorbar() |>
  reverse_x_axis_labels()
```

#### ggplot2

```{r}
#| label: reverse-gg
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_point() +
  stat_summary(aes(fill = treatment), fun = mean, geom = "col",
               alpha = 0.4, color = NA) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  scale_x_discrete(limits = rev)
```

## Themes

Themes change the look without changing the data. tidyplots and ggplot2 both
expose a family of `theme_*()` functions. The tidyplots default is
`theme_tidyplot()`, the closest stock ggplot2 look is `theme_classic()`.

#### tidyplots: `theme_tidyplot()`

```{r}
#| label: theme1-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_sem_errorbar() |>
  add_mean_dash() |>
  theme_tidyplot()
```

#### ggplot2: `theme_classic()`

```{r}
#| label: theme1-gg
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_point() +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", width = 0.5) +
  theme_classic()
```

#### tidyplots: `theme_ggplot2()`

tidyplots even bundles a theme that imitates ggplot2's grey default. The genuine
article is ggplot2's own `theme_grey()` (the default).

```{r}
#| label: theme2-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_sem_errorbar() |>
  add_mean_dash() |>
  theme_ggplot2()
```

#### ggplot2: `theme_grey()`

```{r}
#| label: theme2-gg
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_point() +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", width = 0.5) +
  theme_grey()
```

#### tidyplots: minimal with only a y grid

`theme_minimal_y()` keeps horizontal grid lines only, and `remove_x_axis_line()`
drops the x axis line. In ggplot2, `theme_minimal()` plus blanking the vertical
grid lines gets the same effect.

```{r}
#| label: theme3-tp
study |>
  tidyplot(x = treatment, y = score, color = treatment) |>
  add_data_points() |>
  add_sem_errorbar() |>
  add_mean_dash() |>
  theme_minimal_y() |>
  remove_x_axis_line()
```

#### ggplot2: `theme_minimal()` with vertical grid removed

```{r}
#| label: theme3-gg
ggplot(study, aes(x = treatment, y = score, color = treatment)) +
  geom_point() +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", width = 0.5) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),   # drop vertical grid lines
        panel.grid.minor.x = element_blank())
```

## Split into subplots

`split_plot(by = dose)` breaks one plot into a small multiple per `dose`. This is
exactly ggplot2's `facet_wrap()`.

#### tidyplots

```{r}
#| label: split-tp
#| fig-width: 6
#| fig-height: 3
study |>
  tidyplot(x = group, y = score, color = group) |>
  add_data_points() |>
  add_sem_errorbar() |>
  add_mean_dash() |>
  adjust_size(width = 30, height = 25) |>
  split_plot(by = dose)
```

#### ggplot2

```{r}
#| label: split-gg
#| fig-width: 6
#| fig-height: 3
ggplot(study, aes(x = group, y = score, color = group)) +
  geom_point() +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", width = 0.5) +
  facet_wrap(~ dose)
```

## Output

Both packages save to PDF or PNG at the end of a pipeline. tidyplots uses
`save_plot()`, ggplot2 uses `ggsave()`. The chunks below are not run, they just
show the syntax.

#### tidyplots

```{r}
#| label: save-tp
#| eval: false
study |>
  tidyplot(x = group, y = score, color = group) |>
  add_data_points() |>
  add_sem_errorbar() |>
  add_mean_dash() |>
  save_plot("my_plot.pdf")
```

#### ggplot2

`ggsave()` saves the last plot, or a plot object you pass to it.

```{r}
#| label: save-gg
#| eval: false
p <- ggplot(study, aes(x = group, y = score, color = group)) +
  geom_point() +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.4) +
  stat_summary(fun = mean, fun.min = mean, fun.max = mean,
               geom = "crossbar", width = 0.5)
ggsave("my_plot.pdf", plot = p)
```

## Takeaways

- The translation is mostly one verb to one layer. `add_data_points()` is
  `geom_point()`, `split_plot()` is `facet_wrap()`, `adjust_colors()` is
  `scale_*_manual()`.
- tidyplots packs the mean-and-its-error-bar idiom into named verbs
  (`add_mean_bar()` + `add_sem_errorbar()`). In ggplot2 you reach for
  `stat_summary()` and choose the statistic (`fun`, `fun.data`) and the geom
  separately.
- A few tidyplots conveniences have no one-line ggplot2 twin: the fixed
  plot-area size in millimetres (`adjust_size()`), and the publication-ready
  default theme. Those are exactly the ergonomics tidyplots is selling.
- Original tutorial:
  <https://jbengler.github.io/tidyplots/articles/tidyplots.html>
