---
title: "Spatial transcriptomics with Seurat: imaging-based platforms (MERSCOPE, Xenium, CosMx)"
subtitle: "Local version, reads from the workshop data repo, no network access"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: true
  error: true
---

Local rework of `seurat5_spatial_vignette_2.html`, the Seurat vignette
"Analysis of Image-based Spatial Data in Seurat".

Upstream source: <https://satijalab.org/seurat/articles/seurat5_spatial_vignette_2>
(Rmd: `vignettes/seurat5_spatial_vignette_2.Rmd`, Seurat 5.5.0, compiled April 28, 2026).

This vignette covers three imaging-based spatial platforms, each in its own
section:

- Vizgen MERSCOPE (MERFISH), mouse brain
- 10x Genomics Xenium In Situ, mouse brain
- Nanostring CosMx Spatial Molecular Imager (SMI), FFPE human lung

::: {.callout-important}
## Status: transcribed from upstream, first render is a test

The R code below is transcribed verbatim from the upstream vignette, in order,
with the hardcoded `/brahms/...` input paths replaced by reads from the staged
workshop data repo. The one substantive departure is the Vizgen section, which
reads a pre-built Seurat object rather than calling `LoadVizgen()`, for the
reasons given there. It has **not** been executed yet. `error: true` is set in
the YAML on purpose, so this first test render reports every failure at once
rather than stopping at the first one. All chunks are `eval: true`.
:::

::: {.callout-warning}
## These datasets are the main staging burden

Unlike sequencing-based Visium (one matrix plus one image), imaging platforms
ship as **directories of many raw files** (cell x gene matrix, per-transcript
molecule coordinates, cell segmentation polygons, centroids, metadata). The
loaders `LoadVizgen()`, `LoadXenium()`, and `LoadNanostring()`
each expect a specific directory layout. These are large multi-file downloads
and are the biggest staging job of the whole workshop. Per-platform expected
layouts are stated in callouts in each section below.
:::

## Setup

```{r}
#| label: setup
source(here::here("slides", "find_data_dir.R"))
data_dir <- find_data_dir("spatial_imaging")
# allen_cortex.rds is shared with the Visium tutorial and staged once at the
# top level of the data dir, so read it from the root rather than data_dir.
data_root <- find_data_dir()
report_data_dir("spatial_imaging")
```

```{r}
#| label: libraries
library(Seurat)
library(future)

# The CGU teaching cluster gives each OnDemand RStudio session 4 logical CPUs
# (= 2 physical cores, the nodes are 2 sockets x 10 cores x 2 threads) and a
# 16 GB memory reservation. Match `workers` to physical cores, not threads.
#
# `multisession` workers are separate R processes with no shared memory, so
# `future` serializes every global referenced in a parallel expression and
# ships a full copy to each worker. N workers therefore means N + 1 resident
# copies of the Seurat object. At the upstream vignette's `workers = 10` a
# 1.5 GB object already exceeds 16 GB.
#
# Slurm is not enforcing per-job memory limits on these nodes (no
# /etc/slurm/cgroup.conf on the compute nodes, JobAcctGatherType unset), so
# overshooting is not a clean per-job kill. The node-level OOM killer fires
# and can take out the other students sharing the node.
workers_n <- 2
plan("multisession", workers = workers_n)
library(ggplot2)

# Guardrail, not a cap. This is the size at which `future` refuses to export a
# global and raises an error. Keeping it near the per-worker copy we can afford
# turns "silently make 3 copies and get OOM-killed" into an informative error.
# The upstream vignette's 10 GB silences exactly the warning students need.
options(future.globals.maxSize = 2 * 1024^3)
```

## Mouse Brain: Vizgen MERSCOPE

This dataset was produced using the Vizgen MERSCOPE system, which uses MERFISH
technology. The full public dataset contains nine samples (three full coronal
slices of the mouse brain, three biological replicates per slice). The gene
panel has 483 targets. Here we analyze one sample, slice 2 replicate 1
(`s2r1`). The median number of transcripts detected per cell is 206.

Upstream, the dataset is read with `LoadVizgen()`, which parses the output of
the Vizgen analysis pipeline. The resulting Seurat object holds a count matrix
(483 transcripts per cell, stored in the `Vizgen` assay), plus imaging-specific
slots for cell centroids, cell segmentation boundaries, and individual
molecule positions.

We read a pre-built copy of exactly that object instead, for the reasons in the
callout below.

::: {.callout-important}
## Why this section reads a staged object rather than calling `LoadVizgen()`

`LoadVizgen()` on this dataset takes about **41 minutes** and peaks near
**16 GiB** of memory, almost all of it spent parsing 1242 `HDF5` cell-boundary
files one at a time. It also overflows R's pointer protection stack while
assembling the segmentation polygons, and dies with
`protect(): protection stack overflow` unless R is started with
`--max-ppsize=500000`. That flag can only be set when R starts, so no chunk in
this document could fix it.

`slides/wed_afternoon/staging/stage_vizgen_object.R` does that work once and
saves the result. Reading it back takes about a minute.
:::

::: {.callout-note}
## Expected staged layout: Vizgen MERSCOPE

This tutorial needs one file, `spatial_imaging/vizgen_mouse_brain_s2r1.rds`.

To rebuild it you need the raw MERSCOPE pipeline outputs, which are gated
behind vendor registration, staged at
`spatial_imaging/vizgen_mouse_brain/s2r1/`: the cell-by-gene matrix
(`cell_by_gene.csv`), cell metadata (`cell_metadata.csv`), the detected
transcripts / molecule positions (`detected_transcripts.csv`), and the cell
boundary polygons (an `HDF5` `cell_boundaries/` subdirectory or the newer
`cell_boundaries.parquet`). See `staging/MANUAL_DOWNLOADS.md`. Those raw files
total about 9.5 GiB and are **not** part of the workshop distribution, the
saved object is. Upstream original path was
`/brahms/hartmana/vignette_data/vizgen/s2r1/`.
:::

```{r}
#| label: vizgen-load
# Upstream calls LoadVizgen() here. See the callout above for why we do not.
#   vizgen.obj <- LoadVizgen(data.dir = file.path(data_dir, "vizgen_mouse_brain", "s2r1"),
#                            fov = "s2r1")
vizgen.obj <- readRDS(file.path(data_dir, "vizgen_mouse_brain_s2r1.rds"))
vizgen.obj
```

Cell centroids, one row per cell:

```{r}
#| label: vizgen-centroids
# Get the center position of each centroid. There is one row per cell in this dataframe.
head(GetTissueCoordinates(vizgen.obj[["s2r1"]][["centroids"]]))
```

Cell segmentation boundaries, the polygon vertices describing each cell:

```{r}
#| label: vizgen-segmentation
# Get the coordinates for each segmentation vertice. Each cell will have a variable number of
# vertices describing its shape.
head(GetTissueCoordinates(vizgen.obj[["s2r1"]][["segmentation"]]))
```

Molecule positions, the coordinates of each individual detected molecule:

```{r}
#| label: vizgen-molecules
# Fetch molecules positions for Chrm1
head(FetchData(vizgen.obj[["s2r1"]][["molecules"]], vars = "Chrm1"))
```

### Preprocessing and unsupervised analysis

We treat the data first as an scRNA-seq experiment. We use SCTransform-based
normalization, slightly modifying the default clipping to mitigate outliers
occasionally seen in smFISH data, then run dimensional reduction and
clustering.

```{r}
#| label: vizgen-cluster
vizgen.obj <- SCTransform(vizgen.obj, assay = "Vizgen", clip.range = c(-10, 10))
vizgen.obj <- RunPCA(vizgen.obj, npcs = 30, features = rownames(vizgen.obj))
vizgen.obj <- RunUMAP(vizgen.obj, dims = 1:30)
vizgen.obj <- FindNeighbors(vizgen.obj, reduction = "pca", dims = 1:30)
vizgen.obj <- FindClusters(vizgen.obj, resolution = 0.3)
```

Visualize the clustering in UMAP space with `DimPlot()` or overlaid on the
image with `ImageDimPlot()`.

```{r}
#| label: vizgen-dimplot
DimPlot(vizgen.obj, reduction = "umap")
```

```{r}
#| label: vizgen-imagedimplot
ImageDimPlot(vizgen.obj, fov = "s2r1", cols = "polychrome", axes = TRUE)
```

`ImageDimPlot()` and `ImageFeaturePlot()` accept several customization
parameters: `alpha` (within-cell transparency), `size` (centroid point size),
`cols` (internal shading palette, for example `polychrome`, `glasbey`,
`Paired`, `Set3`, `parade`), `shuffle.cols`, `border.size`, `border.color`,
`dark.background`, and `axes`.

Highlight the cells of a single cluster:

```{r}
#| label: vizgen-highlight
p1 <- ImageDimPlot(vizgen.obj, fov = "s2r1", cols = "red", cells = WhichCells(vizgen.obj, idents = 1))
p2 <- ImageDimPlot(vizgen.obj, fov = "s2r1", cols = "red", cells = WhichCells(vizgen.obj, idents = 15))
p1 + p2
```

Color cells by the quantified expression of a gene with `ImageFeaturePlot()`,
and, because MERFISH images individual molecules, show molecule locations too.

```{r}
#| label: vizgen-feature-molecules
p1 <- ImageFeaturePlot(vizgen.obj, features = "Slc17a7")
p2 <- ImageDimPlot(vizgen.obj, molecules = "Slc17a7", nmols = 10000, alpha = 0.3, mols.cols = "red")
p1 + p2
```

The `nmols` parameter reduces the number of molecules shown to limit
overplotting. `mols.size`, `mols.cols`, and `mols.alpha` further tune the
display. Plotting molecules is especially useful for co-visualizing several
genes.

```{r}
#| label: vizgen-comolecules
p1 <- ImageDimPlot(vizgen.obj, fov = "s2r1", alpha = 0.3, molecules = c("Slc17a7", "Olig1"), nmols = 10000)
markers.14 <- FindMarkers(vizgen.obj, ident.1 = "14")
p2 <- ImageDimPlot(vizgen.obj, fov = "s2r1", alpha = 0.3, molecules = rownames(markers.14)[1:4],
    nmols = 10000)
p1 + p2
```

By default Seurat treats each cell as a point (a centroid), which speeds up
plotting over large areas. We can zoom into a region by creating a new field of
view, then switch `DefaultBoundary()` to show full cell segmentations, and
optionally `Simplify()` the polygons to speed up plotting.

```{r}
#| label: vizgen-crop
# create a Crop
cropped.coords <- Crop(vizgen.obj[["s2r1"]], x = c(1750, 3000), y = c(3750, 5250), coords = "plot")
# set a new field of view (fov)
vizgen.obj[["hippo"]] <- cropped.coords

# visualize FOV using default settings (no cell boundaries)
p1 <- ImageDimPlot(vizgen.obj, fov = "hippo", axes = TRUE, size = 0.7, border.color = "white", cols = "polychrome",
    coord.fixed = FALSE)

# visualize FOV with full cell segmentations
DefaultBoundary(vizgen.obj[["hippo"]]) <- "segmentation"
p2 <- ImageDimPlot(vizgen.obj, fov = "hippo", axes = TRUE, border.color = "white", border.size = 0.1,
    cols = "polychrome", coord.fixed = FALSE)

# simplify cell segmentations
vizgen.obj[["hippo"]][["simplified.segmentations"]] <- Simplify(coords = vizgen.obj[["hippo"]][["segmentation"]],
    tol = 3)
DefaultBoundary(vizgen.obj[["hippo"]]) <- "simplified.segmentations"

# visualize FOV with simplified cell segmentations
DefaultBoundary(vizgen.obj[["hippo"]]) <- "simplified.segmentations"
p3 <- ImageDimPlot(vizgen.obj, fov = "hippo", axes = TRUE, border.color = "white", border.size = 0.1,
    cols = "polychrome", coord.fixed = FALSE)

p1 + p2 + p3
```

The `tol` parameter controls how aggressively segmentations are simplified. A
higher `tol` drops more vertices, speeding up plotting at the cost of some
detail.

Visualize individual molecules at higher resolution after zooming in:

```{r}
#| label: vizgen-zoom-molecules
# Since there is nothing behind the segmentations, alpha will slightly mute colors
ImageDimPlot(vizgen.obj, fov = "hippo", molecules = rownames(markers.14)[1:4], cols = "polychrome",
    mols.size = 1, alpha = 0.5, mols.cols = c("red", "blue", "yellow", "green"))
```

## Mouse Brain: 10x Genomics Xenium In Situ

This section analyzes data from the Xenium platform. `LoadXenium()` reads the
per-transcript molecule locations, the cell-by-gene matrix, cell
segmentations, and centroids. The RNA data is stored in the `Xenium` assay.

Upstream downloads the "Tiny subset" from the 10x Genomics
[Fresh Frozen Mouse Brain for Xenium Explorer Demo](https://www.10xgenomics.com/products/xenium-in-situ/preview-dataset-human-breast)
with:

```bash
wget https://cf.10xgenomics.com/samples/xenium/1.0.2/Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP/Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP_outs.zip
unzip Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP_outs.zip
```

::: {.callout-note}
## Expected staged layout: Xenium

`LoadXenium()` wants the **unzipped Xenium output directory**, staged at
`spatial_imaging/xenium_tiny_subset/`. That directory holds the standard Xenium
`outs`: `cell_feature_matrix.h5` (or the `cell_feature_matrix/` MTX
directory), `cells.csv.gz`, `cell_boundaries.csv.gz`,
`transcripts.csv.gz` (or their `.parquet` equivalents), and
`experiment.xenium`. Upstream original path was
`/brahms/hartmana/vignette_data/xenium_tiny_subset`. Installing `arrow` and
`hdf5r` speeds up loading.
:::

```{r}
#| label: xenium-load
path <- file.path(data_dir, "xenium_tiny_subset")
# Load the Xenium data
xenium.obj <- LoadXenium(path, fov = "fov", segmentations = "cell")
# remove cells with 0 counts
xenium.obj <- subset(xenium.obj, subset = nCount_Xenium > 0)
```

Standard QC violin plots via the `Xenium` assay:

```{r}
#| label: xenium-qc
VlnPlot(xenium.obj, features = c("nFeature_Xenium", "nCount_Xenium"), ncol = 2, pt.size = 0)
```

Plot molecule positions of marker genes with `ImageDimPlot()`:

```{r}
#| label: xenium-molecules
ImageDimPlot(xenium.obj, fov = "fov", molecules = c("Gad1", "Sst", "Pvalb", "Gfap"), nmols = 20000)
```

Per-cell expression of layer markers, with `max.cutoff` set roughly to each
gene's 90th percentile to improve contrast:

```{r}
#| label: xenium-feature
ImageFeaturePlot(xenium.obj, features = c("Cux2", "Rorb", "Bcl11b", "Foxp2"), max.cutoff = c(25,
    35, 12, 10), size = 0.75, cols = c("white", "red"))
```

Zoom in with `Crop()`, then show segmentations plus molecules:

```{r}
#| label: xenium-crop
cropped.coords <- Crop(xenium.obj[["fov"]], x = c(1200, 2900), y = c(3750, 4550), coords = "plot")
xenium.obj[["zoom"]] <- cropped.coords
# visualize cropped area with cell segmentations & selected molecules
DefaultBoundary(xenium.obj[["zoom"]]) <- "segmentation"
ImageDimPlot(xenium.obj, fov = "zoom", axes = TRUE, border.color = "white", border.size = 0.1, cols = "polychrome",
    coord.fixed = FALSE, molecules = c("Gad1", "Sst", "Npy2r", "Pvalb", "Nrn1"), nmols = 10000)
```

SCTransform normalization, dimensional reduction, and clustering:

```{r}
#| label: xenium-cluster
xenium.obj <- SCTransform(xenium.obj, assay = "Xenium")
xenium.obj <- RunPCA(xenium.obj, npcs = 30, features = rownames(xenium.obj))
xenium.obj <- RunUMAP(xenium.obj, dims = 1:30)
xenium.obj <- FindNeighbors(xenium.obj, reduction = "pca", dims = 1:30)
xenium.obj <- FindClusters(xenium.obj, resolution = 0.3)
```

```{r}
#| label: xenium-dimplot
DimPlot(xenium.obj)
```

```{r}
#| label: xenium-featureplot
FeaturePlot(xenium.obj, features = c("Cux2", "Bcl11b", "Foxp2", "Gad1", "Sst", "Gfap"))
```

```{r}
#| label: xenium-imagedimplot
ImageDimPlot(xenium.obj, cols = "polychrome", size = 0.75)
```

### Spatial niches with RCTD annotation

Using each cell's position we compute spatial niches. We use a cortex
reference from the Allen Brain Institute to annotate cells, cropping the
dataset to the cortex first. Below, `Slc17a7` expression helps locate the
cortical region.

```{r}
#| label: xenium-reload-crop
xenium.obj <- LoadXenium(file.path(data_dir, "xenium_tiny_subset"))
p1 <- ImageFeaturePlot(xenium.obj, features = "Slc17a7", axes = TRUE, max.cutoff = "q90")
p1
```

```{r}
#| label: xenium-crop-cortex
crop <- Crop(xenium.obj[["fov"]], x = c(600, 2100), y = c(900, 4700))
xenium.obj[["crop"]] <- crop
p2 <- ImageFeaturePlot(xenium.obj, fov = "crop", features = "Slc17a7", size = 1, axes = TRUE, max.cutoff = "q90")
p2
```

Seurat v5 supports Robust Cell Type Decomposition (RCTD) to deconvolve spatial
data using an scRNA-seq reference. RCTD is implemented in the `spacexr`
package, installed from GitHub:

```{r}
#| label: xenium-install-spacexr
devtools::install_github("dmcable/spacexr", build_vignettes = FALSE)
```

::: {.callout-note}
## Expected staged file: Allen cortex reference

The RCTD reference is a single `.rds`, staged once at the top level of the
workshop data dir as `allen_cortex.rds` (shared with the Visium tutorial).
Upstream downloads it from
<https://www.dropbox.com/s/cuowvm4vrf65pvq/allen_cortex.rds?dl=1> and read it
from `/brahms/shared/vignette-data/allen_cortex.rds`. This one is a plain
Seurat scRNA-seq object, so a staged `.rds` is the natural format (no directory
of raw files needed).
:::

```{r}
#| label: xenium-rctd
library(spacexr)

query.counts <- GetAssayData(xenium.obj, assay = "Xenium", layer = "counts")[, Cells(xenium.obj[["crop"]])]
coords <- GetTissueCoordinates(xenium.obj[["crop"]], which = "centroids")
rownames(coords) <- coords$cell
coords$cell <- NULL
query <- SpatialRNA(coords, query.counts, colSums(query.counts))
# allen.cortex.ref can be downloaded from:
# https://www.dropbox.com/s/cuowvm4vrf65pvq/allen_cortex.rds?dl=1
allen.cortex.ref <- readRDS(file.path(data_root, "allen_cortex.rds"))
allen.cortex.ref <- UpdateSeuratObject(allen.cortex.ref)

Idents(allen.cortex.ref) <- "subclass"
# remove CR cells because there aren't enough of them for annotation
allen.cortex.ref <- subset(allen.cortex.ref, subset = subclass != "CR")
counts <- GetAssayData(allen.cortex.ref, assay = "RNA", layer = "counts")
cluster <- as.factor(allen.cortex.ref$subclass)
names(cluster) <- colnames(allen.cortex.ref)
nUMI <- allen.cortex.ref$nCount_RNA
names(nUMI) <- colnames(allen.cortex.ref)
nUMI <- colSums(counts)
levels(cluster) <- gsub("/", "-", levels(cluster))
reference <- Reference(counts, cluster, nUMI)
# run RCTD with many cores
RCTD <- create.RCTD(query, reference, max_cores = 8)
RCTD <- run.RCTD(RCTD, doublet_mode = "doublet")
annotations.df <- RCTD@results$results_df
annotations <- annotations.df$first_type
names(annotations) <- rownames(annotations.df)
xenium.obj$predicted.celltype <- annotations
keep.cells <- Cells(xenium.obj)[!is.na(xenium.obj$predicted.celltype)]
xenium.obj <- subset(xenium.obj, cells = keep.cells)
```

Seurat v5 introduces niche analysis, which groups cells by the composition of
their local spatial neighborhood. `BuildNicheAssay()` builds a `niche` assay
of neighboring cell-type composition and returns a `niches` metadata column of
k-means cluster assignments.

```{r}
#| label: xenium-niche
xenium.obj <- BuildNicheAssay(object = xenium.obj, fov = "crop", group.by = "predicted.celltype",
    niches.k = 5, neighbors.k = 30)
```

Group cells by cell-type identity or by niche identity:

```{r}
#| label: xenium-niche-plot
celltype.plot <- ImageDimPlot(xenium.obj, group.by = "predicted.celltype", size = 1.5, cols = "polychrome",
    dark.background = F) + ggtitle("Cell type")
niche.plot <- ImageDimPlot(xenium.obj, group.by = "niches", size = 1.5, dark.background = F) + ggtitle("Niches") +
    scale_fill_manual(values = c("#442288", "#6CA2EA", "#B5D33D", "#FED23F", "#EB7D5B"))
celltype.plot | niche.plot
```

Each niche is enriched for distinct cell types:

```{r}
#| label: xenium-niche-table
table(xenium.obj$predicted.celltype, xenium.obj$niches)
```

## Human Lung: Nanostring CosMx Spatial Molecular Imager

This dataset comes from the Nanostring CosMx Spatial Molecular Imager (SMI),
which does multiplexed single-molecule profiling of RNA and protein and works
on FFPE tissue. The full dataset is 8 FFPE samples from 5 non-small-cell lung
cancer (NSCLC) tissues, with a 960-transcript panel. Here we load one sample,
lung 5 replicate 1.

`LoadNanostring()` parses the public download outputs. Instead of unsupervised
analysis, this section maps the profiles onto a precomputed Azimuth Healthy
Human Lung reference (Azimuth 0.4.3, human lung reference 1.0.0). The median
detected transcripts per cell is 249.

::: {.callout-note}
## Expected staged layout: Nanostring CosMx

`LoadNanostring()` wants a **directory** of CosMx outputs, staged at
`spatial_imaging/nanostring_lung5_rep1/`. It matches four files by pattern, and
all four must be in that one directory:

| Pattern | File here |
|---|---|
| `*_exprMat_file.csv` | `Lung5_Rep1_exprMat_file.csv`, 198 MB |
| `*_metadata_file.csv` | `Lung5_Rep1_metadata_file.csv`, 11.5 MB |
| `*_tx_file.csv` | `Lung5_Rep1_tx_file.csv`, 3.40 GB |
| `*-polygons.csv` | `Lung5_Rep1-polygons.csv`, 120 MB |

The first three come from the vendor download, which needs registration. The
cell boundary polygons come from elsewhere, a public Dropbox link, and are also
staged one level up at `spatial_imaging/Lung5_Rep1-polygons.csv` for
convenience. The copy inside the directory is a hard link to that one, because
`ReadNanostring()` will only look in `data.dir`. `..._fov_positions_file.csv`
ships with the download but is not one of the four the loader reads.

Upstream original path was
`/brahms/hartmana/vignette_data/nanostring/lung5_rep1`.

The precomputed Azimuth annotations are a single `.Rds` staged at
`spatial_imaging/nanostring_data.Rds` (a list with `$annotations` and `$umap`).
Upstream original path was `/brahms/hartmana/vignette_data/nanostring_data.Rds`,
downloadable from the Seurat vignette data site.
:::

```{r}
#| label: nanostring-load
nano.obj <- LoadNanostring(data.dir = file.path(data_dir, "nanostring_lung5_rep1"), fov = "lung5.rep1")
# add in precomputed Azimuth annotations
azimuth.data <- readRDS(file.path(data_dir, "nanostring_data.Rds"))
nano.obj <- AddMetaData(nano.obj, metadata = azimuth.data$annotations)
nano.obj[["proj.umap"]] <- azimuth.data$umap
Idents(nano.obj) <- nano.obj$predicted.annotation.l1

nano.obj <- SCTransform(nano.obj, assay = "Nanostring", clip.range = c(-10, 10), verbose = FALSE)

# text display of annotations and prediction scores
head(slot(object = nano.obj, name = "meta.data")[2:5])
```

Cells projected onto the reference-defined UMAP. For this NSCLC sample the
tumor cells are annotated as "basal", the closest cell type in the healthy
reference.

```{r}
#| label: nanostring-dimplot
DimPlot(nano.obj)
```

### Visualization of cell type and expression localization patterns

`ImageDimPlot()` plots cells by spatial location, colored by assigned cell
type. The basal (tumor) population is tightly spatially organized.

```{r}
#| label: nanostring-imagedimplot
ImageDimPlot(nano.obj, fov = "lung5.rep1", axes = TRUE, cols = "glasbey")
```

Highlight a few cell-type groups:

```{r}
#| label: nanostring-highlight
ImageDimPlot(nano.obj, fov = "lung5.rep1", cells = WhichCells(nano.obj, idents = c("Basal", "Macrophage",
    "Smooth Muscle", "CD4 T")), cols = c("red", "green", "blue", "orange"), size = 0.6)
```

Visualize the marker `KRT17` several ways:

```{r}
#| label: nanostring-vln
VlnPlot(nano.obj, features = "KRT17", assay = "Nanostring", layer = "counts", pt.size = 0.1, y.max = 30) +
    NoLegend()
```

```{r}
#| label: nanostring-featureplot
FeaturePlot(nano.obj, features = "KRT17", max.cutoff = "q95")
```

```{r}
#| label: nanostring-imagefeature
p1 <- ImageFeaturePlot(nano.obj, fov = "lung5.rep1", features = "KRT17", max.cutoff = "q95")
p2 <- ImageDimPlot(nano.obj, fov = "lung5.rep1", alpha = 0.3, molecules = "KRT17", nmols = 10000) +
    NoLegend()
p1 + p2
```

Co-visualize several markers: `KRT17` (basal cells), `C1QA` (macrophages),
`IL7R` (T cells), `TAGLN` (smooth muscle).

```{r}
#| label: nanostring-comolecules
# Plot some of the molecules which seem to display spatial correlation with each other
ImageDimPlot(nano.obj, fov = "lung5.rep1", group.by = NA, alpha = 0.3, molecules = c("KRT17", "C1QA",
    "IL7R", "TAGLN"), nmols = 20000)
```

Zoom in on a basal-rich region and show cell boundaries:

```{r}
#| label: nanostring-crop
basal.crop <- Crop(nano.obj[["lung5.rep1"]], x = c(159500, 164000), y = c(8700, 10500))
nano.obj[["zoom1"]] <- basal.crop
DefaultBoundary(nano.obj[["zoom1"]]) <- "segmentation"
ImageDimPlot(nano.obj, fov = "zoom1", cols = "polychrome", coord.fixed = FALSE)
```

```{r}
#| label: nanostring-zoom-molecules
# note the clouds of TPSAB1 molecules denoting mast cells
ImageDimPlot(nano.obj, fov = "zoom1", cols = "polychrome", alpha = 0.3, molecules = c("KRT17", "IL7R",
    "TPSAB1"), mols.size = 0.3, nmols = 20000, border.color = "black", coord.fixed = FALSE)
```

## Session Info

```{r}
#| label: sessioninfo
sessionInfo()
```
