---
title: "Visium HD cell segmentation: from 2 um bins to single cells"
subtitle: "Local version, reads from the workshop data repo, no network access"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: true
  error: true
---

Transcribed from `visiumhd_analysis_cell_segmentations.html`, the Seurat vignette
"Analysis and visualization of Visium HD data with cell segmentations in Seurat"
(compiled May 26, 2026),
<https://satijalab.org/seurat/articles/visiumhd_analysis_cell_segmentations>.

The vignette takes a Visium HD run that 10x Genomics' Space Ranger 4.0+ has
already segmented into single cells, loads both the 8 um bins and the
segmentation polygons, compares bins against segmented cells, and annotates the
segmented cells with Pan-human Azimuth. Source dataset: human kidney (cortex)
FFPE, Space Ranger 4.0.1, from the 10x data repository,
<https://www.10xgenomics.com/datasets/visium-hd-cytassist-gene-expression-libraries-of-human-kidney>.

::: {.callout-important}
## Status: transcribed, not yet executed

Every chunk is transcribed from the upstream vignette and none of them has been
run in this repository. This first render is the test. `error: true` is set on
purpose so that the render does not stop at the first failure and we get the
full list of problems in one pass.
:::

::: {.callout-note}
## Local data layout expected by the staging script

Upstream downloads seven files from the 10x CDN into a directory and then
un-tars three of them (see the "Load data" section below for the exact URLs).
`Load10X_Spatial()` reads the resulting Space Ranger output tree. Stage that
tree under the `spatial_visium_hd_segmentation` subdirectory of the workshop
data repo, so that `data_dir` (resolved in the setup chunk) is the value passed
as `data.dir` to `Load10X_Spatial()`.

The staged `spatial_visium_hd_segmentation/` directory should contain the
extracted Space Ranger outputs, that is, the contents of

- `Visium_HD_Human_Kidney_FFPE_binned_outputs.tar.gz` (the 2, 8, and 16 um
  binned expression matrices),
- `Visium_HD_Human_Kidney_FFPE_segmented_outputs.tar.gz` (segmentation-based
  expression data and the segmentation polygons), and
- `Visium_HD_Human_Kidney_FFPE_spatial.tar.gz` (the H&E tissue images, including
  `tissue_hires_image.png`, and the spatial coordinate mappings),

plus the loose files `Visium_HD_Human_Kidney_FFPE_barcode_mappings.parquet` and
`Visium_HD_Human_Kidney_FFPE_feature_slice.h5`. The `metrics_summary.csv` and
`molecule_info.h5` downloads are not read by this vignette.
:::

::: {.callout-warning}
## Cloud Azimuth needs network access, so it is not fully offline

The cell-type annotation step calls `AzimuthAPI::CloudAzimuth()`, which sends the
object to a 10x/Satija-lab cloud API and returns the annotations. That single
step cannot run without network access, unlike the rest of this vignette. For a
truly offline workshop the annotated object (or at least the Azimuth metadata
columns and the `azimuth_embed` reduction) should be precomputed at staging time
and merged into `kidney_obj`, so that everything downstream of the annotation
runs locally. As transcribed here the call is left in place so the first test
render shows exactly where the network dependency bites.
:::

::: {.callout-warning}
## Slow and high-memory steps, candidates for precomputing at staging time

Several steps are heavy and are good candidates to run once at staging time and
cache as an `.rds`:

- `Load10X_Spatial()` on the full run reads roughly 666,695 8 um bins and
  148,056 segmentation polygons into memory. Loading the high-resolution image
  (`tissue_hires_image.png`) makes the object substantially larger than the
  low-res default.
- `NormalizeData()` on the `Spatial.Polygons` assay, and the `CloudAzimuth()`
  round trip, both operate over all segmented cells.
- `RunUMAP()` over the 128-dimensional `azimuth_embed` reduction.

Precomputing `kidney_obj` (loaded, normalized, annotated, with the UMAP) and
staging it as a single object would let the workshop skip straight to the
visualizations.
:::

## Setup

```{r}
#| label: setup
source(here::here("slides", "find_data_dir.R"))
data_dir <- find_data_dir("spatial_visium_hd_segmentation")
report_data_dir("spatial_visium_hd_segmentation")
```

First, we load Seurat and the other packages necessary for this vignette. Core
functionality in support of Visium HD cell segmentation data was initially
introduced in Seurat v5.4.0 and SeuratObject v5.3.0.

```{r}
#| label: libraries
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
```

## Load data

Upstream, the data is downloaded and extracted from the terminal. These are the
exact commands from the vignette, kept here for reference only. They are shown,
not run, because the staged copy under `data_dir` already holds the extracted
outputs.

```bash
mkdir -p visium_hd_human_kidney
cd visium_hd_human_kidney

wget https://cf.10xgenomics.com/samples/spatial-exp/4.0.1/Visium_HD_Human_Kidney_FFPE/Visium_HD_Human_Kidney_FFPE_binned_outputs.tar.gz
wget https://cf.10xgenomics.com/samples/spatial-exp/4.0.1/Visium_HD_Human_Kidney_FFPE/Visium_HD_Human_Kidney_FFPE_segmented_outputs.tar.gz
wget https://cf.10xgenomics.com/samples/spatial-exp/4.0.1/Visium_HD_Human_Kidney_FFPE/Visium_HD_Human_Kidney_FFPE_spatial.tar.gz
wget https://cf.10xgenomics.com/samples/spatial-exp/4.0.1/Visium_HD_Human_Kidney_FFPE/Visium_HD_Human_Kidney_FFPE_barcode_mappings.parquet
wget https://cf.10xgenomics.com/samples/spatial-exp/4.0.1/Visium_HD_Human_Kidney_FFPE/Visium_HD_Human_Kidney_FFPE_feature_slice.h5
wget https://cf.10xgenomics.com/samples/spatial-exp/4.0.1/Visium_HD_Human_Kidney_FFPE/Visium_HD_Human_Kidney_FFPE_metrics_summary.csv
wget https://cf.10xgenomics.com/samples/spatial-exp/4.0.1/Visium_HD_Human_Kidney_FFPE/Visium_HD_Human_Kidney_FFPE_molecule_info.h5

tar -xzf Visium_HD_Human_Kidney_FFPE_binned_outputs.tar.gz
tar -xzf Visium_HD_Human_Kidney_FFPE_segmented_outputs.tar.gz
tar -xzf Visium_HD_Human_Kidney_FFPE_spatial.tar.gz
```

After downloading the data, we load it into Seurat using the `Load10X_Spatial()`
function. This function reads in the output of the Space Ranger count pipeline.
As of Space Ranger 4.0+, the count pipeline also automatically runs the segment
pipeline for Visium HD and Visium HD 3' data. We specify which data to load with
the `bin.size` argument. By including `"polygons"`, we load the
segmentation-based expression data where transcripts are assigned to individual
cells based on the nuclei detection and cell boundary expansion method. In this
vignette, we also load the 8 um-binned data for comparison.

The `image.name` argument was introduced in Seurat v5.4. Users specify either
`"tissue_lowres_image.png"` (default) or `"tissue_hires_image.png"`. The
high-resolution image is 10 times larger, so it is useful for generating higher
quality visualizations. Users should also consider object size and analysis
needs when deciding what to load.

Upstream sets `path_to_data <- "/brahms/shared/vignette-data/visium_hd_human_kidney"`.
Here we point it at the staged directory instead.

```{r}
#| label: load-data
path_to_data <- data_dir

kidney_obj <- Load10X_Spatial(data.dir = path_to_data,
                              bin.size = c(8, "polygons"),
                              image.name = "tissue_hires_image.png")
```

We can inspect the images and their associated assays to verify the data stored
in the object:

```{r}
#| label: inspect-images
Images(kidney_obj)
kidney_obj[["slice1.008um"]]
kidney_obj[["slice1.polygons"]]
```

## Bins vs. segmentations

One of the key advantages of the segmentation approach is the ability to analyze
and visualize data at single-cell resolution. Before we plot the data on the
whole tissue image, we can first define regions of interest in the tissue image
for focused analysis later.

In order to do so, we begin by inspecting the whole tissue image and its
coordinate ranges.

```{r}
#| label: tissue-image
tissue_image <- SpatialDimPlot(kidney_obj,
                               images = "slice1.008um",
                               alpha = 0,
                               image.scale = "hires") +
                    ggtitle("Tissue image") + NoLegend()

# Define layer to customize axis text
theme_layer <- theme(axis.text.x = element_text(size = 10, angle = 45),
                     axis.text.y = element_text(size = 10),
                     axis.title.x = element_text(size = 10),
                     axis.title.y = element_text(size = 10))

tissue_image_with_axes <- tissue_image +
                              scale_x_continuous(n.breaks = 20) +
                              scale_y_reverse(n.breaks = 20) +
                              theme_layer

tissue_image_with_axes
```

The approach we take here is to extract tissue coordinates and subset our data to
the same coordinate ranges for direct comparison. By using `Crop()`, we can
define a new zoomed field of view (FOV), stored in the images slot.

Here we define two different levels of zoom, based on coordinates from the tissue
image above. For the first level of zoom, we will take a section of 800 x 400
pixels, and apply this to the segmented data.

```{r}
#| label: zoom1
zoom1_xlim <- c(2400, 3200)
zoom1_ylim <- c(3300, 3700)

kidney_obj[["zoom1.polygons"]] <- Crop(object = kidney_obj[["slice1.polygons"]],
                                       x = zoom1_xlim,
                                       y = zoom1_ylim,
                                       scale = "hires")
```

For the second level of zoom, we will take a smaller section of 400 x 300 pixels,
and apply this to both the binned and segmented data.

```{r}
#| label: zoom2
zoom2_xlim <- c(2750, 3150)
zoom2_ylim <- c(3350, 3650)

kidney_obj[["zoom2.polygons"]] <- Crop(object = kidney_obj[["slice1.polygons"]],
                                       x = zoom2_xlim,
                                       y = zoom2_ylim,
                                       scale = "hires")

kidney_obj[["zoom2.008um"]] <- Crop(object = kidney_obj[["slice1.008um"]],
                                    x = zoom2_xlim,
                                    y = zoom2_ylim,
                                    scale = "hires")
```

To see the regions captured in these zooms, we plot just the tissue:

```{r}
#| label: zoom-tissue
tissue_image_zoom1 <- SpatialDimPlot(kidney_obj,
                                     images = "zoom1.polygons",
                                     alpha = 0,
                                     image.scale = "hires") +
                                  ggtitle("Zoom level 1") + NoLegend()

tissue_image_zoom2 <- SpatialDimPlot(kidney_obj,
                                     images = "zoom2.polygons",
                                     alpha = 0,
                                     image.scale = "hires") +
                                  ggtitle("Zoom level 2") + NoLegend()

tissue_image_zoom1 + tissue_image_zoom2
```

Now, we can take a closer look at how bins and cell segmentations capture spatial
information by comparing how they appear when overlaid on the tissue image. Let's
use the second level of zoom for this comparison, since it is more focused on a
smaller region of the tissue.

When generating a spatial plot of segmented data, `plot_segmentations = TRUE`
displays the full boundaries of each segmentation polygon, rather than just the
centroids. This allows us to see the actual shape and size of each detected cell.

```{r}
#| label: bins-vs-segmentations
sdp_8um_zoom <- SpatialDimPlot(kidney_obj,
                               images = "zoom2.008um",
                               pt.size.factor = 18,
                               stroke = 0.1,
                               alpha = 0.6,
                               image.scale = "hires",
                               cols = "turquoise4") +
                    ggtitle("8um bins") + NoLegend()

sdp_segm_zoom <- SpatialDimPlot(kidney_obj,
                                images = "zoom2.polygons",
                                plot_segmentations = TRUE,
                                stroke = 0.1,
                                alpha = 0.6,
                                image.scale = "hires",
                                cols = "turquoise4") +
                      ggtitle("Cell segmentations") + NoLegend()

sdp_8um_zoom + sdp_segm_zoom
```

## Cell type annotation with Pan-human Azimuth

We now annotate cell types in the segmentation-level data.

To perform cell type annotation, we use Pan-human Azimuth, a neural
network-based classifier trained on a reference scRNA-seq dataset of 23 different
types of human tissue. The classifier maps query cells to a hierarchical cell
ontology and provides confidence scores for each annotation. The original
implementation of this tool is in Python as `panhumanpy`. In R, we can access
Pan-human Azimuth through the `AzimuthAPI` package:

```{r}
#| label: install-azimuthapi
if (!requireNamespace("AzimuthAPI", quietly = TRUE)) {
  remotes::install_github("satijalab/AzimuthAPI")
}

library(AzimuthAPI)
```

### Annotating cell segmentations

`AzimuthAPI::CloudAzimuth()` runs Pan-human Azimuth on a Seurat object via a
cloud-based API. The object is returned with the results of cell type annotation
as added cell-level metadata, and the latent embeddings of the classifier as an
added dimensional reduction.

We set the default assay, normalize the data with standard log-normalization,
and then run `CloudAzimuth()`.

```{r}
#| label: cloud-azimuth
DefaultAssay(kidney_obj) <- "Spatial.Polygons"

kidney_obj <- NormalizeData(kidney_obj)
kidney_obj <- CloudAzimuth(kidney_obj,
                           assay = "Spatial.Polygons",
                           model_version = "v1")
```

First, we can inspect how the output of Pan-human Azimuth is stored within the
object metadata. Note that the object contains both 8 um bins and cell
segmentations. After running `CloudAzimuth` above on the `Spatial.Polygons`
assay, the Azimuth metadata columns are populated for (only) cell segmentations.

```{r}
#| label: azimuth-metadata
library(knitr)
knitr::kable(tail(kidney_obj@meta.data %>% select(4:last_col())))
```

We see that in addition to cell type annotations organized hierarchically,
Pan-human Azimuth provides confidence scores for each annotation. Lower
confidence scores indicate uncertainty in hierarchy assignments. We can set a
confidence threshold to retain only high-confidence annotations for downstream
tasks.

```{r}
#| label: subset-high-confidence
kidney_obj_hc <- subset(kidney_obj,
                        !is.na(final_level_confidence) & final_level_confidence > 0.7)
```

In addition to using the confidence score as a QC metric for annotation accuracy,
we also encourage users to examine differentially expressed genes for each
annotated cell type with `AzimuthAPI::make_azimuth_QC_heatmaps()`. For example,
we can look at a heatmap for cells annotated within the epithelial cell
hierarchy:

```{r}
#| label: azimuth-qc-heatmaps
kidney_azimuth_QC_heatmaps <- make_azimuth_QC_heatmaps(kidney_obj_hc)

kidney_azimuth_QC_heatmaps[["Epithelial cell_1"]]
```

Additionally, the `azimuth_embed` dimensional reduction layer stores a
128-dimensional reduction of the Pan-human Azimuth embeddings. This is useful in
downstream analyses. For example, we can visualize the cell type predictions
returned by Pan-human Azimuth in embedding space:

```{r}
#| label: azimuth-umap
kidney_obj_hc <- RunUMAP(kidney_obj_hc,
                         dims = 1:128,
                         reduction = "azimuth_embed",
                         reduction.name = "azimuth_umap")

p2 <- DimPlot(kidney_obj_hc,
              group.by = "final_level_labels",
              label.size = 2.5,
              label = TRUE,
              reduction = "azimuth_umap",
              repel = TRUE) + NoLegend()

p2
```

## Cell type annotations in spatial context

Next, we can visualize where different cell types are located within the tissue.
Mapping the cell type annotations onto the tissue image, we can observe the cell
types present in various anatomically distinct regions, such as glomeruli and
renal tubules.

Podocytes are highly specialized cells found within the glomeruli of the kidney.
We can focus our visualization on only cells annotated as podocytes:

```{r}
#| label: podocytes
kidney_segm_pod <- subset(kidney_obj_hc, final_level_labels == "Podocyte")

Idents(kidney_segm_pod) <- "final_level_labels"

pod_annotations_tissue <- SpatialDimPlot(kidney_segm_pod,
                                         images = "slice1.polygons",
                                         plot_segmentations = TRUE,
                                         image.scale = "hires",
                                         cols = "cyan") + NoLegend()

pod_annotations <- SpatialDimPlot(kidney_segm_pod,
                                  images = "slice1.polygons",
                                  image.alpha = 0,
                                  plot_segmentations = TRUE,
                                  image.scale = "hires",
                                  cols = "turquoise4") + NoLegend()

pod_annotations_tissue +
  pod_annotations +
  plot_annotation(title = 'Cells w/ annotation \'Podocyte\'')
```

We can pick out some additional cell types of interest to examine the spatial
organization of major kidney cell populations.

```{r}
#| label: cell-types-of-interest
cell_types_of_interest <- c("Capillary EC",
                            "Pericyte",
                            "Podocyte",
                            "Renal epithelial cell - distal tubules",
                            "Renal epithelial cell - Loop of Henle",
                            "Type A intercalated cell",
                            "Type B intercalated cell")

kidney_segm_top <- subset(kidney_obj_hc,
                          final_level_labels %in% cell_types_of_interest)

Idents(kidney_segm_top) <- "final_level_labels"

cell_type_colors <- c("#06d31e", "#006326", "#000000", "#453494", "#ffaf01", "#00fffa", "#1da5f9")
names(cell_type_colors) <- cell_types_of_interest

full_image_annotated <- SpatialDimPlot(kidney_segm_top,
                                       images = "slice1.polygons",
                                       plot_segmentations = TRUE,
                                       alpha = 0.7,
                                       stroke = 0.04,
                                       image.scale = "hires",
                                       cols = cell_type_colors) +
                            labs(fill = "Cell type")

full_image_annotated
```

## High-resolution zoom into tissue regions

Finally, zooming in again, we observe cell types annotated in and around some
glomeruli.

```{r}
#| label: glom-zoom
# Define layers to customize the legend by putting it in three rows below the plot
legend_guide_layer <- list(theme(legend.position = "bottom",
                                 legend.text = element_text(size = 12),
                                 legend.title = element_blank()),
                           guides(fill = guide_legend(nrow = 3, byrow = TRUE)))

glom_zoom_annotated <- SpatialDimPlot(kidney_segm_top,
                                      images = "zoom1.polygons",
                                      group.by = "final_level_labels",
                                      plot_segmentations = TRUE,
                                      alpha = 0.7,
                                      stroke = 0.1,
                                      image.scale = "hires",
                                      cols = cell_type_colors) +
                            legend_guide_layer

glom_zoom_annotated
```

In this section of kidney cortex, which includes several glomeruli surrounded by
renal tubules, the predicted cell type annotations align well with the underlying
tissue structure. Podocytes are confined to the glomeruli, while endothelial and
perivascular cells cluster within capillary-rich regions as expected. Distinct
tubular epithelial populations are distributed along surrounding tubules,
consistent with the normal spatial organization of nephron segments.

For more information on spatial analysis workflows, see the other spatial
vignettes and documentation (Seurat, <https://satijalab.org/seurat/>;
SeuratObject, <https://mojaveazure.github.io/seurat-object/>).

## Session info

```{r}
#| label: session-info
sessionInfo()
```
