---
title: "Introduction to Quarto"
subtitle: "Reproducible documents for analytical work<br/>Adapted from <https://r4ds.hadley.nz/quarto.html>"
author: "Steve Rozen"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    embed-resources: true
    code-line-numbers: false
    code-copy: true
    monofont: "Liberation Mono"
    mainfont: "Liberation Sans"
    css: quarto_intro_slides.css
---

## Why Quarto and Markdown?

- **Communicate results, not just code.** One file holds prose, code, and
  output (figures, tables, numbers).
- **Reproducible.** Re-render the document and every number, figure, and
  table is recomputed from the data.
- **Multi-format.** The same `.qmd` source renders to HTML, PDF, Word,
  slides, websites, and books.
- **Plain text.** Diffs cleanly in git, easy to review, easy to share.
- **Separates content from styling.** Authors write Markdown, Quarto handles
  layout.

## Why this matters for analysis

- No more "copy the number from the R console into the Word doc."
- No more "which version of the figure did I paste into the slide?"
- Reviewers and collaborators can rerun the analysis end-to-end.
- Future-you, six months from now, can rerun it too.

## What is a Quarto document?

A `.qmd` file is a plain-text file with three kinds of content:

1. A **YAML header** (metadata and output options).
2. **Markdown text** (prose, headings, lists, links).
3. **Code chunks** (executable R, Python, Julia, etc.).

You render it with `quarto render file.qmd` (or the Render button in
RStudio / VS Code / Positron).

## 1. The YAML header

``` yaml
---
title: "Diamond sizes"
author: "Hadley Wickham"
date: 2026-06-29
format: html
---
```

- Lives at the top, between `---` lines.
- Controls title, author, output format, and execution options.
- `format:` can be `html`, `pdf`, `docx`, `revealjs`, ...

## 2. Markdown text

```` markdown
# Heading 1
## Heading 2

*italic*, **bold**, `code`.

- bulleted
- list

1. numbered
2. list

[a link](https://quarto.org)
````

Lightweight conventions; render to any output format.

## 3. Code chunks

```` markdown
```{r}
#| label: setup
#| message: false
library(tidyverse)
```
````

- Fenced with triple backticks and the engine name in braces (`{r}`).
- Options go after `#|` at the top of the chunk.
- Output (text, tables, figures) is inserted into the rendered document.

## Common chunk options

| Option       | Effect                                          |
|--------------|-------------------------------------------------|
| `echo`       | Show the source code?                           |
| `eval`       | Run the code?                                   |
| `include`    | Include code *and* output? (`false` hides both) |
| `message`    | Show messages?                                  |
| `warning`    | Show warnings?                                  |
| `error`      | Keep rendering on error? (`true` to continue)   |
| `results`    | `"hide"` to suppress text output                |
| `fig-show`   | `"hide"` to suppress figures                    |

Set globally in YAML under `execute:` or per-chunk with `#|`.

## Figures

Control figure size and captions with chunk options:

``` r
#| label: fig-carat
#| fig-cap: "Diamond price vs carat."
#| fig-width: 6
#| fig-height: 4
#| out-width: "80%"
ggplot(diamonds, aes(carat, price)) + geom_hex()
```

- Label starting with `fig-` makes the chunk cross-referenceable.
- Reference it in prose as `@fig-carat`.

## Tables

```` markdown
```{r}
#| label: tbl-diamonds
#| tbl-cap: "First five diamonds."
knitr::kable(head(diamonds, 5))
```
````

- Label starting with `tbl-` makes it a cross-referenceable table.
- Reference in prose as `@tbl-diamonds`.
- For richer tables: `gt::gt()`, `DT::datatable()`, `kableExtra`.

## Inline code

Embed values directly in your prose:

``` markdown
We analyzed `r nrow(diamonds)` diamonds with a mean price of
$`r round(mean(diamonds$price))`.
```

Renders as:

> We analyzed 53940 diamonds with a mean price of $3933.

This is the single most important habit for keeping prose and data
in sync.

## Caching

For slow chunks:

``` r
#| cache: true
#| dependson: "raw_data"
```

- Saved results are reused until the chunk (or a `dependson` chunk)
  changes.
- Speeds up iterative writing; be careful when upstream data changes
  outside R's view.

## Editors

- **RStudio** has a *visual editor* (WYSIWYG-style) and a *source editor*
  (plain text). We will work in the **source editor** today.
- VS Code and Positron also have first-class Quarto support.
- A `.qmd` is just text, so any editor works.

## Typical workflow

1. Edit the `.qmd` in your editor.
2. Run chunks interactively to develop the analysis.
3. Render the whole document (`quarto render`) to catch reproducibility
   problems early.
4. Commit the `.qmd` (and usually not the rendered output) to git.

## Exercises, Quarto basics

1. Create a new Quarto document (File > New File > Quarto Document in
   RStudio). Read the instructions. Practice running the chunks
   individually. Render the document using the button and the keyboard
   shortcut. Modify the code, re-run, and verify the output changes.
2. Create one new Quarto document for each of the three built-in
   formats: HTML, PDF, and Word. Render each. How do the outputs
   differ? How do the inputs differ? (You may need to install LaTeX
   for PDF, RStudio will prompt you.)

## Exercises, source editor

1. Create a brief CV. Title is your name. Include headings for
   education or employment. Each section has a bulleted list of
   jobs / degrees. Put the year in **bold**.
2. Using the source editor and a Markdown quick reference, figure
   out how to:
   a. Add a footnote.
   b. Add a horizontal rule.
   c. Add a block quote.
3. Copy `diamond-sizes.qmd` from
   <https://github.com/hadley/r4ds/tree/main/quarto> into a local
   Quarto document. Run it, then add text after the frequency
   polygon describing its most striking features.

## Exercises, code chunks

1. Add a section that explores how diamond sizes vary by `cut`,
   `color`, and `clarity`. Assume the reader does not know R, so
   instead of setting `echo: false` on every chunk, set it as a
   global option.
2. Download `diamond-sizes.qmd` from
   <https://github.com/hadley/r4ds/tree/main/quarto>. Add a section
   that describes the largest 20 diamonds, including a table of
   their most important attributes.
3. Modify `diamond-sizes.qmd` to use `label_comma()` for nicely
   formatted output. Include the percentage of diamonds larger than
   2.5 carats.

## Exercises, figures

1. Find an image of a diamond, paste it into `diamond-sizes.qmd`,
   add a caption, resize it, and render. Note where the image is
   saved.
2. Edit the label of the plot chunk in `diamond-sizes.qmd` to start
   with `fig-` and add `fig-cap`. Add a cross-reference to the
   figure in the text above the chunk.
3. Change the figure size with each of these options in turn, render,
   and describe what changes:
   a. `fig-width: 10`
   b. `fig-height: 3`
   c. `out-width: "100%"`
   d. `out-width: "20%"`

## Exercises, tables

1. Insert a code chunk into `diamond-sizes.qmd` and use
   `knitr::kable()` to show the first 5 rows of `diamonds`.
2. Display the same table with `gt::gt()` instead.
3. Add a chunk label starting with `tbl-` and a caption via
   `tbl-cap`. Add a cross-reference to the table in the text above
   the chunk.

## Exercises, caching

1. Set up a network of chunks where `d` depends on `c` and `b`, and
   both `b` and `c` depend on `a`. Have each chunk print
   `lubridate::now()`, set `cache: true`, and verify your
   understanding of caching.

## Summary

- A `.qmd` is YAML header + Markdown + code chunks.
- Render with `quarto render` to keep prose, code, and output in sync.
- Chunk options control what is shown and what is run.
- Inline code (`` `r ` ``) keeps numbers in the prose accurate.
- Cross-reference figures (`fig-`) and tables (`tbl-`).
- Cache slow chunks so iteration stays fast.

Reference: <https://r4ds.hadley.nz/quarto.html>
