---
title: "Introduction to Quarto"
subtitle: "Reproducible documents for analytical work<br/>Adapted from <https://r4ds.hadley.nz/quarto.html>"
author: "Steve Rozen"
date: "2026-08-03"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    embed-resources: true
    code-line-numbers: false
    code-copy: true
    monofont: "Liberation Mono"
    mainfont: "Liberation Sans"
    css: quarto_intro_slides.css
editor: 
  markdown: 
    wrap: 72
---

## Why Quarto and Markdown?

- **Communicate results, methods, and code in one reproducible
  document**
- **Reproducible.** Re-render the document and every number, figure, and
  table is recomputed from the data.
- **Multi-format.** The same `.qmd` source renders to HTML, dynamic
  dashboards, PDF, Word, slides (and websites and books).
- **Plain text.** Diffs cleanly in git, easy to review, easy to share.

## Why this matters for analysis

- No more "copy the number from the R console into the Word doc."
- No more "which version of the figure did I paste into the slide?"
- Reviewers and collaborators can rerun the analysis end-to-end.
- Future-you, six months from now, can rerun it too.

## Use cases

- Communicate to decision-makers, who want to focus on the conclusions,
  not the code behind the analysis.

- Collaborate with other data scientists (including future you!), who
  are interested in both your conclusions, and how you reached them
  (i.e. the code).

- Use Quarto as a modern-day analysis notebook where you capture not
  only what you did, but also what you were thinking.

## What is a Quarto document?

A `.qmd` "Quarto markdown" file is a plain-text file with three kinds of
content:

1.  **YAML header** (metadata and output options).
2.  **Markdown text** (prose, headings, lists, links).
3.  **Code chunks** (executable R, Python, Julia, etc.).

You "render it" with `quarto render file.qmd` (or the Render button in
RStudio / VS Code / Positron)

"Render" means execute the code and do the formatting to a final output
format (e.g. HTML, dashboard, PDF, etc.) FYI, I wrote these slide in
Quarto.

## 1. The YAML header

``` yaml
---
title: "Diamond sizes"
author: "Hadley Wickham"
date: 2026-06-29
format: html
---
```

- Lives at the top, between `---` lines.
- Controls title, author, output format, and execution options.
- `format:` can be `html`, `pdf`, `docx`, `revealjs` (html slides),
  `dashboard` (html dashboard), ...

## 2.1 Markdown text examples

``` markdown

## Heading 2 (Most quarto docs use this the major section heading)

*italic*, **bold**, `code`, [a link](https://quarto.org)

- bullet
- list

1. numbered
2. list

| tiny | table |
-------|-------|
|  a   |   b   |
```

Lightweight conventions; render to any output format; easy to read as
plain text.

## Heading 2 (Technically but most quarto docs use this the major section heading)

*italic*, **bold**, `code`, [a link](https://quarto.org)

- bullet
- list

1.  numbered
2.  list

| tiny | table |
|------|-------|
| a    | b     |

## 3. Code chunks

```` markdown
```{{r}}
#| label: simple-addition

1 + 1
2 + 2
```
````

- Fenced with triple backticks and the engine name in braces (`{r}`).
- Options go after `#|` at the top of the chunk.
- Output (text, tables, figures) is inserted into the rendered document;
  simple-addition:

```{r}
#| label: simple-addition

1 + 1
2 + 2
```

## Common chunk options

| Option     | Effect                                          |
|------------|-------------------------------------------------|
| `echo`     | Show the source code?                           |
| `eval`     | Run the code?                                   |
| `include`  | Include code *and* output? (`false` hides both) |
| `message`  | Show messages?                                  |
| `warning`  | Show warnings?                                  |
| `error`    | Keep rendering on error? (`true` to continue)   |
| `results`  | `"hide"` to suppress text output                |
| `fig-show` | `"hide"` to suppress figures                    |

Set globally in YAML under `execute:` or per-chunk with `#|`.

## Figures

Control figure size and captions with chunk options:

```{r}
#| label: setup
library(tidyverse)
```

``` r
#| label: fig-carat
#| fig-cap: "Diamond price vs carat."
#| fig-width: 6
#| fig-height: 4
#| out-width: "80%"
ggplot(diamonds, aes(carat, price)) + geom_hex()
```

- Label starting with `fig-` makes the chunk cross-referenceable.
- Reference it in prose as `@fig-carat`.

## Figure output

```{r}
#| label: fig-carat
#| fig-cap: "Diamond price vs carat."
#| fig-width: 6
#| fig-height: 4
#| out-width: "80%"
ggplot(diamonds, aes(carat, price)) + geom_hex()
```


## More on figures

Sizing figures and the text in figures in Quarto is
tricky, because of mutliple frames of reference.
On the next slide is the output of

```{{r}}
library(tidyplots)
tidyplot(diamonds, carat, price) |>
    add(geom_hex(aes(fill = after_stat(count)), bins = 40)) |>
    add(scale_fill_viridis_c())
```

## Output from previous slide

```{r}
#| label: fig-carat2
library(tidyplots)
tidyplot(diamonds, carat, price) |>
    add(geom_hex(aes(fill = after_stat(count)), bins = 40)) |>
    add(scale_fill_viridis_c())
```

(`tidyplots` uses fixed output size by default)

## Try again

```{{r}}
tidyplot(diamonds, carat, price, height = NA, width = NA) |>
    add(geom_hex(aes(fill = after_stat(count)), bins = 40)) |>
    add(scale_fill_viridis_c())
```

```{r}
library(tidyplots)
tidyplot(diamonds, carat, price, height = NA, width = NA) |>
    add(geom_hex(aes(fill = after_stat(count)), bins = 40)) |>
    add(scale_fill_viridis_c())
```

## Try again again


```{{r}}
tidyplot(diamonds, carat, price, height = NA, width = NA) |>
    add(geom_hex(aes(fill = after_stat(count)), bins = 40)) |>
    add(scale_fill_viridis_c())
```

```{r}
#| fig-width: 4
#| fig-height: 4
library(tidyplots)
tidyplot(diamonds, carat, price, height = NA, width = NA) |>
    add(geom_hex(aes(fill = after_stat(count)), bins = 40)) |>
    add(scale_fill_viridis_c())
```

## Try yet again


```{{r}}
tidyplot(diamonds, carat, price, height = NA, width = NA) |>
    add(geom_hex(aes(fill = after_stat(count)), bins = 40)) |>
    add(scale_fill_viridis_c()) |>
    adjust_font(fontsize = 14)
```

```{r}
#| fig-width: 4
#| fig-height: 4
library(tidyplots)
tidyplot(diamonds, carat, price, height = NA, width = NA) |>
    add(geom_hex(aes(fill = after_stat(count)), bins = 40)) |>
    add(scale_fill_viridis_c()) |>
    adjust_font(fontsize = 14)
```

## Coding agents seem to often get figure sizes and placement and text sizes wrong at first; skills and SOPs can help

Guide to work arounds: can it be codified as an agent skill:

Different ways to specify font sizes in different parts of `ggplot2` and
different interactions with scaling in the output.

https://tidyverse.org/blog/2020/08/taking-control-of-plot-scaling/


## Figures --- wait, there's more

:::: {.columns}

::: {.column width="45%"}
```{.r style="font-size: 0.7em;"}
#| label: fig-carat-ggplotly
#| fig-cap: "Diamond price vs carat."
#| fig-width: 6
#| fig-height: 4
#| out-width: "80%"
p <- ggplot(diamonds, aes(carat, price)) +
  geom_hex()
plotly::ggplotly(p)
```
:::

::: {.column width="55%"}
```{r}
#| label: fig-carat-ggplotly
#| fig-cap: "Diamond price vs carat."
#| fig-width: 6
#| fig-height: 4
#| out-width: "80%"
p <- ggplot(diamonds, aes(carat, price)) + geom_hex()
plotly::ggplotly(p)
```
:::

::::

## Figures -- a more interesting ggplotly example

```{r}
#| label: fig-gapminder-ggplotly
data(gapminder, package = "gapminder")
gg <- ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(aes(size = pop, frame = year, ids = country)) +
  scale_x_log10()
plotly::ggplotly(gg)
```

## Figures -- code for previous slide

The key is the argument `frame = year` in the call to `geom_point`

```{.r}
data(gapminder, package = "gapminder")
gg <- ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(aes(size = pop, frame = year, ids = country)) +
  scale_x_log10()
plotly::ggplotly(gg)
```

You can also call `plotly` directly. There's a whole book by Carson Sieviert: https://plotly-r.com/


## Tables

```{{r}}
#| label: tbl-diamonds
#| tbl-cap: "First five diamonds."
knitr::kable(head(diamonds, 5))
```

 - Label starting with `tbl-` makes it a cross-referenceable table. -
Reference in prose as `@tbl-diamonds`. - For richer tables: `gt::gt()`,
`DT::datatable()`, `kableExtra`.

## Table output {.smaller}

```{r}
#| label: tbl-diamonds
#| tbl-cap: "First five diamonds."
knitr::kable(head(diamonds, 5))
```

## [The `DT` package for interactive tables]{style="font-size: 0.85em;"} {.smaller}

Sort by any column, search, page through rows, all client-side in the
rendered HTML.

```{r}
#| label: tbl-diamonds-dt
htmltools::div(
  style = "font-size: 0.65em;",
  DT::datatable(
    head(diamonds, 100),
    rownames  = FALSE,
    filter    = "top",
    options   = list(
      pageLength = 5,
      scrollX    = TRUE,
      order      = list(list(6, "desc"))
    )
  )
)
```

## Code for previous table

``` r
  DT::datatable(
    head(diamonds, 100),
    rownames  = FALSE,
    filter    = "top",
    options   = list(
      pageLength = 5,
      scrollX    = TRUE,
      order      = list(list(6, "desc"))
    )
  )
```

## Inline code

Embed values directly in your prose:

``` markdown
We analyzed `r knitr::inline_expr("nrow(diamonds)")` diamonds with a mean price of
$`r knitr::inline_expr("round(mean(diamonds$price))")`.
```

Renders as:

> We analyzed `r nrow(diamonds)` diamonds with a mean price of
> $`r round(mean(diamonds$price))`.

If the numbers come from your data / computation never ever ever *ever*
write them literally in your prose; if your data or computation change
then the numbers in the text will be out of sync.

## Caching

For slow chunks quarto provides a option to cache computations and
specify when the cached value needs to be re-computed. These
specificaitons can be error-prone and need to be done carefully.

## A functional caching pattern

````         
``` {{r}}
#| label: raw-data
#| cache: true
#| cache.extra: !expr file.mtime("a_very_large_file.csv")
rawdata <- readr::read_csv("a_very_large_file.csv")
```
````

Various writing here ...

````         
``` {{r}}
#| label: processed-data
#| cache: true
#| dependson: "raw-data"
processed_data <- rawdata |> 
  filter(!is.na(import_var)) |> 
  mutate(new_variable = complicated_transformation(x, y, z))
```
````

## Editors

- **RStudio** has a *visual editor* (What-you-see-is-what-you-get-style)
  and a *source editor* (plain text). We will work in the **source
  editor** today.
- VS Code and Positron also have first-class Quarto support.
- A `.qmd` is just text, so any editor works.

## Typical workflow

1.  Edit the `.qmd` in your editor.
2.  Run chunks interactively to develop the analysis.
3.  Render the whole document (`quarto render`) to catch reproducibility
    problems early.
4.  Commit the `.qmd` (and usually not the rendered output) to git.

## Debugging in Quarto

Demo how one does this, executing code blocks in quarto and then a copy of the code block in the console.

## Dynamic plots / plotly

## Exercises, Quarto basics

Markdown: https://quarto.org/docs/authoring/markdown-basics.html

Other cheatsheets: https://rstudio.github.io/cheatsheets/quarto.pdf

https://res.cloudinary.com/dyd911kmh/image/upload/v1676540721/Marketing/Blog/Quarto_Cheat_Sheet.pdf

1.  Create a new Quarto document (File \> New File \> Quarto Document in
    RStudio). Read the instructions. Practice running the chunks
    individually. Render the document using the button and the keyboard
    shortcut. Modify the code, re-run, and verify the output changes.
2.  Create one new Quarto document for each of the three built-in
    formats: HTML, PDF, and Word. Render each. How do the outputs
    differ? How do the inputs differ? (You may need to install LaTeX for
    PDF, RStudio will prompt you.)

## Exercises, source editor

1.  Create a brief CV. Title is your name. Include headings for
    education and employment. Each section has a bulleted list of jobs /
    degrees. Put the year in **bold**.
2.  Using the source editor and a Markdown quick reference, figure out
    how to:
    a.  Add a footnote.
    b.  Add a horizontal rule.
    c.  Add a block quote.
3.  Copy `diamond-sizes.qmd` from
    <https://github.com/hadley/r4ds/tree/main/quarto> into a local
    Quarto document. Run it, then add text after the frequency polygon
    describing its most striking features.

## Exercises, code chunks

1.  Add a section that explores how diamond sizes vary by `cut`,
    `color`, and `clarity`. Assume the reader does not know R, so
    instead of setting `echo: false` on every chunk, set it as a global
    option.
2.  Download `diamond-sizes.qmd` from
    <https://github.com/hadley/r4ds/tree/main/quarto>. Add a section
    that describes the largest 20 diamonds, including a table of their
    most important attributes.
3.  Modify `diamond-sizes.qmd` to use `label_comma()` for nicely
    formatted output. Include the percentage of diamonds larger than 2.5
    carats.

## Exercises, figures

1.  Find an image of a diamond, paste it into `diamond-sizes.qmd`, add a
    caption, resize it, and render. Note where the image is saved.
2.  Edit the label of the plot chunk in `diamond-sizes.qmd` to start
    with `fig-` and add `fig-cap`. Add a cross-reference to the figure
    in the text above the chunk.
3.  Change the figure size with each of these options in turn, render,
    and describe what changes:
    a.  `fig-width: 10`
    b.  `fig-height: 3`
    c.  `out-width: "100%"`
    d.  `out-width: "20%"`

## Exercises, tables

1.  Insert a code chunk into `diamond-sizes.qmd` and use
    `knitr::kable()` to show the first 5 rows of `diamonds`.
2.  Display the same table with `gt::gt()` instead.
3.  Add a chunk label starting with `tbl-` and a caption via `tbl-cap`.
    Add a cross-reference to the table in the text above the chunk.

## Quarto drawbacks / pain points

- Debugging
- Figure sizes and text sizes in figures
- Different output formats not all that uniform at the quarto level, e.g. embedded html, css files, 
- Caching
- Passing in parameters

## Summary

- A `.qmd` is YAML header + Markdown formatting + code chunks.
- Render with `quarto render` to keep prose, code, and output in sync.
- Chunk options control what is shown and what is run.

Why:

- **Communicate results, methods, and code in one reproducible
  document**
- **Reproducible.** Re-render the document and every number, figure, and
  table is recomputed from the data.
- **Multi-format.** The same `.qmd` source renders to HTML, dynamic
  dashboards, PDF, Word, slides (and websites and books).
- **Plain text.** Diffs cleanly in git, easy to review, easy to share.

Reference: <https://r4ds.hadley.nz/quarto.html>
