---
title: "The Complete ggplot2 Tutorial, Part 2, with tidyplots equivalents"
subtitle: "Selva Prabhakaran's theme-customizing guide, each step paired with the tidyplots way"
author: "Bioinformatics & Research Computing"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 6
    fig-height: 4
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
---

## About this handout

This follows Selva Prabhakaran's
[Complete ggplot2 Tutorial, Part 2: Customizing the look and feel](https://r-statistics.co/Complete-Ggplot2-Tutorial-Part2-Customizing-Theme-With-R-Code.html)
step by step. For each step we keep the original **ggplot2** code and plot, then
add the **tidyplots** code that produces the same figure.

Part 2 is about the *look and feel*: titles, legends, annotations, flipping,
faceting, and backgrounds. This is where the two packages differ most in
philosophy:

- **ggplot2** exposes one enormous `theme()` function with dozens of
  `element_text()` / `element_rect()` / `element_line()` slots.
- **tidyplots** wraps the common adjustments in small named verbs
  (`adjust_title()`, `adjust_legend_position()`, `adjust_font()`). For deep
  surgery it offers `adjust_theme_details(...)`, which accepts raw ggplot2
  `theme()` arguments, and the escape hatch `add(ggplot2::...)` for anything else.

The recurring lesson: tidyplots covers the frequent 90% with tidy verbs, and
hands you raw ggplot2 for the rest.

## Setup

```{r}
#| label: setup
library(ggplot2)
library(tidyplots)
library(ggrepel)   # repelling text labels, used by both packages here
data("midwest", package = "ggplot2")
data("mpg", package = "ggplot2")
```

Both packages reuse one base scatterplot through most of Part 2: `area` vs
`poptotal`, colored by `state`, sized by `popdensity`, with a loess fit. We build
it once each way.

```{r}
#| label: base-plots
# ggplot2 base
gg <- ggplot(midwest, aes(x = area, y = poptotal)) +
  geom_point(aes(col = state, size = popdensity)) +
  geom_smooth(method = "loess", se = FALSE) +
  xlim(c(0, 0.1)) + ylim(c(0, 500000)) +
  labs(title = "Area Vs Population", y = "Population", x = "Area",
       caption = "Source: midwest")

# tidyplots base. tidyplots' grouping aesthetic is color, so we keep point size
# static here rather than mapping it to popdensity (a size-by-variable mapping
# would itself need the ggplot2 escape hatch).
#
# Note the same color-grouping trap as Part 1: color = state is global, so a bare
# add_curve_fit() would fit one loess curve PER state (with a ribbon each). The
# ggplot2 base sets color only inside the points' aes(), so its geom_smooth() is a
# single global line with se = FALSE. We match that with the escape hatch:
# geom_smooth(aes(color = NULL, group = 1)) cancels the grouping for the fit only.
#
# Second trap, specific to the loess curve's SHAPE: ggplot2's xlim()/ylim() DELETE
# every point outside the box before the loess is fit (here the lone area = 0.11
# point and the high-poptotal points), so its smooth never sees them. tidyplots'
# adjust_x_axis(limits=)/adjust_y_axis(limits=) only ZOOM the view, they keep all
# points, so the loess would still fit the outliers and extrapolate to a different
# mode. To make the curve identical we filter the data to ggplot2's box up front.
midwest_box <- subset(midwest, area <= 0.1 & poptotal <= 500000)
tp <- midwest_box |>
  tidyplot(x = area, y = poptotal, color = state) |>
  add_data_points(size = 2) |>
  add(ggplot2::geom_smooth(ggplot2::aes(color = NULL, group = 1),
                           method = "loess", se = FALSE)) |>
  adjust_x_axis(limits = c(0, 0.1)) |>
  adjust_y_axis(limits = c(0, 500000)) |>
  adjust_title("Area Vs Population") |>
  adjust_x_axis_title("Area") |>
  adjust_y_axis_title("Population") |>
  adjust_caption("Source: midwest")
```

## 1. Plot and axis titles

In ggplot2 every text element is styled inside `theme()` with `element_text()`
(size, color, face, angle). tidyplots gives you focused verbs: `adjust_title()`
(and friends) take `fontsize`, `face`, and `color` directly, and
`adjust_x_axis(rotate_labels = ...)` rotates tick text. For the long tail of
elements, `adjust_theme_details()` accepts the same `element_text()` arguments as
ggplot2.

#### ggplot2

```{r}
#| label: titles-gg
gg + theme(
  plot.title  = element_text(size = 20, face = "bold", color = "tomato",
                             hjust = 0.5, lineheight = 1.2),
  plot.caption = element_text(size = 15),
  axis.title.x = element_text(vjust = 10, size = 15),
  axis.title.y = element_text(size = 15),
  axis.text.x  = element_text(size = 10, angle = 30, vjust = .5),
  axis.text.y  = element_text(size = 10))
```

#### tidyplots

```{r}
#| label: titles-tp
tp |>
  adjust_title("Area Vs Population", fontsize = 20, face = "bold",
               color = "tomato") |>
  adjust_x_axis_title(fontsize = 15) |>
  adjust_y_axis_title(fontsize = 15) |>
  adjust_x_axis(rotate_labels = 30) |>
  adjust_font(fontsize = 10)        # base font for the remaining text
```

## 2. Modifying the legend

### Change the legend title

ggplot2 offers three routes (`labs()`, `guides()`, `scale_*()`). tidyplots has one
verb: `adjust_legend_title()`. The base plot has two legends (color for `state`,
size for `popdensity`), and each verb names the one it targets.

#### ggplot2

```{r}
#| label: legtitle-gg
gg + labs(color = "State", size = "Density")
```

#### tidyplots

```{r}
#| label: legtitle-tp
tp |>
  adjust_legend_title("State")   # the color legend; size legend kept as is
```

### Change legend labels and point colors

ggplot2 uses `scale_color_manual(values = , labels = )`. tidyplots uses
`adjust_colors()`, which takes the colors in `new_colors` and the relabelled
categories in `labels`.

#### ggplot2

```{r}
#| label: leglabels-gg
gg + scale_color_manual(
  name   = "State",
  labels = c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin"),
  values = c("IL" = "blue", "IN" = "red", "MI" = "green",
             "OH" = "brown", "WI" = "orange"))
```

#### tidyplots

```{r}
#| label: leglabels-tp
tp |>
  adjust_legend_title("State") |>
  adjust_colors(
    new_colors = c("IL" = "blue", "IN" = "red", "MI" = "green",
                   "OH" = "brown", "WI" = "orange"),
    labels = c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin"))
```

### Change the order of the legends

Reordering which legend appears first is fine-grained `guides()` control with no
tidyplots verb, so this is a natural escape-hatch case.

#### ggplot2

```{r}
#| label: legorder-gg
gg + guides(colour = guide_legend(order = 1),
            size   = guide_legend(order = 2))
```

#### tidyplots

```{r}
#| label: legorder-tp
tp |>
  add(ggplot2::guides(colour = ggplot2::guide_legend(order = 1),
                      size   = ggplot2::guide_legend(order = 2)))
```

### Style the legend title, text, and key

Styling the legend key background and overriding key glyph sizes is again deep
`theme()` / `guides()` territory. tidyplots routes the `theme()` parts through
`adjust_theme_details()` and the `guides()` part through `add()`.

#### ggplot2

```{r}
#| label: legstyle-gg
gg + theme(legend.title = element_text(size = 12, color = "firebrick"),
           legend.text  = element_text(size = 10),
           legend.key   = element_rect(fill = "springgreen")) +
  guides(colour = guide_legend(override.aes = list(size = 2, stroke = 1.5)))
```

#### tidyplots

```{r}
#| label: legstyle-tp
tp |>
  adjust_theme_details(
    legend.title = ggplot2::element_text(size = 12, color = "firebrick"),
    legend.text  = ggplot2::element_text(size = 10),
    legend.key   = ggplot2::element_rect(fill = "springgreen")) |>
  add(ggplot2::guides(
    colour = ggplot2::guide_legend(override.aes = list(size = 2, stroke = 1.5))))
```

### Remove the legend and change its position

This is where tidyplots shines: `adjust_legend_position()` takes
`"none"`, `"left"`, `"right"`, `"top"`, or `"bottom"` in one verb. Placing the
legend *inside* the plot at arbitrary coordinates has no tidyplots verb, so that
last variant uses the escape hatch.

#### ggplot2

```{r}
#| label: legpos-gg
gg + theme(legend.position = "None") + labs(subtitle = "No Legend")
```

```{r}
#| label: legpos-gg2
gg + theme(legend.position = "bottom", legend.box = "horizontal") +
  labs(subtitle = "Legend at Bottom")
```

#### tidyplots

```{r}
#| label: legpos-tp
tp |>
  remove_legend()
```

```{r}
#| label: legpos-tp2
tp |>
  adjust_legend_position("bottom")
```

For a legend pinned inside the plot, drop back to ggplot2:

```{r}
#| label: legpos-tp3
tp |>
  add(ggplot2::theme(
    legend.justification = c(1, 0),
    legend.position      = c(0.95, 0.05),
    legend.background     = ggplot2::element_blank(),
    legend.key            = ggplot2::element_blank()))
```

## 3. Text, labels, and annotation

### Text and labels around points

The tutorial labels only the largest counties. We subset the data the same way,
then ggplot2 uses `geom_text()` / `ggrepel::geom_text_repel()` with that subset.
tidyplots has native verbs: `add_data_labels()` and `add_data_labels_repel()`,
both of which accept a `data =` argument for the subset.

```{r}
#| label: textprep
midwest_sub <- midwest[midwest$poptotal > 300000, ]
midwest_sub$large_county <- midwest_sub$county
```

#### ggplot2

```{r}
#| label: text-gg
gg + ggrepel::geom_text_repel(aes(label = large_county), size = 2,
                              data = midwest_sub) +
  labs(subtitle = "With ggrepel::geom_text_repel") +
  theme(legend.position = "None")
```

#### tidyplots

```{r}
#| label: text-tp
tp |>
  remove_legend() |>
  add_data_labels_repel(label = large_county, data = midwest_sub, fontsize = 7)
```

### Annotation anywhere inside the plot

ggplot2 builds a `grob` and places it with `annotation_custom()`. tidyplots has a
direct verb, `add_annotation_text()`, that takes the text and its x/y position in
data coordinates.

#### ggplot2

```{r}
#| label: annot-gg
library(grid)
my_grob <- grid::grobTree(grid::textGrob(
  "This text is at x=0.05 and y=400000!", x = 0.5, y = 0.8,
  gp = grid::gpar(col = "firebrick", fontsize = 12, fontface = "bold")))
gg + annotation_custom(my_grob)
```

#### tidyplots

```{r}
#| label: annot-tp
tp |>
  add_annotation_text("This text is at x=0.05 and y=400000!",
                      x = 0.05, y = 400000, fontsize = 12,
                      color = "firebrick", fontface = "bold")
```

## 4. Flipping and reversing axes

`coord_flip()` swaps x and y. tidyplots has a verb for exactly this:
`flip_plot()`. Reversing a scale is `scale_x_reverse()` in ggplot2 and
`adjust_x_axis(transform = "reverse")` in tidyplots.

#### ggplot2: flip

```{r}
#| label: flip-gg
gg + coord_flip() + theme(legend.position = "None")
```

#### tidyplots: flip

```{r}
#| label: flip-tp
tp |>
  remove_legend() |>
  flip_plot()
```

#### ggplot2: reverse both scales

Watch out for a gotcha here. The base `gg` clips with `xlim()`/`ylim()`, but those
are themselves x and y scales. Adding `scale_x_reverse()`/`scale_y_reverse()`
*replaces* the existing scales, so the limits are discarded and every clipped-out
county (including the huge-`popdensity` outliers up to 5,000,000) snaps back into
view, distorting the dot sizes and the axis range. The fix is to put the limits on
the reverse scales themselves, given high-to-low because the axis is reversed.

```{r}
#| label: rev-gg
gg +
  scale_x_reverse(limits = c(0.1, 0)) +     # reversed x, same clip as the base
  scale_y_reverse(limits = c(500000, 0)) +  # reversed y, same clip as the base
  theme(legend.position = "None")
```

#### tidyplots: reverse both scales

```{r}
#| label: rev-tp
tp |>
  remove_legend() |>
  adjust_x_axis(transform = "reverse") |>
  adjust_y_axis(transform = "reverse")
```

## 5. Faceting

Faceting switches to the `mpg` dataset: highway mileage (`hwy`) against engine
displacement (`displ`).

```{r}
#| label: facet-base
# ggplot2 base
gf <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()

# tidyplots base
tf <- mpg |>
  tidyplot(x = displ, y = hwy) |>
  add_data_points() |>
  # se = FALSE drops the confidence ribbon to match ggplot2's geom_smooth(se = FALSE)
  add_curve_fit(method = "lm", se = FALSE)
```

### Facet wrap

`facet_wrap(~ class)` becomes `split_plot(by = class)`. Both control the layout
(ggplot2's `nrow`, tidyplots' `ncol`/`nrow`).

There is one real difference to watch. ggplot2's `facet_wrap()` gives every panel
the *same* x and y range by default (`scales = "fixed"`), so panels are directly
comparable. tidyplots' `split_plot()` builds each panel as an independent plot, so
by default each one gets its own range. Its `scales = "fixed"` argument
synchronizes only the x axis here, the y (measure) axis still varies panel to
panel. To reproduce ggplot2's fully shared scales we use the same escape hatch as
the facet-grid example below, `ggplot2::facet_wrap()`, which keeps tidyplots'
styling but ggplot2's faceting (and its shared scales).

#### ggplot2

```{r}
#| label: wrap-gg
#| fig-width: 8
#| fig-height: 6
gf + facet_wrap(~ class, nrow = 3) +
  labs(title = "hwy vs displ", subtitle = "Faceting by class")
```

#### tidyplots

```{r}
#| label: wrap-tp
#| fig-width: 8
#| fig-height: 6
# split_plot(by = class) is the native verb, but it free-scales each panel.
# The escape hatch gives ggplot2's shared fixed scales so panels are comparable.
tf |>
  adjust_title("hwy vs displ") |>
  add(ggplot2::facet_wrap(~ class, ncol = 3))
```

### Facet grid (two variables)

`facet_grid(cyl ~ class)` lays one variable across rows and another across
columns. tidyplots can do this two ways. `split_plot(rows = cyl, cols = class)` is
the native verb, but it builds each cell as a standalone plot, so it repeats the
axis titles in every cell and draws no row/column strip labels (you cannot tell
which row is which `cyl`). The escape hatch `ggplot2::facet_grid()` keeps the
strips, but tidyplots' minimal theme does not render faceted panels (no axis ticks
or tick labels appear), so we re-apply a full ggplot2 theme afterward. Using
`theme_bw()` also matches the ggplot2 version above.

#### ggplot2

```{r}
#| label: grid-gg
#| fig-width: 8
#| fig-height: 6
gf + facet_grid(cyl ~ class) +
  labs(title = "hwy vs displ", subtitle = "Facet grid: cyl ~ class")
```

#### tidyplots

```{r}
#| label: grid-tp
#| fig-width: 8
#| fig-height: 6
tf |>
  adjust_title("hwy vs displ") |>
  add(ggplot2::facet_grid(cyl ~ class)) |>     # rows = cyl, cols = class
  # tidyplots' minimal theme leaves faceted panels without axes; re-apply a full
  # ggplot2 theme (theme_bw matches the ggplot2 version) to restore ticks + labels
  add(ggplot2::theme_bw())
```

## 6. Plot background, grid, and borders

These are pure `theme()` adjustments in ggplot2. tidyplots passes the same
arguments through `adjust_theme_details()`, so the translation is almost
mechanical: wrap the `theme(...)` contents in `adjust_theme_details(...)`.

### Change the plot background

#### ggplot2

```{r}
#| label: bg-gg
gf + theme(
  panel.background = element_rect(fill = "khaki"),
  panel.grid.major = element_line(colour = "burlywood", linewidth = 1.5),
  panel.grid.minor = element_line(colour = "tomato", linewidth = .25,
                                  linetype = "dashed"),
  panel.border = element_blank(),
  axis.line.x = element_line(colour = "darkorange", linewidth = 1.5),
  axis.line.y = element_line(colour = "darkorange", linewidth = 1.5)) +
  labs(title = "Modified Background")
```

#### tidyplots

```{r}
#| label: bg-tp
tf |>
  adjust_title("Modified Background") |>
  adjust_theme_details(
    panel.background = ggplot2::element_rect(fill = "khaki"),
    panel.grid.major = ggplot2::element_line(colour = "burlywood", linewidth = 1.5),
    panel.grid.minor = ggplot2::element_line(colour = "tomato", linewidth = .25,
                                             linetype = "dashed"),
    panel.border = ggplot2::element_blank(),
    axis.line.x = ggplot2::element_line(colour = "darkorange", linewidth = 1.5),
    axis.line.y = ggplot2::element_line(colour = "darkorange", linewidth = 1.5))
```

### Remove grid, border, axis title, text, and ticks

In ggplot2 you blank each element. tidyplots has dedicated `remove_*()` verbs for
the axis pieces, and `adjust_theme_details()` for the grid and border.

#### ggplot2

```{r}
#| label: strip-gg
gf + theme(
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.border = element_blank(),
  axis.title = element_blank(),
  axis.text  = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "Stripped Down")
```

#### tidyplots

```{r}
#| label: strip-tp
tf |>
  adjust_title("Stripped Down") |>
  remove_x_axis_title() |> remove_y_axis_title() |>
  remove_x_axis_labels() |> remove_y_axis_labels() |>
  remove_x_axis_ticks() |> remove_y_axis_ticks() |>
  # tidyplots' theme draws axis LINES (not a panel border), and the remove_*()
  # verbs above drop only ticks/labels/titles, so blank axis.line too
  adjust_theme_details(panel.grid.major = ggplot2::element_blank(),
                       panel.grid.minor = ggplot2::element_blank(),
                       panel.border     = ggplot2::element_blank(),
                       axis.line        = ggplot2::element_blank())
```

## Takeaways

- For the *frequent* customizations, tidyplots has a one-word verb where ggplot2
  needs a `theme()` slot: `adjust_legend_position()`, `flip_plot()`,
  `add_data_labels_repel()`, `add_annotation_text()`, and the `remove_*_axis_*()`
  family.
- For *deep* theme surgery (legend key fills, panel backgrounds, grid line
  styling), tidyplots does not invent new names. `adjust_theme_details()` takes
  the very same ggplot2 `theme()` arguments, so you reuse what you already know.
- A few things have no tidyplots verb at all: legend ordering, a legend pinned
  inside the plot, and a true two-way `facet_grid()`. Those use the escape hatch
  `add(ggplot2::...)`, the same pattern as Part 1.
- Net: Part 2 is mostly cosmetic, and tidyplots' value here is fewer, clearer
  verbs for the common cases, while staying fully ggplot2-compatible underneath.
- Original tutorial:
  <https://r-statistics.co/Complete-Ggplot2-Tutorial-Part2-Customizing-Theme-With-R-Code.html>
