---
title: "The Complete ggplot2 Tutorial, Part 1, with tidyplots equivalents"
subtitle: "Selva Prabhakaran's ggplot2 intro, each step paired with the tidyplots way"
author: "Bioinformatics & Research Computing"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 6
    fig-height: 4
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
---

## About this handout

This follows Selva Prabhakaran's
[Complete ggplot2 Tutorial, Part 1](https://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html)
step by step. For each step we keep the original **ggplot2** code and plot, then
add the **tidyplots** code that produces the same figure.

The two grammars in one line:

- **ggplot2** maps columns to aesthetics inside `aes()`, then stacks `geom_*()`
  and `scale_*()` layers with `+`.
- **tidyplots** pipes the data frame in, then chains `add_*()`, `adjust_*()`, and
  `remove_*()` verbs with the native pipe `|>`.

This tutorial is a good stress test for tidyplots, because it is built around a
*continuous* scatterplot with a regression line, which is not tidyplots' core use
case (tidyplots leans toward categorical statistical plots). Most steps map
cleanly. Where tidyplots has no native verb, we use its escape hatch
`add(ggplot2::...)`, which drops a raw ggplot2 layer into a tidyplot.

## Setup

```{r}
#| label: setup
options(scipen = 999)   # turn off scientific notation like 1e+06
library(ggplot2)        # the tutorial's package
library(tidyplots)      # the package we translate each step into
data("midwest", package = "ggplot2")   # 437 US-midwest counties
```

The `midwest` data has one row per county. We plot county `area` against total
population `poptotal`, later colored by `state`.

## 1. Understanding the ggplot syntax (the blank canvas)

Both start by declaring the data and the x/y mapping but drawing no geom, so both
panels are empty. In ggplot2 the mapping goes in `aes()`. In tidyplots it goes in
`tidyplot()`.

#### ggplot2

```{r}
#| label: blank-gg
ggplot(midwest, aes(x = area, y = poptotal))
```

#### tidyplots

```{r}
#| label: blank-tp
midwest |>
  tidyplot(x = area, y = poptotal)
```

## 2. A simple scatterplot (and a smoothing line)

Adding points is `geom_point()` in ggplot2 and `add_data_points()` in tidyplots.
The line of best fit is `geom_smooth(method = "lm")` in ggplot2. tidyplots wraps
the same machinery in `add_curve_fit(method = "lm")`.

#### ggplot2

```{r}
#| label: scatter-gg
ggplot(midwest, aes(x = area, y = poptotal)) +
  geom_point() +
  geom_smooth(method = "lm")   # se = FALSE turns off the confidence band
```

#### tidyplots

```{r}
#| label: scatter-tp
midwest |>
  tidyplot(x = area, y = poptotal) |>
  add_data_points() |>
  add_curve_fit(method = "lm")   # the linear fit, ribbon shown by default
```

## 3. Adjusting the x and y axis limits

The tutorial shows two distinct ideas. They look similar but behave differently,
and the distinction carries over to tidyplots.

### Method 1: delete points outside the range

`xlim()` / `ylim()` *drop* the points outside the range, so the regression line
is refit on what remains. In tidyplots, setting `limits` in `adjust_x_axis()` /
`adjust_y_axis()` does the same thing, it clips the data.

#### ggplot2

```{r}
#| label: lim-gg
ggplot(midwest, aes(x = area, y = poptotal)) +
  geom_point() +
  geom_smooth(method = "lm") +
  xlim(c(0, 0.1)) +
  ylim(c(0, 1000000))   # points outside are dropped, so the fit changes
```

#### tidyplots

```{r}
#| label: lim-tp
midwest |>
  tidyplot(x = area, y = poptotal) |>
  add_data_points() |>
  add_curve_fit(method = "lm") |>
  adjust_x_axis(limits = c(0, 0.1)) |>
  adjust_y_axis(limits = c(0, 1000000))   # clips data, just like xlim/ylim
```

### Method 2: zoom in without deleting points

`coord_cartesian()` zooms the viewport but keeps every point, so the fit is
unchanged. tidyplots has no native zoom verb, so this is a natural place to use
the escape hatch and add `ggplot2::coord_cartesian()` directly.

#### ggplot2

```{r}
#| label: zoom-gg
g <- ggplot(midwest, aes(x = area, y = poptotal)) +
  geom_point() +
  geom_smooth(method = "lm")
g + coord_cartesian(xlim = c(0, 0.1), ylim = c(0, 1000000))   # fit unchanged
```

#### tidyplots

```{r}
#| label: zoom-tp
midwest |>
  tidyplot(x = area, y = poptotal) |>
  add_data_points() |>
  add_curve_fit(method = "lm") |>
  # escape hatch: tidyplots has no coord_cartesian verb, so drop in ggplot2's
  add(ggplot2::coord_cartesian(xlim = c(0, 0.1), ylim = c(0, 1000000)))
```

## 4. Title and axis labels

ggplot2 sets every label in one `labs()` call. tidyplots spreads them across
`adjust_*` verbs: `adjust_title()`, `adjust_x_axis_title()`,
`adjust_y_axis_title()`, and `adjust_caption()`. tidyplots has no subtitle verb,
so we fold the subtitle into the caption (or use the escape hatch with
`ggplot2::labs(subtitle = ...)`).

#### ggplot2

```{r}
#| label: titles-gg
ggplot(midwest, aes(x = area, y = poptotal)) +
  geom_point() +
  geom_smooth(method = "lm") +
  coord_cartesian(xlim = c(0, 0.1), ylim = c(0, 1000000)) +
  labs(title    = "Area Vs Population",
       subtitle = "From midwest dataset",
       y        = "Population",
       x        = "Area",
       caption  = "Midwest Demographics")
```

#### tidyplots

```{r}
#| label: titles-tp
midwest |>
  tidyplot(x = area, y = poptotal) |>
  add_data_points() |>
  add_curve_fit(method = "lm") |>
  add(ggplot2::coord_cartesian(xlim = c(0, 0.1), ylim = c(0, 1000000))) |>
  adjust_title("Area Vs Population") |>
  adjust_x_axis_title("Area") |>
  adjust_y_axis_title("Population") |>
  adjust_caption("Midwest Demographics") |>
  add(ggplot2::labs(subtitle = "From midwest dataset"))   # no native subtitle verb
```

## 5. Color and size of points

### Static color and size

When a color or size is a *fixed value* (not tied to data), it goes outside
`aes()` in ggplot2. In tidyplots it goes as a plain argument to the `add_*()`
verb.

#### ggplot2

```{r}
#| label: static-gg
ggplot(midwest, aes(x = area, y = poptotal)) +
  geom_point(col = "steelblue", size = 3) +
  geom_smooth(method = "lm", col = "firebrick") +
  coord_cartesian(xlim = c(0, 0.1), ylim = c(0, 1000000)) +
  labs(title = "Area Vs Population", y = "Population", x = "Area",
       caption = "Midwest Demographics")
```

#### tidyplots

```{r}
#| label: static-tp
midwest |>
  tidyplot(x = area, y = poptotal) |>
  add_data_points(color = "steelblue", size = 3) |>
  add_curve_fit(method = "lm", color = "firebrick") |>
  add(ggplot2::coord_cartesian(xlim = c(0, 0.1), ylim = c(0, 1000000))) |>
  adjust_title("Area Vs Population") |>
  adjust_x_axis_title("Area") |>
  adjust_y_axis_title("Population") |>
  adjust_caption("Midwest Demographics")
```

### Color by a category

To color by a *column*, ggplot2 puts `col = state` inside `aes()`. tidyplots puts
`color = state` in the initial `tidyplot()` call. The legend appears automatically
in both.

#### ggplot2

```{r}
#| label: bycat-gg
gg <- ggplot(midwest, aes(x = area, y = poptotal)) +
  geom_point(aes(col = state), size = 3) +
  geom_smooth(method = "lm", col = "firebrick", size = 2) +
  coord_cartesian(xlim = c(0, 0.1), ylim = c(0, 1000000)) +
  labs(title = "Area Vs Population", subtitle = "From midwest dataset",
       y = "Population", x = "Area", caption = "Midwest Demographics")
gg
```

#### tidyplots

There is a subtle trap here. In tidyplots `color = state` is set globally in
`tidyplot()`, so **every** later layer inherits that grouping, including
`add_curve_fit()`. That would fit one regression line *per state*, whereas the
ggplot2 version fits a single global line (because its `col = "firebrick"` sits
*outside* `aes()` in `geom_smooth()`, carrying no grouping).

To match ggplot2's single line we drop the inherited grouping for the fit only.
The cleanest way is the escape hatch: a raw `geom_smooth()` with
`aes(color = NULL, group = 1)`, which removes the color mapping and forces one
group. The points keep their per-state colors.

```{r}
#| label: bycat-tp
tp <- midwest |>
  tidyplot(x = area, y = poptotal, color = state) |>
  add_data_points(size = 3) |>
  # single global fit: aes(color = NULL, group = 1) cancels the inherited grouping
  add(ggplot2::geom_smooth(ggplot2::aes(color = NULL, group = 1),
                           method = "lm", color = "firebrick", linewidth = 1)) |>
  add(ggplot2::coord_cartesian(xlim = c(0, 0.1), ylim = c(0, 1000000))) |>
  adjust_title("Area Vs Population") |>
  adjust_x_axis_title("Area") |>
  adjust_y_axis_title("Population") |>
  adjust_caption("Midwest Demographics")
tp
```

**What to notice:** this is a real difference in the two grammars, not a bug. A
global `color` aesthetic in tidyplots groups *all* statistical layers by default.
If you instead want a per-group fit (one line per state), keep the original
`add_curve_fit(method = "lm")` and drop the `color = "firebrick"` override. The
ggplot2 equivalent of *that* would be moving `color = state` into the shared
`aes()` so `geom_smooth()` inherits it too.

### Remove the legend

`theme(legend.position = "None")` in ggplot2. `remove_legend()` in tidyplots.

#### ggplot2

```{r}
#| label: nolegend-gg
gg + theme(legend.position = "None")
```

#### tidyplots

```{r}
#| label: nolegend-tp
tp |>
  remove_legend()
```

### Change the color palette

The tutorial swaps in RColorBrewer's "Set1" via `scale_colour_brewer()`. In
tidyplots you pass a color vector to `adjust_colors()`. You can hand it the exact
same brewer palette, or use one of tidyplots' own built-in schemes such as
`colors_discrete_friendly`.

#### ggplot2

```{r}
#| label: palette-gg
gg + scale_colour_brewer(palette = "Set1")
```

#### tidyplots

```{r}
#| label: palette-tp
# Option A: the very same RColorBrewer palette (midwest has 5 states)
tp |>
  adjust_colors(new_colors = RColorBrewer::brewer.pal(5, "Set1"))
```

```{r}
#| label: palette-tp2
# Option B: a tidyplots built-in colorblind-friendly scheme
tp |>
  adjust_colors(new_colors = colors_discrete_friendly)
```

## 6. X axis texts and tick locations

### Set the breaks

`scale_x_continuous(breaks = ...)` in ggplot2 becomes the `breaks` argument of
`adjust_x_axis()` in tidyplots.

#### ggplot2

```{r}
#| label: breaks-gg
gg + scale_x_continuous(breaks = seq(0, 0.1, 0.01))
```

#### tidyplots

```{r}
#| label: breaks-tp
tp |>
  adjust_x_axis(breaks = seq(0, 0.1, 0.01))
```

### Set the breaks and custom labels

`labels` takes a vector the same length as `breaks` in both packages. Here we use
the letters a-k just to show the mechanism.

#### ggplot2

```{r}
#| label: labels-gg
gg + scale_x_continuous(breaks = seq(0, 0.1, 0.01), labels = letters[1:11])
```

#### tidyplots

```{r}
#| label: labels-tp
tp |>
  adjust_x_axis(breaks = seq(0, 0.1, 0.01), labels = letters[1:11])
```

### Reverse the axis

`scale_x_reverse()` in ggplot2. tidyplots exposes `transform = "reverse"` on
`adjust_x_axis()`.

#### ggplot2

```{r}
#| label: reverse-gg
gg + scale_x_reverse()
```

#### tidyplots

```{r}
#| label: reverse-tp
tp |>
  adjust_x_axis(transform = "reverse")
```

### Format the axis label text

The tutorial formats the x axis as percentages (`sprintf`) and the y axis from
1000s to a "K" scale (custom function). Both packages accept a formatting function
in `labels`. tidyplots also has a shortcut: `cut_short_scale = TRUE` auto-abbreviates
large numbers (200000 becomes 200K), which is exactly the y-axis effect here.

#### ggplot2

```{r}
#| label: format-gg
gg +
  scale_x_continuous(breaks = seq(0, 0.1, 0.01),
                     labels = sprintf("%1.2f%%", seq(0, 0.1, 0.01))) +
  scale_y_continuous(breaks = seq(0, 1000000, 200000),
                     labels = function(x) paste0(x / 1000, "K"))
```

#### tidyplots

```{r}
#| label: format-tp
tp |>
  adjust_x_axis(breaks = seq(0, 0.1, 0.01),
                labels = sprintf("%1.2f%%", seq(0, 0.1, 0.01))) |>
  adjust_y_axis(breaks = seq(0, 1000000, 200000),
                cut_short_scale = TRUE)   # 200000 -> 200K automatically
```

### Pre-built themes

ggplot2 ships `theme_bw()`, `theme_classic()`, and more. tidyplots ships its own
family: the default `theme_tidyplot()`, plus `theme_minimal_xy()`,
`theme_minimal_x()`, `theme_minimal_y()`, and `theme_ggplot2()` (which imitates
ggplot2's grey look). You can also drop any ggplot2 theme into a tidyplot through
the escape hatch.

#### ggplot2

```{r}
#| label: theme-gg
gg2 <- gg + scale_x_continuous(breaks = seq(0, 0.1, 0.01))
gg2 + theme_bw() + labs(subtitle = "BW Theme")
```

```{r}
#| label: theme-gg2
gg2 + theme_classic() + labs(subtitle = "Classic Theme")
```

#### tidyplots

```{r}
#| label: theme-tp
# A tidyplots-native minimal theme (closest to theme_bw)
tp |>
  adjust_x_axis(breaks = seq(0, 0.1, 0.01)) |>
  theme_minimal_xy()
```

```{r}
#| label: theme-tp2
# Or drop a genuine ggplot2 theme in through the escape hatch
tp |>
  adjust_x_axis(breaks = seq(0, 0.1, 0.01)) |>
  add(ggplot2::theme_classic())
```

## Takeaways

- Most steps map one verb to one layer: `geom_point()` is `add_data_points()`,
  `geom_smooth(method = "lm")` is `add_curve_fit(method = "lm")`,
  `scale_colour_brewer()` is `adjust_colors()`, axis tweaks are
  `adjust_x_axis()` / `adjust_y_axis()`.
- ggplot2 bundles all labels into one `labs()`. tidyplots splits them across
  several `adjust_*` verbs, one concern per verb.
- This is a continuous scatterplot tutorial, which is outside tidyplots' main
  categorical-statistics wheelhouse. Two things needed the escape hatch
  `add(ggplot2::...)`: zooming with `coord_cartesian()` and adding a plot
  subtitle. That is the design promise, a tidy verb vocabulary that never traps
  you, because any tidyplot can absorb raw ggplot2 layers.
- tidyplots throws in conveniences ggplot2 makes you spell out, for example
  `cut_short_scale = TRUE` to abbreviate large axis numbers as "K".
- Original tutorial:
  <https://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html>
