---
title: "Elegant ggplot2 (First Steps), with tidyplots equivalents"
subtitle: "Slides version of `elegant_ggplot2_vs_tidyplot_2.qmd`"
author: "Steve Rozen (sr110@duke.edu, steverozen@pm.me)"
date: now
date-format: "YYYY-MM-DD HH:mm"
format:
  revealjs:
    embed-resources: true
    slide-number: true
    incremental: false
    code-fold: false
    code-line-numbers: false
    fig-width: 6
    fig-height: 4
    theme: simple
    width: 1280
    height: 800
execute:
  warning: false
  message: false
  echo: true
---

## About this deck

This follows the *First steps* chapter of Wickham, Navarro, and Pedersen,
[*ggplot2: Elegant Graphics for Data Analysis* (3rd ed.)](https://ggplot2-book.org/getting-started.html).

For each step we keep the original **ggplot2** code and plot on one slide, then
the **tidyplots** equivalent on the next.

The two grammars in one line:

- **ggplot2** maps columns to aesthetics inside `aes()`, then stacks `geom_*()`
  and `scale_*()` layers with `+`.
- **tidyplots** pipes the data frame in, then chains `add_*()`, `adjust_*()`,
  and `remove_*()` verbs with the native pipe `|>`.

## Two recurring frictions

1. **No transformed expressions in mappings.** ggplot2 takes
   `aes(y = unemploy / pop)` or `colour = factor(cyl)`. tidyplots' `tidyplot()`
   wants bare column names, so we precompute with `mutate()` first.
2. **No native verb for a few geoms** (frequency polygons, paths). For those
   we use the escape hatch `add(ggplot2::...)`, which drops a raw ggplot2 layer
   into a tidyplot.

## Setup --- ggplot2

```{r}
#| label: setup-gg
library(ggplot2)        # the book's package
library(dplyr)          # for the mutate()/filter() precompute steps
library(mgcv)           # for the GAM smoother example
data("mpg", package = "ggplot2")          # 234 car models
data("economics", package = "ggplot2")    # US economic time series
```

The `mpg` data has one row per car model, with engine displacement (`displ`),
highway mileage (`hwy`), city mileage (`cty`), drivetrain (`drv`), and `class`.

## Setup --- tidyplots

```{r}
#| label: setup-tp
library(tidyplots)      # the package we translate each step into
```

tidyplots layers on top of the same data and the same `dplyr` precompute
helpers loaded on the previous slide.

# 1. Key components: a basic scatterplot

## Basic scatterplot --- ggplot2

Every plot needs data, an aesthetic mapping, and at least one geom.

```{r}
#| label: scatter-gg
ggplot(mpg, aes(x = displ, y = hwy)) +   # data + x/y mapping
  geom_point()                           # draw points
```

## Basic scatterplot --- tidyplots

The mapping goes inside `tidyplot()`; points are `add_data_points()`.

```{r}
#| label: scatter-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points()
```

# 2. Colour, size, shape

## Mapping colour --- ggplot2

`aes(colour = class)` puts the variable in the mapping.

```{r}
#| label: aes-gg
ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point()
```

## Mapping colour --- tidyplots

`color = class` goes inside `tidyplot()`.

```{r}
#| label: aes-tp
mpg |>
  tidyplot(x = displ, y = hwy, color = class, width = NA, height = NA) |>
  add_data_points()
```

## Fixed colour --- ggplot2

A constant colour is *not* an aesthetic, it goes *outside* `aes()`.

```{r}
#| label: fixed-gg
ggplot(mpg, aes(displ, hwy)) +
  geom_point(colour = "blue")   # fixed colour, no legend
```

## Fixed colour --- tidyplots

A plain argument to the geom verb, no legend.

```{r}
#| label: fixed-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points(color = "blue")
```

# 3. Faceting

## Faceting --- ggplot2

`facet_wrap(~ class)` makes one panel per `class`, with shared scales.

```{r}
#| label: facet-gg
#| fig-width: 7
#| fig-height: 5
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~ class)
```

## Faceting --- tidyplots

`split_plot()` free-scales each panel; the escape hatch `facet_wrap()` keeps
ggplot2's shared scales.

```{r}
#| label: facet-tp
#| fig-width: 7
#| fig-height: 5
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points() |>
  add(ggplot2::facet_wrap(~ class))
```

# 4. Plot geoms

## Smoother --- ggplot2

```{r}
#| label: smooth-gg
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth()           # loess by default, with a ribbon
```

## Smoother --- tidyplots

```{r}
#| label: smooth-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points() |>
  add_curve_fit()         # loess by default, with a ribbon
```

## Wigglier loess (`span`) --- ggplot2

```{r}
#| label: span-gg
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(span = 0.2)
```

## Wigglier loess (`span`) --- tidyplots

```{r}
#| label: span-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points() |>
  add_curve_fit(span = 0.2)
```

## GAM smoother --- ggplot2

```{r}
#| label: gam-gg
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x))
```

## GAM smoother --- tidyplots

```{r}
#| label: gam-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points() |>
  add_curve_fit(method = "gam", formula = y ~ s(x))
```

## Linear fit --- ggplot2

```{r}
#| label: lm-gg
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(method = "lm")
```

## Linear fit --- tidyplots

```{r}
#| label: lm-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points() |>
  add_curve_fit(method = "lm")
```

## Overplotting at discrete x --- ggplot2

```{r}
#| label: overplot-gg
ggplot(mpg, aes(drv, hwy)) +
  geom_point()
```

## Overplotting at discrete x --- tidyplots

```{r}
#| label: overplot-tp
mpg |>
  tidyplot(x = drv, y = hwy, width = NA, height = NA) |>
  add_data_points()
```

## Jitter --- ggplot2

```{r}
#| label: jitter-gg
ggplot(mpg, aes(drv, hwy)) + geom_jitter()
```

## Jitter --- tidyplots

```{r}
#| label: jitter-tp
mpg |>
  tidyplot(x = drv, y = hwy, width = NA, height = NA) |>
  add_data_points_jitter(jitter_width = 0.8)
```

## Boxplot --- ggplot2

```{r}
#| label: box-gg
ggplot(mpg, aes(drv, hwy)) + geom_boxplot()
```

## Boxplot --- tidyplots

```{r}
#| label: box-tp
mpg |>
  tidyplot(x = drv, y = hwy, width = NA, height = NA) |>
  add_boxplot()
```

## Violin --- ggplot2

```{r}
#| label: violin-gg
ggplot(mpg, aes(drv, hwy)) + geom_violin()
```

## Violin --- tidyplots

```{r}
#| label: violin-tp
mpg |>
  tidyplot(x = drv, y = hwy, width = NA, height = NA) |>
  add_violin()
```

## Histogram --- ggplot2

```{r}
#| label: hist-gg
ggplot(mpg, aes(hwy)) + geom_histogram()
```

## Histogram --- tidyplots

```{r}
#| label: hist-tp
mpg |>
  tidyplot(x = hwy, width = NA, height = NA) |>
  add_histogram()
```

## Frequency polygon --- ggplot2

```{r}
#| label: freqpoly-gg
ggplot(mpg, aes(hwy)) + geom_freqpoly(binwidth = 2.5)
```

## Frequency polygon --- tidyplots

No native verb, use the escape hatch.

```{r}
#| label: freqpoly-tp
mpg |>
  tidyplot(x = hwy, width = NA, height = NA) |>
  add(ggplot2::geom_freqpoly(binwidth = 2.5))
```

## Freqpoly by group --- ggplot2

```{r}
#| label: freqpoly-grp-gg
ggplot(mpg, aes(displ, colour = drv)) +
  geom_freqpoly(binwidth = 0.5)
```

## Freqpoly by group --- tidyplots

```{r}
#| label: freqpoly-grp-tp
mpg |>
  tidyplot(x = displ, color = drv, width = NA, height = NA) |>
  add(ggplot2::geom_freqpoly(binwidth = 0.5))
```

## Histogram faceted by group --- ggplot2

```{r}
#| label: hist-facet-gg
#| fig-height: 5
ggplot(mpg, aes(displ, fill = drv)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(~ drv, ncol = 1)
```

## Histogram faceted by group --- tidyplots

```{r}
#| label: hist-facet-tp
#| fig-height: 5
mpg |>
  tidyplot(x = displ, color = drv, width = NA, height = NA) |>
  add_histogram(binwidth = 0.5) |>
  add(ggplot2::facet_wrap(~ drv, ncol = 1))
```

## Bar chart (counts) --- ggplot2

```{r}
#| label: bar-gg
ggplot(mpg, aes(manufacturer)) + geom_bar()
```

## Bar chart (counts) --- tidyplots

```{r}
#| label: bar-tp
mpg |>
  tidyplot(x = manufacturer, width = NA, height = NA) |>
  add_count_bar()
```

## Bar chart (identity) --- ggplot2

```{r}
#| label: bar-id-gg
drugs <- data.frame(drug = c("a", "b", "c"), effect = c(4.2, 9.7, 6.1))
ggplot(drugs, aes(drug, effect)) + geom_bar(stat = "identity")
```

## Bar chart (identity) --- tidyplots

```{r}
#| label: bar-id-tp
drugs <- data.frame(drug = c("a", "b", "c"), effect = c(4.2, 9.7, 6.1))
drugs |>
  tidyplot(x = drug, y = effect, width = NA, height = NA) |>
  add_sum_bar()
```

## Line plot --- ggplot2

ggplot2 computes the y ratio inside `aes()`.

```{r}
#| label: line-gg
ggplot(economics, aes(date, unemploy / pop)) +
  geom_line()
```

## Line plot --- tidyplots

tidyplots needs a real column, so we `mutate()` it first.

```{r}
#| label: line-tp
economics |>
  mutate(unemploy_rate = unemploy / pop) |>
  tidyplot(x = date, y = unemploy_rate, width = NA, height = NA) |>
  add_line()
```

## Path coloured by year --- ggplot2

```{r}
#| label: path-gg
year <- function(x) as.POSIXlt(x)$year + 1900
ggplot(economics, aes(unemploy / pop, uempmed)) +
  geom_path(colour = "grey50") +
  geom_point(aes(colour = year(date)))
```

## Path coloured by year --- tidyplots

Both transformed columns precomputed; the path goes in via the escape hatch.

```{r}
#| label: path-tp
economics |>
  mutate(unemploy_rate = unemploy / pop,
         year = as.POSIXlt(date)$year + 1900) |>
  tidyplot(x = unemploy_rate, y = uempmed, color = year,
           width = NA, height = NA) |>
  add(ggplot2::geom_path(ggplot2::aes(color = NULL), colour = "grey50")) |>
  add_data_points()
```

# 5. Modifying the axes

## Axis titles --- ggplot2

```{r}
#| label: xlab-gg
ggplot(mpg, aes(cty, hwy)) +
  geom_point(alpha = 1 / 3) +
  xlab("city driving (mpg)") +
  ylab("highway driving (mpg)")
```

## Axis titles --- tidyplots

```{r}
#| label: xlab-tp
mpg |>
  tidyplot(x = cty, y = hwy, width = NA, height = NA) |>
  add_data_points(alpha = 1 / 3) |>
  adjust_x_axis_title("city driving (mpg)") |>
  adjust_y_axis_title("highway driving (mpg)")
```

## Removing titles --- ggplot2

```{r}
#| label: nolab-gg
ggplot(mpg, aes(cty, hwy)) +
  geom_point(alpha = 1 / 3) +
  xlab(NULL) + ylab(NULL)
```

## Removing titles --- tidyplots

```{r}
#| label: nolab-tp
mpg |>
  tidyplot(x = cty, y = hwy, width = NA, height = NA) |>
  add_data_points(alpha = 1 / 3) |>
  remove_x_axis_title() |>
  remove_y_axis_title()
```

## Axis limits --- ggplot2

```{r}
#| label: lim-gg
ggplot(mpg, aes(drv, hwy)) +
  geom_jitter(width = 0.25) +
  xlim("f", "r") +
  ylim(20, 30)
```

## Axis limits --- tidyplots

No category-picking limit, so we filter the data instead.

```{r}
#| label: lim-tp
mpg |>
  dplyr::filter(drv %in% c("f", "r")) |>
  tidyplot(x = drv, y = hwy, width = NA, height = NA) |>
  add_data_points_jitter() |>
  adjust_y_axis(limits = c(20, 30))
```

# 6. Output

## Saving the plot

Both return an object you can store, print, and save. tidyplots returns a
ggplot under the hood, so `ggsave()` still works, but tidyplots also ships its
own `save_plot()` (note the argument order: the plot comes first).

```{r}
#| label: output-tp
#| eval: false
p <- mpg |>
  mutate(cyl_f = factor(cyl)) |>
  tidyplot(x = displ, y = hwy, color = cyl_f, width = NA, height = NA) |>
  add_data_points()

p                                                                 # print it
save_plot(p, "plot.png", width = 10, height = 10, units = "cm")   # tidyplots
ggplot2::ggsave("plot.png", p, width = 5, height = 5)             # also works
```

## Takeaways

- For the everyday plots in this chapter, tidyplots has a one-to-one verb:
  `add_data_points()`, `add_curve_fit()`, `add_boxplot()`, `add_violin()`,
  `add_histogram()`, `add_count_bar()`, `add_line()`.
- The grammar difference is where the mapping lives: inside `aes()` for ggplot2,
  inside `tidyplot()` for tidyplots, with grouping always carried by `color`.
- tidyplots will not evaluate transformed expressions in a mapping, so
  precompute with `mutate()`.
- It has no verb for a handful of geoms like `geom_freqpoly()` and `geom_path()`,
  so reach for the escape hatch `add(ggplot2::...)`.
- `split_plot()` is tidyplots' faceting verb, but it free-scales each panel.
  Use `add(ggplot2::facet_wrap(...))` when you need ggplot2's shared scales.

## References

- Wickham H, Navarro D, Pedersen TL. *ggplot2: Elegant Graphics for Data
  Analysis* (3rd edition). The source for this deck (chapter *First steps*).
  <https://ggplot2-book.org/getting-started.html>
- Engler JB (2025). "Tidyplots empowers life scientists with easy code-based
  data visualization." *iMeta*, e70018.
  [doi:10.1002/imt2.70018](https://doi.org/10.1002/imt2.70018)
- Wood SN (2011). "Fast stable restricted maximum likelihood and marginal
  likelihood estimation of semiparametric generalized linear models." *JRSS (B)*
  73(1):3-36.
  [doi:10.1111/j.1467-9868.2010.00749.x](https://doi.org/10.1111/j.1467-9868.2010.00749.x)
