---
title: "Elegant ggplot2 (First Steps), with tidyplots equivalents"
subtitle: "Source: `elegant_ggplot2_vs_tidyplot_2.qmd`"
author: "Steve Rozen (sr110@duke.edu, steverozen@pm.me)"
date: now
date-format: "YYYY-MM-DD HH:mm"
format:
  html:
    embed-resources: true
    page-layout: full
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 6
    fig-height: 4
    theme: cosmo
    include-in-header:
      text: |
        <style>
          .quarto-container, .page-columns, .main-container {
            max-width: 100% !important;
          }
        </style>
execute:
  warning: false
  message: false
  echo: true
---

## About this handout

This follows the *First steps* chapter of Hadley Wickham, Danielle Navarro, and
Thomas Lin Pedersen's
[*ggplot2: Elegant Graphics for Data Analysis* (3rd edition)](https://ggplot2-book.org/getting-started.html).
For each step we keep the original **ggplot2** code and plot, then add the
**tidyplots** code that produces the same figure.

The two grammars in one line:

- **ggplot2** maps columns to aesthetics inside `aes()`, then stacks `geom_*()`
  and `scale_*()` layers with `+`.
- **tidyplots** pipes the data frame in, then chains `add_*()`, `adjust_*()`, and
  `remove_*()` verbs with the native pipe `|>`.

This chapter is mostly scatterplots and distributions, so most steps map cleanly
to a tidyplots verb. Two recurring frictions show up and are worth naming up front:

1. **No transformed expressions in mappings.** ggplot2 happily takes
   `aes(y = unemploy / pop)` or `colour = factor(cyl)`. tidyplots' `tidyplot()`
   wants bare column names, so we precompute those columns with `mutate()` first.
2. **No native verb for a few geoms** (frequency polygons, paths). For those we
   use the escape hatch `add(ggplot2::...)`, which drops a raw ggplot2 layer into
   a tidyplot.

## Setup

```{r}
#| label: setup
library(ggplot2)        # the book's package
library(tidyplots)      # the package we translate each step into
library(dplyr)          # for the mutate()/filter() precompute steps
library(mgcv)           # for the GAM smoother example
data("mpg", package = "ggplot2")          # 234 car models
data("economics", package = "ggplot2")    # US economic time series
```

The `mpg` data has one row per car model, with engine displacement (`displ`),
highway mileage (`hwy`), city mileage (`cty`), drivetrain (`drv`), and `class`.

## 1. Key components: a basic scatterplot

Every plot needs data, an aesthetic mapping, and at least one geom. The mapping
goes inside `aes()` in ggplot2 and inside `tidyplot()` in tidyplots. Points are
`geom_point()` versus `add_data_points()`.

#### ggplot2

```{r}
#| label: scatter-gg
ggplot(mpg, aes(x = displ, y = hwy)) +   # data + x/y mapping
  geom_point()                           # draw points
```

#### tidyplots

```{r}
#| label: scatter-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>        # data + x/y mapping
  add_data_points()                      # draw points
```

## 2. Colour, size, shape, and other aesthetic attributes

To map a variable to colour, put it in the mapping: ggplot2's
`aes(colour = class)` becomes tidyplots' `color = class` inside `tidyplot()`.

#### ggplot2

```{r}
#| label: aes-gg
ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point()
```

#### tidyplots

```{r}
#| label: aes-tp
mpg |>
  tidyplot(x = displ, y = hwy, color = class, width = NA, height = NA) |>
  add_data_points()
```

**Mapped vs. fixed.** A constant colour is *not* an aesthetic, it is a fixed
argument. In ggplot2 it goes *outside* `aes()`, in tidyplots it goes as a plain
argument to the geom verb. (Putting `"blue"` inside the mapping instead makes
ggplot2 invent a one-level legend, the classic beginner trap.)

#### ggplot2

```{r}
#| label: fixed-gg
ggplot(mpg, aes(displ, hwy)) +
  geom_point(colour = "blue")   # fixed colour, no legend
```

#### tidyplots

```{r}
#| label: fixed-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points(color = "blue")   # fixed colour, no legend
```

## 3. Faceting

`facet_wrap(~ class)` makes one panel per `class`. tidyplots' native verb is
`split_plot(by = class)`, but there is a real difference to know about.
`facet_wrap()` gives every panel the *same* x and y range, while `split_plot()`
builds each panel as an independent plot and free-scales it, so every panel ends up
with its own y axis and the panels are no longer directly comparable. To match
ggplot2's shared scales we use the escape hatch `add(ggplot2::facet_wrap(~ class))`,
which keeps tidyplots' styling but ggplot2's faceting (and its shared scales).

#### ggplot2

```{r}
#| label: facet-gg
#| fig-width: 7
#| fig-height: 5
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~ class)
```

#### tidyplots

```{r}
#| label: facet-tp
#| fig-width: 7
#| fig-height: 5
mpg |>
  # tidyplots defaults to a fixed absolute panel size (50 x 50 mm, hence its
  # stubborn 1:1 panels). width = NA, height = NA lets the panels flex to fill
  # the figure like ggplot2, which is what makes faceting share scales cleanly.
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points() |>
  # split_plot(by = class) free-scales each panel; the escape hatch facet_wrap
  # keeps ggplot2's shared x and y scales so panels are comparable.
  add(ggplot2::facet_wrap(~ class))
```

## 4. Plot geoms

### 4.1 Adding a smoother

`geom_smooth()` adds a fitted curve with a confidence band. tidyplots wraps the
same machinery in `add_curve_fit()`. Both default to a loess fit. Extra arguments
(`se`, `span`, `method`, `formula`) pass straight through in both.

#### ggplot2

```{r}
#| label: smooth-gg
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth()           # loess by default, with a ribbon
```

#### tidyplots

```{r}
#| label: smooth-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points() |>
  add_curve_fit()         # loess by default, with a ribbon
```

**Controlling the wiggliness.** `span` sets how local the loess fit is (smaller is
wigglier). It passes through `add_curve_fit()` unchanged.

#### ggplot2

```{r}
#| label: span-gg
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(span = 0.2)   # very local, wiggly
```

#### tidyplots

```{r}
#| label: span-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points() |>
  add_curve_fit(span = 0.2)
```

**Other models.** A GAM and a straight line are just different `method=` (and for
the GAM, a `formula=`). Identical arguments in both packages.

#### ggplot2

```{r}
#| label: gam-gg
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x))
```

#### tidyplots

```{r}
#| label: gam-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points() |>
  add_curve_fit(method = "gam", formula = y ~ s(x))
```

#### ggplot2

```{r}
#| label: lm-gg
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(method = "lm")
```

#### tidyplots

```{r}
#| label: lm-tp
mpg |>
  tidyplot(x = displ, y = hwy, width = NA, height = NA) |>
  add_data_points() |>
  add_curve_fit(method = "lm")
```

### 4.2 Boxplots and jittered points

With a discrete x, raw points overplot badly. Three fixes: jitter, boxplot,
violin. tidyplots has a verb for each.

#### ggplot2

```{r}
#| label: overplot-gg
ggplot(mpg, aes(drv, hwy)) +
  geom_point()   # heavy overplotting at three x positions
```

#### tidyplots

```{r}
#| label: overplot-tp
mpg |>
  tidyplot(x = drv, y = hwy, width = NA, height = NA) |>
  add_data_points()
```

**Jitter.** `geom_jitter()` becomes `add_data_points_jitter()`.

#### ggplot2

```{r}
#| label: jitter-gg
ggplot(mpg, aes(drv, hwy)) + geom_jitter()
```

#### tidyplots

```{r}
#| label: jitter-tp
# add_data_points_jitter() defaults to jitter_width = 0.2 (the full spread).
# geom_jitter()'s default width = 0.4 is a half-width (points displaced +/- 0.4),
# so the full ggplot2 spread is ~0.8; set jitter_width = 0.8 to match.
mpg |> tidyplot(x = drv, y = hwy, width = NA, height = NA) |>
  add_data_points_jitter(jitter_width = 0.8)
```

**Boxplot.** `geom_boxplot()` becomes `add_boxplot()`.

#### ggplot2

```{r}
#| label: box-gg
ggplot(mpg, aes(drv, hwy)) + geom_boxplot()
```

#### tidyplots

```{r}
#| label: box-tp
mpg |> tidyplot(x = drv, y = hwy, width = NA, height = NA) |> add_boxplot()
```

**Violin.** `geom_violin()` becomes `add_violin()`.

#### ggplot2

```{r}
#| label: violin-gg
ggplot(mpg, aes(drv, hwy)) + geom_violin()
```

#### tidyplots

```{r}
#| label: violin-tp
mpg |> tidyplot(x = drv, y = hwy, width = NA, height = NA) |> add_violin()
```

### 4.3 Histograms and frequency polygons

A single continuous variable. `geom_histogram()` maps to `add_histogram()`.
tidyplots has no frequency-polygon verb, so `geom_freqpoly()` uses the escape
hatch.

#### ggplot2

```{r}
#| label: hist-gg
ggplot(mpg, aes(hwy)) + geom_histogram()
```

#### tidyplots

```{r}
#| label: hist-tp
mpg |> tidyplot(x = hwy, width = NA, height = NA) |> add_histogram()
```

**Frequency polygon (escape hatch).**

#### ggplot2

```{r}
#| label: freqpoly-gg
ggplot(mpg, aes(hwy)) + geom_freqpoly(binwidth = 2.5)
```

#### tidyplots

```{r}
#| label: freqpoly-tp
mpg |>
  tidyplot(x = hwy, width = NA, height = NA) |>
  add(ggplot2::geom_freqpoly(binwidth = 2.5))   # no native freqpoly verb
```

**Comparing distributions across a subgroup.** Map `drv` to colour for the
frequency polygon, or to fill plus a facet for the histogram. In tidyplots the
grouping aesthetic is always `color`, which maps to the fill for a histogram.

#### ggplot2

```{r}
#| label: freqpoly-grp-gg
ggplot(mpg, aes(displ, colour = drv)) +
  geom_freqpoly(binwidth = 0.5)
```

#### tidyplots

```{r}
#| label: freqpoly-grp-tp
mpg |>
  tidyplot(x = displ, color = drv, width = NA, height = NA) |>
  add(ggplot2::geom_freqpoly(binwidth = 0.5))
```

#### ggplot2

```{r}
#| label: hist-facet-gg
#| fig-height: 5
ggplot(mpg, aes(displ, fill = drv)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(~ drv, ncol = 1)
```

#### tidyplots

```{r}
#| label: hist-facet-tp
#| fig-height: 5
mpg |>
  # width = NA, height = NA frees tidyplots' fixed 1:1 panel size so the three
  # stacked panels flex to fill the figure (no clipped bottom panel).
  tidyplot(x = displ, color = drv, width = NA, height = NA) |>   # color = fill
  add_histogram(binwidth = 0.5) |>
  # split_plot(by = drv) free-scales each panel's y axis; the escape hatch
  # facet_wrap keeps ggplot2's shared y scale so panel heights are comparable.
  add(ggplot2::facet_wrap(~ drv, ncol = 1))
```

### 4.4 Bar charts

Two distinct cases. With raw data, `geom_bar()` *counts* rows per category, which
is `add_count_bar()` in tidyplots.

#### ggplot2

```{r}
#| label: bar-gg
ggplot(mpg, aes(manufacturer)) + geom_bar()
```

#### tidyplots

```{r}
#| label: bar-tp
mpg |>
  tidyplot(x = manufacturer, width = NA, height = NA) |>
  add_count_bar()
```

With data already summarized, ggplot2 needs `stat = "identity"` to plot the value
as is. tidyplots names that intent directly: `add_sum_bar()` draws the (single)
value per category.

#### ggplot2

```{r}
#| label: bar-id-gg
drugs <- data.frame(drug = c("a", "b", "c"), effect = c(4.2, 9.7, 6.1))
ggplot(drugs, aes(drug, effect)) + geom_bar(stat = "identity")
```

#### tidyplots

```{r}
#| label: bar-id-tp
drugs |>
  tidyplot(x = drug, y = effect, width = NA, height = NA) |>
  add_sum_bar()   # one value per group, plotted as is
```

### 4.5 Time series: line and path

Lines join points in x order, paths join them in data order. `geom_line()` maps to
`add_line()`. `geom_path()` has no native verb, so it uses the escape hatch.

Note the transformed y, `unemploy / pop`. ggplot2 computes it inside `aes()`, but
tidyplots needs a real column, so we `mutate()` it first.

#### ggplot2

```{r}
#| label: line-gg
ggplot(economics, aes(date, unemploy / pop)) +
  geom_line()
```

#### tidyplots

```{r}
#| label: line-tp
economics |>
  mutate(unemploy_rate = unemploy / pop) |>   # precompute, no aes() math
  tidyplot(x = date, y = unemploy_rate, width = NA, height = NA) |>
  add_line()
```

**Path coloured by year.** Both the y ratio and the `year(date)` colour are
transformed, so both become precomputed columns for tidyplots. The path itself is
the escape hatch, drawn uncoloured, then the coloured points go on top.

#### ggplot2

```{r}
#| label: path-gg
year <- function(x) as.POSIXlt(x)$year + 1900
ggplot(economics, aes(unemploy / pop, uempmed)) +
  geom_path(colour = "grey50") +
  geom_point(aes(colour = year(date)))
```

#### tidyplots

```{r}
#| label: path-tp
economics |>
  mutate(unemploy_rate = unemploy / pop,
         year = as.POSIXlt(date)$year + 1900) |>
  tidyplot(x = unemploy_rate, y = uempmed, color = year, width = NA, height = NA) |>
  add(ggplot2::geom_path(ggplot2::aes(color = NULL), colour = "grey50")) |>
  add_data_points()
```

## 5. Modifying the axes

**Axis titles.** `xlab()`/`ylab()` become `adjust_x_axis_title()`/
`adjust_y_axis_title()`, and dropping a title is a `remove_*_axis_title()` verb
rather than passing `NULL`. The semi-transparent points use a fixed `alpha` in
both.

#### ggplot2

```{r}
#| label: xlab-gg
ggplot(mpg, aes(cty, hwy)) +
  geom_point(alpha = 1 / 3) +
  xlab("city driving (mpg)") +
  ylab("highway driving (mpg)")
```

#### tidyplots

```{r}
#| label: xlab-tp
mpg |>
  tidyplot(x = cty, y = hwy, width = NA, height = NA) |>
  add_data_points(alpha = 1 / 3) |>
  adjust_x_axis_title("city driving (mpg)") |>
  adjust_y_axis_title("highway driving (mpg)")
```

**Removing the titles.**

#### ggplot2

```{r}
#| label: nolab-gg
ggplot(mpg, aes(cty, hwy)) +
  geom_point(alpha = 1 / 3) +
  xlab(NULL) + ylab(NULL)
```

#### tidyplots

```{r}
#| label: nolab-tp
mpg |>
  tidyplot(x = cty, y = hwy, width = NA, height = NA) |>
  add_data_points(alpha = 1 / 3) |>
  remove_x_axis_title() |>
  remove_y_axis_title()
```

**Axis limits.** Continuous limits are `ylim(20, 30)` versus
`adjust_y_axis(limits = c(20, 30))`. Selecting which *categories* of a discrete
axis to keep is `xlim("f", "r")` in ggplot2, but tidyplots has no category-picking
limit, so we filter the data instead.

#### ggplot2

```{r}
#| label: lim-gg
ggplot(mpg, aes(drv, hwy)) +
  geom_jitter(width = 0.25) +
  xlim("f", "r") +     # keep only front- and rear-wheel drive
  ylim(20, 30)         # clip y to 20..30
```

#### tidyplots

```{r}
#| label: lim-tp
mpg |>
  dplyr::filter(drv %in% c("f", "r")) |>   # the category pick, done as a filter
  tidyplot(x = drv, y = hwy, width = NA, height = NA) |>
  add_data_points_jitter() |>
  adjust_y_axis(limits = c(20, 30))
```

## 6. Output

Both packages return an object you can store, print, and save. tidyplots returns a
ggplot under the hood, so `ggsave()` still works, but tidyplots also ships its own
`save_plot()` (note the argument order: the plot comes first).

```{r}
#| label: output-tp
#| eval: false
# build and store
p <- mpg |>
  mutate(cyl_f = factor(cyl)) |>           # factor(cyl) precomputed for tidyplots
  tidyplot(x = displ, y = hwy, color = cyl_f, width = NA, height = NA) |>
  add_data_points()

p                                          # print it

save_plot(p, "plot.png", width = 10, height = 10, units = "cm")   # tidyplots
ggplot2::ggsave("plot.png", p, width = 5, height = 5)             # also works
```

## Takeaways

- For the everyday plots in this chapter, tidyplots has a one-to-one verb:
  `add_data_points()`, `add_curve_fit()`, `add_boxplot()`, `add_violin()`,
  `add_histogram()`, `add_count_bar()`, `add_line()`.
- The grammar difference is where the mapping lives: inside `aes()` for ggplot2,
  inside `tidyplot()` for tidyplots, with grouping always carried by `color`.
- Two things force you off the happy path. tidyplots will not evaluate transformed
  expressions in a mapping (so precompute with `mutate()`), and it has no verb for
  a handful of geoms like `geom_freqpoly()` and `geom_path()` (so reach for the
  escape hatch `add(ggplot2::...)`).
- `split_plot()` is tidyplots' faceting verb, but it free-scales each panel. Use
  `add(ggplot2::facet_wrap(...))` when you need ggplot2's shared scales.

## References

- Wickham H, Navarro D, Pedersen TL. *ggplot2: Elegant Graphics for Data Analysis*
  (3rd edition). The source for this handout (chapter *First steps*).
  <https://ggplot2-book.org/getting-started.html>
- Engler JB (2025). "Tidyplots empowers life scientists with easy code-based data
  visualization." *iMeta*, e70018.
  [doi:10.1002/imt2.70018](https://doi.org/10.1002/imt2.70018)
- Wood SN (2011). "Fast stable restricted maximum likelihood and marginal
  likelihood estimation of semiparametric generalized linear models." *Journal of
  the Royal Statistical Society (B)* 73(1):3-36.
  [doi:10.1111/j.1467-9868.2010.00749.x](https://doi.org/10.1111/j.1467-9868.2010.00749.x)
  (the `mgcv` GAM smoother used in the smoother section).

::: {.callout-note collapse="true"}
## Session Info

```{r}
#| label: session-info
#| echo: false
#| comment: ""
if (requireNamespace("sessioninfo", quietly = TRUE)) {
  sessioninfo::session_info(info = "all")
} else {
  sessionInfo()
}
```

:::
