---
title: "Two grammars, one figure set"
subtitle: "ggplot2 vs. tidyplots for scRNA-seq visualization"
author: "Bioinformatics & Research Computing"
format:
  revealjs:
    theme: simple
    slide-number: true
    code-overflow: wrap
    fig-align: center
    width: 1280
    height: 800
    margin: 0.06
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

```{=html}
<style>
.reveal pre code { font-size: 0.60em; line-height: 1.25; }
.reveal .columns p strong { font-size: 0.95em; }
.takeaway { background:#FBF1ED; border-radius:8px; padding:.45em .8em;
  font-size:.66em; margin-top:.5em; }
.takeaway::before { content:"TAKEAWAY"; display:block; color:#D9694C;
  font-weight:700; letter-spacing:.08em; font-size:.8em; margin-bottom:.15em; }
.gg { color:#3B4A8C; font-weight:700; }
.tp { color:#0E8C7F; font-weight:700; }
.reveal section img { box-shadow:none; border:none; }
</style>
```

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(tidyplots)
library(ggbeeswarm)   # geom_quasirandom()
library(ggsignif)     # geom_signif()

set.seed(42)
cell_types <- c("T cell", "B cell", "Monocyte")
conditions <- c("Control", "Stimulated")
genes      <- c("CD3D", "MS4A1", "LYZ", "IL2")

mean_expr <- function(gene, cell_type, condition) {
  base <- dplyr::case_when(
    gene == "CD3D"  & cell_type == "T cell"   ~ 3.0,
    gene == "MS4A1" & cell_type == "B cell"   ~ 3.2,
    gene == "LYZ"   & cell_type == "Monocyte" ~ 3.5,
    gene == "IL2"   & cell_type == "T cell"   ~ 1.0,
    TRUE ~ 0.3
  )
  base + ifelse(gene == "IL2" & cell_type == "T cell" &
                  condition == "Stimulated", 1.6, 0)
}

sc <- tidyr::expand_grid(cell_type = cell_types, condition = conditions,
                         gene = genes, cell = 1:40) |>
  mutate(mu         = purrr::pmap_dbl(list(gene, cell_type, condition), mean_expr),
         expression = pmax(0, rnorm(n(), mu, 0.8)))

il2     <- sc |> filter(gene == "IL2", cell_type == "T cell")
markers <- sc |> filter(gene %in% c("CD3D", "MS4A1", "LYZ"))
heat    <- sc |> group_by(gene, cell_type) |>
  summarise(mean_expr = mean(expression), .groups = "drop")
dot     <- sc |> group_by(gene, cell_type) |>
  summarise(pct = mean(expression > 0) * 100,
            avg = mean(expression), .groups = "drop") |>
  group_by(gene) |> mutate(avg_z = as.numeric(scale(avg))) |> ungroup()
```

## The big idea {.smaller}

Same **grammar of graphics** — opposite ergonomics.

:::: {.columns}

::: {.column width="50%"}
**[ggplot2]{.gg}**

- Map data to aesthetics in `aes()`, then stack `geom_*`, `stat_*`, `scale_*`, `theme_*`.
- Maximum control; more typing.
- Stats & significance need helpers (`stat_summary`, `ggsignif`).
- Faceting via `facet_wrap()`.
:::

::: {.column width="50%"}
**[tidyplots]{.tp}**

- Pipe a data frame into `tidyplot()`, then `add_<statistic>_<representation>()`.
- Publication-styled defaults out of the box.
- Means, error bars, tests are built-in verbs.
- Faceting via `split_plot(by = ...)`.
:::

::::

::: {.takeaway}
tidyplots is a stats-first on-ramp that never traps you: `add(ggplot2::geom_*())` drops any tidyplot back into raw ggplot2.
:::

## A small synthetic scRNA-seq dataset {.smaller}

4 marker genes × 3 cell types × 2 conditions, 40 cells per group → 960 rows, tidy long format. `IL2` is wired to be **induced by stimulation in T cells** — a real signal for the test.

```{r}
#| echo: true
sc |> slice_sample(n = 6) |> knitr::kable()
```

# The figures, built both ways

## Plot 1 — points + mean ± SEM + test {.smaller .scrollable}

:::: {.columns}

::: {.column width="50%"}
**[ggplot2]{.gg}**
```{r}
#| fig-width: 4.2
#| fig-height: 3.4
ggplot(il2, aes(condition, expression, color = condition)) +
  stat_summary(aes(fill = condition), fun = mean,
               geom = "col", alpha = .4, color = NA) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = .2) +
  geom_quasirandom(width = .25, size = 1) +
  geom_signif(comparisons = list(c("Control", "Stimulated")),
              map_signif_level = TRUE, color = "black") +
  labs(x = NULL, y = "IL2 expression") +
  theme_classic(base_size = 12) + theme(legend.position = "none")
```
:::

::: {.column width="50%"}
**[tidyplots]{.tp}**
```{r}
#| fig-width: 4.2
#| fig-height: 3.4
il2 |>
  tidyplot(x = condition, y = expression, color = condition) |>
  add_mean_bar(alpha = .4) |>
  add_sem_errorbar() |>
  add_data_points_beeswarm() |>
  add_test_asterisks()
```
:::

::::

::: {.takeaway}
Statistic and visual are named together. Significance is one verb in tidyplots vs. an extra package (`ggsignif`) plus `stat_summary` in ggplot2.
:::

## Plot 2 — marker-gene violins (faceting) {.smaller .scrollable}

**[ggplot2]{.gg}** — `geom_violin` + `geom_boxplot` + `facet_wrap`
```{r}
#| fig-width: 9
#| fig-height: 2.7
ggplot(markers, aes(cell_type, expression, fill = cell_type)) +
  geom_violin(scale = "width", color = NA, alpha = .7) +
  geom_boxplot(width = .12, outlier.size = .4) +
  facet_wrap(~ gene, scales = "free_y") +
  labs(x = NULL) + theme_bw(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

**[tidyplots]{.tp}** — `add_violin` + `add_boxplot` + `split_plot`
```{r}
#| fig-width: 9
#| fig-height: 2.7
markers |>
  tidyplot(x = cell_type, y = expression, color = cell_type) |>
  add_violin() |>
  add_boxplot(width = .15) |>
  split_plot(by = gene)
```

::: {.takeaway}
`facet_wrap()` keeps one object; `split_plot()` returns independent panels. The closest the two packages ever feel.
:::

## Plot 3 — mean-expression heatmap {.smaller .scrollable}

:::: {.columns}

::: {.column width="50%"}
**[ggplot2]{.gg}**
```{r}
#| fig-width: 4.2
#| fig-height: 3.2
ggplot(heat, aes(cell_type, gene, fill = mean_expr)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = NULL, y = NULL, fill = "Mean") +
  theme_minimal(base_size = 12)
```
:::

::: {.column width="50%"}
**[tidyplots]{.tp}**
```{r}
#| fig-width: 4.2
#| fig-height: 3.2
heat |>
  tidyplot(x = cell_type, y = gene, color = mean_expr) |>
  add_heatmap()
```
:::

::::

::: {.takeaway}
Both `summarise()` first. tidyplots brings the continuous scale for free; ggplot2 leaves the palette to you.
:::

## Plot 4 — the dot plot: the ceiling {.smaller .scrollable}

Shared precompute — two statistics + a per-gene z-score (neither package computes these for you):

```{r}
#| eval: false
dot <- sc |>
  group_by(gene, cell_type) |>
  summarise(pct = mean(expression > 0) * 100,   # % expressing
            avg = mean(expression)) |>          # mean expression
  group_by(gene) |>
  mutate(avg_z = scale(avg))                    # z-score per gene
```

:::: {.columns}

::: {.column width="50%"}
**[ggplot2]{.gg}**
```{r}
#| fig-width: 4.2
#| fig-height: 3.0
ggplot(dot, aes(cell_type, gene)) +
  geom_point(aes(size = pct, color = avg_z)) +
  scale_size_area(max_size = 9) +
  scale_color_gradient2(midpoint = 0, low = "grey80",
                        mid = "lightsteelblue", high = "firebrick") +
  labs(x = NULL, y = NULL) + theme_bw(base_size = 12)
```
:::

::: {.column width="50%"}
**[tidyplots]{.tp}**
```{r}
#| fig-width: 4.2
#| fig-height: 3.0
dot |>
  tidyplot(x = cell_type, y = gene, color = avg_z) |>
  add(ggplot2::geom_point(ggplot2::aes(size = pct))) |>
  add(ggplot2::scale_size_area(max_size = 9))
```
:::

::::

::: {.takeaway}
Two statistics on one mark → no native `add_*`. You precompute and drop to ggplot2 via `add()`. The escape hatch becomes the only way through.
:::

## When to reach for which {.smaller}

- **[tidyplots]{.tp}** — teaching, exploration, and standard scientific figures. Stats-first verbs and publication defaults get students to a clean, correct plot fast.
- **[ggplot2]{.gg}** — full control and anything non-standard. The grammar has no ceiling.
- **both, together** — they share an engine. Start tidy, then `add(ggplot2::geom_*())` the moment a figure outgrows the `add_*` model.

::: {.takeaway}
High-level grammars are wonderful until a figure encodes multiple derived quantities — and the dot plot is the canonical scRNA-seq case where that happens.
:::
