---
title: "ggplot2 vs. tidyplots: the same scRNA-seq figures, two grammars"
subtitle: "A side-by-side teaching handout"
author: "Bioinformatics & Research Computing"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    fig-width: 4.5
    fig-height: 3.5
    theme: cosmo
execute:
  warning: false
  message: false
  echo: true
---

## What this handout shows

Both packages sit on top of the **grammar of graphics**, but they ask you to
think differently:

| | **ggplot2** | **tidyplots** |
|---|---|---|
| Mental model | map data → aesthetics, then stack *geoms* and *stats* | pipe a data frame in, then `add_*()` named components |
| Verb style | `geom_*()`, `stat_*()`, `scale_*()`, `theme_*()` | `add_<statistic>_<representation>()` |
| Defaults | general-purpose grey theme | publication-styled (small fonts, clean axes) out of the box |
| Stats / significance | bring a helper (`ggsignif`, `ggpubr`, `stat_summary`) | built in: `add_mean_bar()`, `add_sem_errorbar()`, `add_test_*()` |
| Faceting | `facet_wrap()` / `facet_grid()` | `split_plot(by = ...)` |
| Escape hatch | — *is* the escape hatch | `add(ggplot2::geom_*())` drops back to raw ggplot2 |

**Teaching point:** tidyplots is a gentle, statistics-first on-ramp that stays
ggplot2-compatible — students learn a tidy `add_*()` vocabulary but never get
trapped, because any tidyplot can absorb raw ggplot2 layers.

## Setup

```{r}
#| label: setup
library(tidyverse)   # ggplot2, dplyr, tidyr, purrr
library(tidyplots)   # install.packages("tidyplots")

# Only needed to make the *ggplot2* version match tidyplots' built-ins.
# The fact that these are extra installs is itself part of the lesson.
library(ggbeeswarm)  # geom_quasirandom()  -> beeswarm points
library(ggsignif)    # geom_signif()        -> significance brackets
```

## A tiny synthetic scRNA-seq dataset

We simulate log-normalized expression for four marker genes across three cell
types and two conditions. `IL2` is wired to be **induced by stimulation in T
cells** so we have a real effect to test.

```{r}
#| label: simulate
set.seed(42)

cell_types <- c("T cell", "B cell", "Monocyte")
conditions <- c("Control", "Stimulated")
genes      <- c("CD3D", "MS4A1", "LYZ", "IL2")

# biologically-flavored mean expression on a log-normalized scale
mean_expr <- function(gene, cell_type, condition) {
  base <- dplyr::case_when(
    gene == "CD3D"  & cell_type == "T cell"   ~ 3.0,   # T-cell marker
    gene == "MS4A1" & cell_type == "B cell"   ~ 3.2,   # B-cell marker
    gene == "LYZ"   & cell_type == "Monocyte" ~ 3.5,   # monocyte marker
    gene == "IL2"   & cell_type == "T cell"   ~ 1.0,   # low at baseline
    TRUE ~ 0.3
  )
  bump <- ifelse(gene == "IL2" & cell_type == "T cell" &
                   condition == "Stimulated", 1.6, 0)  # the effect
  base + bump
}

n_cells <- 40
sc <- tidyr::expand_grid(
  cell_type = cell_types,
  condition = conditions,
  gene      = genes,
  cell      = seq_len(n_cells)
) |>
  dplyr::mutate(
    cell_id    = paste(cell_type, condition, cell, sep = "_"),
    mu         = purrr::pmap_dbl(list(gene, cell_type, condition), mean_expr),
    expression = pmax(0, rnorm(dplyr::n(), mean = mu, sd = 0.8))
  )

dplyr::glimpse(sc)
```

## Plot 1 — Expression by condition, with points + mean ± SEM + test

The headline contrast. tidyplots gives you beeswarm points, a mean bar, SEM
error bars, *and* a significance test in five readable verbs. ggplot2 needs
`stat_summary()` plus two extra packages to match it.

```{r}
#| label: prep-il2
il2 <- sc |> dplyr::filter(gene == "IL2", cell_type == "T cell")
```

#### ggplot2

```{r}
#| label: p1-ggplot2
#| fig-cap: "ggplot2: stat_summary + ggbeeswarm + ggsignif"
ggplot(                                  # start the plot
  il2,                                   # the data frame to plot
  aes(                                   # aes() maps data columns to visual roles
    x = condition,                       # condition on the x axis
    y = expression                       # expression on the y axis
  )
) +                                      # + adds the next layer
  stat_summary(                          # compute a summary statistic per x group and draw it
    aes(fill = condition),               # fill (bar interior color) varies by condition
    fun = mean,                          # the statistic to compute is the mean of y
    geom = "col",                        # draw the result as a column (bar)
    width = 0.6,                         # bar width (1 = bars touch)
    alpha = 0.4,                         # 40% opacity, so points behind show through
    color = NA                           # no outline around the bars
  ) +
  stat_summary(                          # a second summary layer, for the error bars
    fun.data = mean_se,                  # compute the mean and its +/- standard-error range
    geom = "errorbar",                   # draw that range as an error bar
    width = 0.2                          # width of the little horizontal caps
  ) +
  ggbeeswarm::geom_quasirandom(          # draw the individual cells as beeswarm points (spread sideways so they do not overlap)
    aes(color = condition),              # point color varies by condition
    width = 0.25,                        # how far points are allowed to spread sideways
    size = 1                             # point size
  ) +
  ggsignif::geom_signif(                 # draw a significance bracket and run the test
    comparisons = list(                  # which x groups to compare
      c("Control", "Stimulated")         # compare Control vs Stimulated
    ),
    map_signif_level = TRUE              # show stars (*, **, ***) instead of the raw p-value
  ) +
  labs(                                  # set the labels
    x = NULL,                            # remove the x-axis title
    y = "IL2 expression",                # y-axis title
    title = "IL2 in T cells"             # plot title
  ) +
  theme_classic(                         # a clean built-in look (white background, axis lines, no grid)
    base_size = 11                       # base font size in points
  ) +
  theme(                                 # override individual style elements
    legend.position = "none"             # hide the legend (the x axis already labels the conditions)
  )
```

#### tidyplots

```{r}
#| label: p1-tidyplots
#| fig-cap: "tidyplots: five add_*() verbs, stats included"
il2 |>
  tidyplot(x = condition, y = expression, color = condition) |>
  add_mean_bar(alpha = 0.4) |>
  add_sem_errorbar() |>
  add_data_points_beeswarm() |>
  add_test_asterisks() |>
  add_title("IL2 in T cells") |>
  adjust_legend_position("none")
```

**What to notice**

- The statistic and its visual form are *named together* in tidyplots
  (`add_mean_bar`, `add_sem_errorbar`), versus ggplot2's `stat_summary(fun = ..., geom = ...)`.
- Significance testing is one verb in tidyplots; in ggplot2 it's an add-on package.
- ggplot2 wins when you need an unusual `stat_summary` (e.g. a custom function) —
  full control, more typing.

## Plot 2 — Marker-gene violins across cell types (faceting)

```{r}
#| label: prep-markers
markers <- sc |> dplyr::filter(gene %in% c("CD3D", "MS4A1", "LYZ"))
```

#### ggplot2

```{r}
#| label: p2-ggplot2
#| fig-width: 7
#| fig-height: 3
#| fig-cap: "ggplot2: geom_violin + geom_boxplot + facet_wrap"
ggplot(
  markers,                               # the data frame to plot
  aes(
    x = cell_type,                       # cell_type on the x axis
    y = expression,                      # expression on the y axis
    fill = cell_type                     # fill color varies by cell_type
  )
) +
  geom_violin(                           # draw a violin (a mirrored density of the distribution)
    scale = "width",                     # give every violin the same maximum width, so shapes are comparable across groups of different size
    color = NA,                          # no outline
    alpha = 0.7                          # 70% opacity
  ) +
  geom_boxplot(                          # overlay a narrow box-and-whisker on each violin
    width = 0.12,                        # make the box narrow so the violin stays visible
    outlier.size = 0.4,                  # draw outlier points small
    alpha = 0.9                          # nearly opaque
  ) +
  facet_wrap(                            # split the plot into one small panel per gene
    ~ gene,                              # the variable to split panels by
    scales = "free_y"                    # let each panel have its own y-axis range
  ) +
  labs(
    x = NULL,                            # remove the x-axis title
    y = "Expression"                     # y-axis title
  ) +
  theme_bw(                              # the black-and-white theme (white panel, grey gridlines, border)
    base_size = 11                       # base font size in points
  ) +
  theme(
    legend.position = "none",            # hide the legend
    axis.text.x = element_text(          # style the x-axis tick labels
      angle = 30,                        # rotate them 30 degrees
      hjust = 1                          # right-justify so long labels do not overlap
    )
  )
```

#### tidyplots

```{r}
#| label: p2-tidyplots
#| fig-width: 7
#| fig-height: 3
#| fig-cap: "tidyplots: add_violin + add_boxplot + split_plot"
markers |>
  tidyplot(x = cell_type, y = expression, color = cell_type) |>
  add_violin() |>
  add_boxplot(width = 0.15) |>
  adjust_legend_position("none") |>
  split_plot(by = gene)
```

**What to notice**

- `facet_wrap()` keeps everything in one ggplot object; `split_plot()` returns a
  set of independent tidyplots stitched together — handy when each panel needs
  different adjustments later.
- Same layered idea, near-identical line count here. Faceting is where the two
  packages feel most similar.

## Plot 3 (bonus) — Mean-expression heatmap

Both need a `group_by() |> summarise()` first; the plotting line is where they diverge.

```{r}
#| label: prep-heat
heat <- sc |>
  dplyr::group_by(gene, cell_type) |>
  dplyr::summarise(mean_expr = mean(expression), .groups = "drop")
```

#### ggplot2

```{r}
#| label: p3-ggplot2
#| fig-width: 4
#| fig-height: 3
ggplot(
  heat,                                  # the data frame to plot
  aes(
    x = cell_type,                       # cell_type on the x axis
    y = gene,                            # gene on the y axis
    fill = mean_expr                     # the continuous mean_expr drives the fill color
  )
) +
  geom_tile() +                          # draw one rectangle per x/y combination (the heatmap cells), colored by fill
  scale_fill_viridis_c() +               # apply the viridis color scale for a continuous (_c) fill; viridis is perceptually uniform and colorblind-friendly
  labs(
    x = NULL,                            # remove the x-axis title
    y = NULL,                            # remove the y-axis title
    fill = "Mean"                        # title for the color legend
  ) +
  theme_minimal(                         # a light theme with no axis lines and faint gridlines
    base_size = 11                       # base font size in points
  )
```

#### tidyplots

```{r}
#| label: p3-tidyplots
#| fig-width: 4
#| fig-height: 3
heat |>
  tidyplot(x = cell_type, y = gene, color = mean_expr) |>
  add_heatmap()
```

**What to notice**

- tidyplots maps the continuous fill through `color =` and `add_heatmap()` brings
  a sensible diverging/continuous scale for free; ggplot2 leaves the palette
  choice (here `scale_fill_viridis_c()`) explicitly to you.

## Plot 4 — The dot plot: where both grammars hit their ceiling

The Seurat `DotPlot` / Scanpy `sc.pl.dotplot` grid is the one canonical scRNA-seq
figure that neither package draws from raw data. Each dot encodes **two** derived
statistics at once:

- **size** = % of cells in the group expressing the gene (`mean(expr > 0)`)
- **color** = mean expression, **z-scored within each gene** across groups

Every earlier figure mapped to *one mark plus at most one computed statistic*.
This one needs a genuine precompute — two summaries *and* a per-gene
normalization — before anything reaches the canvas.

```{r}
#| label: prep-dot
dot <- sc |>
  dplyr::group_by(gene, cell_type) |>
  dplyr::summarise(
    pct_expr = mean(expression > 0) * 100,   # % of cells expressing
    avg_expr = mean(expression),             # mean expression
    .groups  = "drop"
  ) |>
  dplyr::group_by(gene) |>
  dplyr::mutate(avg_scaled = as.numeric(scale(avg_expr))) |>  # z-score per gene
  dplyr::ungroup()

dot
```

#### ggplot2

Natural here, but fully manual: you wire up the two scales yourself. Note the
naming trap — `geom_dotplot()` is a Wilkinson dot *histogram*, unrelated to this.

```{r}
#| label: p4-ggplot2
#| fig-width: 4.5
#| fig-height: 3.5
#| fig-cap: "ggplot2: geom_point with size + color mapped to two statistics"
ggplot(
  dot,                                   # the data frame to plot
  aes(
    x = cell_type,                       # cell_type on the x axis
    y = gene                             # gene on the y axis
  )
) +
  geom_point(                            # draw one dot per group, encoding TWO statistics at once
    aes(
      size = pct_expr,                   # dot area grows with % of cells expressing the gene
      color = avg_scaled                 # dot color shows the per-gene z-scored mean expression
    )
  ) +
  scale_size_area(                       # map the size value to dot AREA (not radius), the honest way to encode magnitude
    max_size = 9,                        # area of the largest dot
    name = "% expressing"                # legend title for the size scale
  ) +
  scale_color_gradient2(                 # a diverging color scale with a defined midpoint
    low = "grey80",                      # color at the low end
    mid = "lightsteelblue",              # color at the midpoint
    high = "firebrick",                  # color at the high end
    midpoint = 0,                        # 0 (the z-score mean) maps to the mid color
    name = "Avg expr\n(z-score)"         # legend title ("\n" puts the rest on a second line)
  ) +
  labs(
    x = NULL,                            # remove the x-axis title
    y = NULL                             # remove the y-axis title
  ) +
  theme_bw(
    base_size = 11                       # base font size in points
  ) +
  theme(
    axis.text.x = element_text(          # style the x-axis tick labels
      angle = 30,                        # rotate them 30 degrees
      hjust = 1                          # right-justify so long labels do not overlap
    ),
    panel.grid.minor = element_blank()   # element_blank() removes an element; here it hides the minor gridlines
  )
```

#### tidyplots

There is **no** `add_*()` for two statistics on one mark, so the idiomatic move
is to precompute (above) and drop back to raw ggplot2 through `add()`. The color
mapping set in `tidyplot()` is inherited by the added `geom_point()`; the size
scale you still supply yourself. In other words, for this one figure you leave
the tidy grammar behind.

```{r}
#| label: p4-tidyplots
#| fig-width: 4.5
#| fig-height: 3.5
#| fig-cap: "tidyplots: precompute, then add(ggplot2::geom_point()) — the escape hatch as necessity"
dot |>
  tidyplot(x = cell_type, y = gene, color = avg_scaled) |>
  add(                                   # add() drops a raw ggplot2 layer into a tidyplot
    ggplot2::geom_point(                 # draw the dots (the color avg_scaled is inherited from tidyplot() above)
      ggplot2::aes(size = pct_expr)      # map the second statistic to dot size
    )
  ) |>
  add(                                   # add the same area-based size scale used in the pure-ggplot2 version
    ggplot2::scale_size_area(
      max_size = 9,                      # area of the largest dot
      name = "% expressing"              # legend title for the size scale
    )
  ) |>
  adjust_size(                           # set the plotting-area size (a tidyplots verb)
    width = 55,                          # width in millimetres
    height = 45                          # height in millimetres
  )
```

**What to notice**

- The interesting work is in dplyr, not the plotting verb: neither library
  computes "% expressing" or a per-gene z-score for you.
- ggplot2 handles it cleanly once the table exists — two scales, by hand.
- tidyplots' `add_<statistic>_<representation>()` model assumes one statistic per
  mark, so a two-statistic dot has no native verb. `add()` stops being a
  curiosity and becomes the only way through.
- **The teaching line:** high-level grammars are wonderful until a figure encodes
  multiple derived quantities — and the DotPlot is the canonical scRNA-seq case
  where that happens.

## Discussion prompts for students

- When is ggplot2's explicitness an advantage rather than overhead?
- Rewrite Plot 1's ggplot2 version *without* `ggbeeswarm`/`ggsignif`. What do you lose?
- Use `add(ggplot2::geom_*())` to bolt a custom ggplot2 layer onto a tidyplot —
  confirm the escape hatch actually works.
- In Plot 4, what would break if you z-scored `avg_expr` *globally* instead of
  per gene? Why does the per-gene scaling matter for reading a marker dot plot?
